\bookheader{Esther}
\labelbook{Esth}

\bookpretitle{The Book of}
\booktitle{Esther}

\labelchapt{1}
\passage{The Wealth of King Ahasuerus}

\chapt{1}
\v{1}This is a record of\fnote{\fbackref{1:1} The Heb. lacks \fbib{a record of}} what happened during the reign\fnote{\fbackref{1:1} Lit. \fbib{days}} of Ahasuerus,\fnote{\fbackref{1:1} Or possibly \fbib{Xerxes}, and so throughout the book} the Ahasuerus who ruled over 127 provinces from India to Cush.\fnote{\fbackref{1:1} I.e. the upper Nile region} \v{2}At that time King Ahasuerus was ruling from\fnote{\fbackref{1:2} Lit. \fbib{was sitting on his royal throne in}} Susa the capital.\fnote{\fbackref{1:2} Or \fbib{Susa the fortress,} and so throughout the book} \v{3}In the third year of his reign, he gave a banquet for all his officials and ministers,\fnote{\fbackref{1:3} Or \fbib{his servants}} and the military leaders\fnote{\fbackref{1:3} Lit. \fbib{army}} of Persia and Media, the nobles, and the provincial officials were present.\fnote{\fbackref{1:3} Lit. \fbib{before him}} \v{4}He displayed the enormous wealth of his kingdom, along with its splendid beauty and greatness for many days---for 180 days in all.\fnote{\fbackref{1:4} The Heb. lacks \fbib{in all}}
\passage{The Banquet of King Ahasuerus}

\v{5}When those days were over, the king held a seven-day banquet in the courtyard of the garden of his\fnote{\fbackref{1:5} Lit. \fbib{the king's}} palace for all the people who were present in Susa the capital, from the greatest to the least important. \v{6}There were curtains of white and blue linen tied with cords of fine linen and purple material\fnote{\fbackref{1:6} The Heb. lacks \fbib{material}} to silver rings on\fnote{\fbackref{1:6} Lit. \fbib{and}} marble columns. There were couches of gold and silver on a mosaic pavement of porphyry, marble, mother-of-pearl and other precious stones. \v{7}Drinks were served in gold vessels of various kinds, and there was plenty of royal wine because the king was generous.\fnote{\fbackref{1:7} Lit. \fbib{wine according to the hand of the king}} \v{8}According to the king's\fnote{\fbackref{1:8} The Heb. lacks \fbib{king's}} decree the drinking was not compulsory because the king instructed\fnote{\fbackref{1:8} Lit. \fbib{established}} every steward in his house to serve each individual what he desired. \v{9}Queen Vashti also held a banquet in the royal palace of King Ahasuerus for the women.
\passage{Vashti Refuses to Obey the King}

\v{10}A week later, when the king was under the influence of all that wine,\fnote{\fbackref{1:10} Lit. \fbib{the heart of the king was happy with wine}} he ordered Mehuman, Biztha, Harbona, Bigtha, Abagtha, Zethar, and Carcas, the seven eunuchs who served King Ahasuerus, \v{11}to bring Queen Vashti to the king, wearing\fnote{\fbackref{1:11} Lit. \fbib{with}} the royal crown to display her beauty to the people and the officials, since she was lovely to look at. \v{12}Queen Vashti refused to come at the king's order that was brought by the eunuchs.
\passage{The King Removes Vashti as Queen}

Then the king flew into a rage. \v{13}The king spoke to the wise men who understood the times, for it was the king's custom to consult\fnote{\fbackref{1:13} Lit. \fbib{to speak before}} all those who understood law and justice. \v{14}(His closest advisors\fnote{\fbackref{1:14} Lit. \fbib{Those closest to him}} were: Carshena, Shethar, Admatha, Tarshish, Meres, Marsena, and Memucan, the seven officials of Persia and Media who had direct access\fnote{\fbackref{1:14} Lit. \fbib{saw the face of} } to the king and who held the highest rank\fnote{\fbackref{1:14} Lit. \fbib{sat first in the kingdom}} in the kingdom.) \v{15}The king inquired,\fnote{\fbackref{1:15} The Heb. lacks \fbib{The king inquired}} ``According to law, what should be done to Queen Vashti because she did not obey the order of King Ahasuerus that was delivered by the eunuchs?''

\v{16}Then Memucan replied in the presence of the king and his officials, ``It is not the king alone whom Vashti has wronged, but rather all of the officials and all of the people who are in the provinces of King Ahasuerus. \v{17}When the report about the queen goes out to all the women, it will cause them to despise their husbands.\fnote{\fbackref{1:17} Lit. \fbib{husbands in their eyes}} They'll say, `King Ahasuerus ordered Queen Vashti to be brought before him, but she wouldn't come.' \v{18}This very day the wives of the officials\fnote{\fbackref{1:18} Or \fbib{women of nobility}} of Persia and Media who hear the report about the queen will speak in the same way to all the officials of the king, and there will be more than enough contempt and anger. \v{19}If it seems good to the king, let a royal decree go out from him and let it be written in the laws of Persia and Media, which cannot be repealed, that Vashti is never again to enter the presence of King Ahasuerus. Let the king give her royal position to another woman who is better than she. \v{20}When the edict of the king that he issues is heard throughout his kingdom---for it's vast---then all the women will give honor to their husbands, from the greatest to the least important.''

\v{21}This seemed like a good idea\fnote{\fbackref{1:21} The Heb. lacks \fbib{idea}} to the king and his officials, so the king did what Memucan suggested.\fnote{\fbackref{1:21} Lit. \fbib{according to the word of Memucan}} \v{22}He sent letters to all the provinces of the king, written in the script of that province,\fnote{\fbackref{1:22} Lit. \fbib{to each province according to its writing}} and to each people in their own language, ordering that every man should be the master in his house and speak the language of his own people.
\labelchapt{2}
\passage{The King Searches for a New Queen}

\chapt{2}
\v{1}After this, when the anger of King Ahasuerus had subsided, he remembered Vashti, what she had done, and what had been decreed about her. \v{2}Then the young men who attended the king suggested, ``Let beautiful young virgins be sought for the king. \v{3}Let the king appoint officials in all the provinces of his kingdom to bring all the beautiful young virgins into the harem in Susa the capital. Let them be placed under the care of Hegai, the king's eunuch, who is in charge of the women to give them their beauty treatments.\fnote{\fbackref{2:3} Lit. \fbib{their massages}} \v{4}Then let the young woman who pleases the king rule in place of Vashti.'' This advice\fnote{\fbackref{2:4} The Heb. lacks \fbib{advice}} pleased the king, and he did this.
\passage{Esther's Background}

\v{5}In Susa the capital there was a Jewish man from the tribe of Benjamin, Jair's son Mordecai, who was a descendant of Kish's son Shimei the descendant of Benjamin. \v{6}He had been taken into captivity from Jerusalem along with the exiles who had been deported with Jeconiah, king of Judah, whom Nebuchadnezzar, king of Babylon had taken into exile.\fnote{\fbackref{2:6} This deportation took place in 597 B.C.} \v{7}Mordecai\fnote{\fbackref{2:7} Lit. \fbib{He}} had raised his cousin\fnote{\fbackref{2:7} Lit. \fbib{his uncle's daughter}} Hadassah, or Esther,\fnote{\fbackref{2:7} The Heb. name \fbib{Hadassah} means \fbib{Myrtle}; The Persian name \fbib{Esther} means \fbib{Star}} because she had no father or mother. The young woman had a beautiful figure and was very attractive. When her mother and father died Mordecai had taken her as his daughter.

\v{8}The king's order and edict was proclaimed, and many young women were brought to Susa the capital under the care of Hegai. Esther was taken to the palace into the care of Hegai, who was in charge of the women. \v{9}The young woman pleased him and gained his favor. He quickly provided her beauty treatments and gave her portions of food to her. He also assigned her seven suitable young women from the palace and transferred her and her young women to the best place in the harem. \v{10}Esther did not make known her people or heritage\fnote{\fbackref{2:10} Or \fbib{her ancestry}} because Mordecai had instructed her not to make it known. \v{11}Every day Mordecai would walk back and forth in front of the court of the harem to find out about Esther's well-being and what was happening to her.
\passage{Esther Becomes Queen}

\v{12}Each young woman's turn came to go in to King Ahasuerus at the end of the twelve month period, at which time she was treated according to the regulations for women. This process\fnote{\fbackref{2:12} The Heb. lacks \fbib{process}} completed the period of her beauty treatments---six months with oil of myrrh and six months with spices and cosmetics for women. \v{13}After that the young woman would go in to the king, and whatever she asked for would be given to her to take with her from the harem to the palace. \v{14}In the evening she would go into the palace\fnote{\fbackref{2:14} The Heb. lacks \fbib{the palace}} and in the morning she would return to the second harem, into the care of Shaashgaz, the king's eunuch who was in charge of the mistresses.\fnote{\fbackref{2:14} Or \fbib{concubines}; i.e. secondary wives} She would not go again to the king unless the king wanted her and she was called for by name. \v{15}Now Esther was the daughter of Abihail, who had been Mordecai's uncle. Mordecai had taken Esther in as his own\fnote{\fbackref{2:15} The Heb. lacks \fbib{own}} daughter. When her turn came to go in to the king, she did not want anything except what Hegai, the king's eunuch in charge of the harem, advised. Esther found favor with everyone who saw her. \v{16}Esther was taken to King Ahasuerus to his royal palace in the tenth month, which is the month Tebeth, in the seventh year of his reign.

\v{17}The king loved Esther more than any of the other women, so he favored her and was kinder to her than he was to any of the other virgins. He put the royal crown on her head and made her queen in place of Vashti. \v{18}The king put on a great banquet for all his officials and ministers\fnote{\fbackref{2:18} Or \fbib{servants}} to honor Esther. He declared a holiday for the provinces and gave generous gifts.\fnote{\fbackref{2:18} Lit. \fbib{gave gifts according to the hand of the king}}
\passage{Mordecai Thwarts a Plot to Kill Ahasuerus}

\v{19}When the virgins were gathered a second time, Mordecai was sitting in the king's gate. \v{20}Now Esther had not declared her heritage\fnote{\fbackref{2:20} Or \fbib{her ancestry}} or her people, just as Mordecai had instructed her, for Esther did what Mordecai told her just as she had done when she was raised by him. \v{21}At that time when Mordecai was sitting in the king's gate, Bigthan and Teresh, two of the king's eunuchs among those who guarded the threshold,\fnote{\fbackref{2:21} I.e. the entrance to the restricted areas of the palace} became angry and conspired to assassinate\fnote{\fbackref{2:21} Lit. \fbib{to send a hand against}} King Ahasuerus. \v{22}When Mordecai learned about the plot, he told Queen Esther, and she told the king in Mordecai's name. \v{23}After the matter had been fully investigated, Bigthan and Teresh\fnote{\fbackref{2:23} Lit. \fbib{investigated, the two of them}} were hanged\fnote{\fbackref{2:23} Or \fbib{impaled}} on a pole, and this was recorded in the Book of the Chronicles in the presence of the king.
\labelchapt{3}
\passage{Haman is Promoted by Ahasuerus}

\chapt{3}
\v{1}Sometime later King Ahasuerus promoted Hammedatha the Agagite's son Haman, elevating him to a position above\fnote{\fbackref{3:1} Lit. \fbib{setting his seat above}} all the other officials who were with him. \v{2}All the king's ministers\fnote{\fbackref{3:2} Or \fbib{servants}} who were in the king's gate would kneel and bow down to Haman, because the king had commanded that Haman\fnote{\fbackref{3:2} Lit. \fbib{commanded concerning him that he}} be honored in this way. Mordecai, however, would not kneel and would not bow down.

\v{3}The king's ministers\fnote{\fbackref{3:3} Or \fbib{servants}} who were in the king's gate asked Mordecai, ``Why are you disobeying the king's command?'' \v{4}They asked him this day after day, and he would not listen to them, so they told Haman to see whether or not Mordecai would get away with his disobedience,\fnote{\fbackref{3:4} Lit. \fbib{would stand}} since he also had told them that he was Jewish. \v{5}When Haman saw that Mordecai would not kneel and bow down to him, he\fnote{\fbackref{3:5} Lit. \fbib{Haman}} flew into a rage. \v{6}Because they had told him who the people of Mordecai were, Haman\fnote{\fbackref{3:6} Lit. \fbib{he}} found it unacceptable\fnote{\fbackref{3:6} Lit. \fbib{contemptible}} to kill\fnote{\fbackref{3:6} Lit. \fbib{sending a hand against}} only Mordecai. So Haman sought to destroy all of Mordecai's people, the Jewish people, who were in all the kingdom of Ahasuerus.
\passage{Haman's Plot against the Jewish People}

\v{7}In the twelfth year of the reign of\fnote{\fbackref{3:7} The Heb. lacks \fbib{the reign of}} King Ahasuerus, in the first month (the month Nisan), the pur (that is, the lot) was cast in Haman's presence to determine the best day and month to carry out his plot.\fnote{\fbackref{3:7} Lit. \fbib{before Haman from day to day and month to month}} The lot indicated the twelfth month, the month Adar.\fnote{\fbackref{3:7} Lit. \fbib{day and month, the twelfth, the month Adar}} \v{8}Then Haman told King Ahasuerus, ``There is a certain people scattered and divided among the people throughout the provinces of your kingdom. Their laws are different than all the other people, they don't obey the king's laws, and it's not in the king's best interest\fnote{\fbackref{3:8} Lit. \fbib{there is no advantage for the king}} to leave them alone. \v{9}If the king approves, let it be decreed\fnote{\fbackref{3:9} Lit. \fbib{written}} that they're to be destroyed, and I'll measure out 10,000 silver talents\fnote{\fbackref{3:9} I.e. about 750,000 pounds; a talent weighed about 75 pounds} and bring it to the king's treasury for those who will do the work.''

\v{10}The king removed his signet ring from his hand and gave it to Hammedatha the Agagite's son Haman, the enemy of the Jewish people. \v{11}The king told Haman, ``The silver is given to you, along with the people, to do with them as you see fit.''

\v{12}The king's scribes were summoned on the thirteenth day of the first month, and all that Haman commanded was written to the regional authorities\fnote{\fbackref{3:12} Lit. \fbib{satraps}; Persian government officials similar in authority to a governor} of the king, to the governors who were over each province, and to the officials of each people. This order\fnote{\fbackref{3:12} The Heb. lacks \fbib{order}} was translated in the name of King Ahasuerus into the language of each province\fnote{\fbackref{3:12} Lit. \fbib{Ahasuerus to the people, the script}} and bore the seal of the king's signet ring. \v{13}Letters were sent by couriers to all of the king's provinces to annihilate, to kill, and to destroy all the Jewish people, both young and old, women and children, and to confiscate their goods\fnote{\fbackref{3:13} Lit. \fbib{spoil}} on a single day---the thirteenth day of the twelfth month of Adar. \v{14}A copy of the letter was to be issued as an edict in every province and published to all the people, telling them\fnote{\fbackref{3:14} The Heb. lacks \fbib{telling them}} to be ready for that day. \v{15}The couriers went out, urged on by the king's command, and the edict was issued in Susa the capital. The king and Haman sat down to drink, while the city of Susa was thrown into confusion.
\labelchapt{4}
\passage{Mordecai Seeks Esther's Help}

\chapt{4}
\v{1}When Mordecai learned all that had been done, he tore his garments and clothed himself in sackcloth and ashes. He went into the middle of the city and cried out with a loud and bitter cry. \v{2}He came as far as\fnote{\fbackref{4:2} Lit. \fbib{came to}} the front of the king's gate, because no one was allowed to enter the king's gate clothed in sackcloth. \v{3}In every province where the order of the king and his edict reached, among the Jewish people there was great mourning, fasting, weeping, and lamenting, and many lay down on sackcloth and ashes.

\v{4}When Esther's young women and her eunuchs came and told her, the queen was greatly distressed. She sent clothes for Mordecai to put on so he could take off the sackcloth that he had on, but he would not take them. \v{5}Then Esther summoned Hathach, one of the king's eunuchs, whom he had assigned to her, and she ordered him to go to Mordecai to find out what was happening and why it was happening. \v{6}Hathach went to Mordecai in the city square that was in front of the king's gate. \v{7}Mordecai told him everything that had happened and the exact amount of money that Haman had said he would pay into the king's treasury in order to destroy the Jewish people. \v{8}Mordecai\fnote{\fbackref{4:8} Lit. \fbib{He}} gave Hathach\fnote{\fbackref{4:8} Lit. \fbib{him}} a copy of the written decree ordering the Jews' destruction that had been issued in Susa. Mordecai\fnote{\fbackref{4:8} Lit. \fbib{He}} wanted him to show it to Esther, to explain it to her, and then to instruct her to go in to the king to seek his favor and plead with him for her people.

\v{9}Hathach went and told Esther what Mordecai had said.\fnote{\fbackref{4:9} Lit. \fbib{the words of Mordecai}} \v{10}Then Esther spoke to Hathach, instructing him\fnote{\fbackref{4:10} Lit. \fbib{and she instructed him}} to go back\fnote{\fbackref{4:10} The Heb. lacks \fbib{to go back}} to Mordecai with this message:\fnote{\fbackref{4:10} The Heb. lacks \fbib{with this message:}} \v{11}``Every servant of the king and every person in the king's provinces knows that for any man or woman who goes to the king in the inner court without being summoned there is only\fnote{\fbackref{4:11} The Heb. lacks \fbib{only}} one law---that he be put to death---unless the king holds out the golden scepter to him. Only\fnote{\fbackref{4:11} The Heb. lacks \fbib{Only}} then he will live. For these last\fnote{\fbackref{4:11} The Heb. lacks \fbib{last} } 30 days I've not been summoned to come to the king.''

\v{12}They reported Esther's message to Mordecai.

\v{13}Mordecai told them to reply to Esther, ``Don't suppose that because you are in the palace, you will escape any more than the other Jewish people.\fnote{\fbackref{4:13} Lit. \fbib{than all the Jews}} \v{14}Indeed, if you are silent at this time, relief and deliverance will come to the Jewish people from another place, but you and your father's family will perish. Who knows but that you were brought to the kingdom for a time like this?''

\v{15}Then Esther replied to Mordecai, \v{16}``Go and gather all the Jewish people who are in Susa and fast for me. Don't eat or drink for three days, night or day. Both I and my young women will also fast in the same way, and then I'll go in to the king, even though it's against the law. And if I perish, I perish.''

\v{17}Then Mordecai left and did everything that Esther had ordered him.
\labelchapt{5}
\passage{Esther Goes before the King}

\chapt{5}
\v{1}On the third day, Esther put on her royal attire and stood in the inner courtyard of the palace in front of the king's quarters.\fnote{\fbackref{5:1} Or \fbib{house}} The king was sitting on his royal throne in the throne room, opposite the entrance to the building. \v{2}When the king saw Queen Esther standing in the courtyard, she won his favor, and the king extended to Esther the gold scepter that he was holding. Esther approached and touched the top of the scepter. \v{3}The king asked her, ``What do you want, Queen Esther? What is your request? Even if it's half of the kingdom, it will be given to you.''

\v{4}Esther replied, ``If it pleases the king, let the king and Haman come today to the banquet I've prepared for him.''

\v{5}The king responded, ``Bring Haman quickly so we may do what Esther has requested.'' So the king and Haman went to the banquet that Esther had prepared.

\v{6}While they were drinking wine,\fnote{\fbackref{5:6} Lit. \fbib{at the banquet of wine}} the king asked Esther, ``What's your petition? It will be given to you. What's your request? Up to half of the kingdom, and it will be done.''

\v{7}Esther answered, ``This is my petition and my request: \v{8}If I've found favor with the king and if it pleases the king to grant my petition and to honor my request, let the king and Haman come to the banquet that I'll prepare for them tomorrow, and then I'll do what the king has said.''
\passage{Haman's Plan to Kill Mordecai}

\v{9}Haman went out that day pleased and happy, but when Haman saw Mordecai in the king's gate, and that he did not stand up and tremble in his presence, Haman was furious with Mordecai. \v{10}Haman restrained himself, went to his house, and sent for\fnote{\fbackref{5:10} Lit. \fbib{sent and brought}} his friends and his wife Zeresh. \v{11}Then Haman told them about his splendid wealth, the number\fnote{\fbackref{5:11} Lit. \fbib{multitude}} of his sons, all the ways the king had honored\fnote{\fbackref{5:11} Or \fbib{promoted}} him, and that he had promoted him above all the other officials and ministers\fnote{\fbackref{5:11} Or \fbib{servants}} of the king.

\v{12}Then Haman said, ``Even Queen Esther brought no one except me with the king to the banquet that she held. Furthermore, I (along with the king) have also been invited by her tomorrow. \v{13}But all this does not satisfy me every time I see Mordecai the Jew sitting at the king's gate.''

\v{14}Then Zeresh his wife and all his friends said, ``Have a pole made 50 cubits\fnote{\fbackref{5:14}I.e. about 75 feet; a cubit was about eighteen inches} high, and then in the morning speak to the king and have Mordecai hanged\fnote{\fbackref{5:14} Or \fbib{impaled}} on it. Then go with the king to the banquet happy.'' This advice pleased Haman, and he had the pole made.
\labelchapt{6}
\passage{Haman's Plan Begins to Unravel}

\chapt{6}
\v{1}That night the king could not sleep, so he gave instructions to bring the book of records, the chronicles, and they were read to the king. \v{2}It was found recorded there that Mordecai had reported about Bigthana and Teresh, two of the king's eunuchs who guarded the entrance to the restricted areas of the palace,\fnote{\fbackref{6:2} Lit. \fbib{guarded the threshold}} and that they had conspired to assassinate\fnote{\fbackref{6:2} Lit. \fbib{to send a hand against}} King Ahasuerus. \v{3}So the king asked, ``What honor and distinction was bestowed on Mordecai for this?''

The young men who served the king answered, ``Nothing was done for him.''

\v{4}The king said, ``Who is in the courtyard?'' Now Haman had just entered the outer courtyard of the palace to speak to the king about having Mordecai hanged on the pole he had set up.

\v{5}The king's young men told him, ``Look, Haman is standing in the courtyard.''

The king said, ``Let him come in.''

\v{6}After Haman came in, the king asked him, ``What should be done for the man whom the king desires to honor?''

Haman told himself, ``Whom would the king desire to honor more than me?'' \v{7}Haman answered the king, ``For a man whom the king desires to honor, \v{8}let them bring royal robes that the king has worn and a horse on which the king has ridden, with a royal crown placed on its head. \v{9}Then give the robes and the horse to one of the king's most noble officials. Let them put the robes on the man whom the king desires to honor, and let them put him on the horse in the main\fnote{\fbackref{6:9} The Heb. lacks \fbib{main}} square of the city. Then let them announce in front of him, `This is what is done for the man whom the king desires to honor.'\,''

\v{10}Then the king told Haman, ``Quick! Take the clothes and the horse just as you have suggested and do this for Mordecai the Jew who sits in the king's gate. And don't let anything you've suggested fall through the cracks.''\fnote{\fbackref{6:10} The Heb. lacks \fbib{through the cracks}}

\v{11}So Haman took the clothes and the horse, dressed Mordecai, and put him on the horse in the main\fnote{\fbackref{6:11} The Heb. lacks \fbib{main}} square of the city. He cried out in front of him, ``This is what is done for the man whom the king desires to honor.''

\v{12}Then Mordecai returned to the king's gate, while Haman hurried to his house, mourning and hiding his face.\fnote{\fbackref{6:12} Lit. \fbib{and covering his head}} \v{13}Haman told his wife Zeresh and all his friends everything that had happened to him. His wise friends and his wife Zeresh told him, ``If Mordecai, before whom you have begun to fall, is one of the Jewish people,\fnote{\fbackref{6:13} Lit. \fbib{of the seed of the Jews}} you won't prevail against him. Instead, you will surely fall before him.''

\v{14}While they were still talking to him, the king's eunuchs arrived, and they quickly took him to the banquet that Esther had prepared.
\labelchapt{7}
\passage{Haman is Executed}

\chapt{7}
\v{1}The king and Haman went in to have a drink with Queen Esther. \v{2}On the second day the king again told Esther as they drank wine, ``What's your petition, Queen Esther? It will be given to you. What's your request? Up to half of the kingdom, and it will be done.''

\v{3}Queen Esther answered: ``If I've found favor with you, your majesty, and if it seems good to the king, let my life be given to me as my petition and my people as my request. \v{4}Indeed, I and my people have been sold to be annihilated, killed, and destroyed. If we had just been sold as male and female slaves, I would have kept quiet, because the trouble wouldn't have been sufficient to bother the king.''\fnote{\fbackref{7:4} Or \fbib{no enemy could compensate for this damage to the king}}

\v{5}Then King Ahasuerus asked Queen Esther, ``Who is this, and where is the person who would dare\fnote{\fbackref{7:5} Lit. \fbib{whose heart has filled him}} do this?''

\v{6}Esther replied, ``An adversary and an enemy---it's this wicked Haman!'' So Haman was terrified before the king and the queen. \v{7}The king got up from the banquet in anger and went out to the palace garden, while Haman stood there begging Queen Esther to spare his life,\fnote{\fbackref{7:7} Lit. \fbib{the queen for his life}} because he realized that the king intended to harm him.\fnote{\fbackref{7:7} Lit. \fbib{to bring evil on him}}

\v{8}When the king returned to the banquet hall from the palace garden, Haman was prostrate on the couch where Esther was. The king asked, ``Will this man\fnote{\fbackref{7:8} Lit. \fbib{he}} even assault the queen with me in the house?'' The king had no sooner spoken than they covered Haman's face. \v{9}Then Harbonah, one of the eunuchs attending the king, observed, ``Look there! A pole is standing 50 cubits\fnote{\fbackref{7:9} I.e. about 75 feet; a cubit was about eighteen inches} high at Haman's house that he prepared for Mordecai, whose report benefitted\fnote{\fbackref{7:9} Lit. \fbib{who spoke good for}} the king!''

The king said, ``Hang\fnote{\fbackref{7:9} Or \fbib{Impale}} him on it.'' \v{10}So they hanged\fnote{\fbackref{7:10} Or \fbib{impaled}} Haman on the pole he had set up for Mordecai, and then the king's anger subsided.
\labelchapt{8}
\passage{The Promotion of Mordecai}

\chapt{8}
\v{1}That day King Ahasuerus gave Queen Esther the property\fnote{\fbackref{8:1} Lit. \fbib{house}} of Haman, the enemy of the Jewish people, and Mordecai came into the king's presence because Esther had told him how Mordecai\fnote{\fbackref{8:1} Lit. \fbib{he}} was related to her. \v{2}The king took off his signet ring that he had taken from Haman and gave it to Mordecai. Esther then put Mordecai in charge of Haman's property.\fnote{\fbackref{8:2} Lit. \fbib{house}}
\passage{Esther Asks that the Jewish People be Spared}

\v{3}Then Esther spoke to the king again and fell at his feet. She wept and pleaded with him for mercy to overturn the evil plan devised\fnote{\fbackref{8:3} The Heb. lacks \fbib{devised}} by Haman the Agagite and his plot against the Jewish people. \v{4}The king extended the golden scepter to Esther, and she got up and stood before the king. \v{5}She said, ``If it pleases the king, and if I've found favor with him, and if the matter is proper in the king's opinion, and if I'm pleasing to the king, let an order be issued\fnote{\fbackref{8:5} Lit. \fbib{let it be written}} revoking the letters devised by Hammedatha the Agagite's son Haman, which ordered\fnote{\fbackref{8:5} Lit. \fbib{were written}} the destruction of the Jewish people throughout the king's provinces. \v{6}Indeed, how can I bear to see this disaster happen to my people? How can I bear to see the destruction of my kinsmen?''

\v{7}King Ahasuerus told Queen Esther and Mordecai the Jew, ``Look, I've given Haman's property\fnote{\fbackref{8:7} Lit. \fbib{house}} to Esther, and they have hanged\fnote{\fbackref{8:7} Or \fbib{impaled}} him on the pole because he tried to harm\fnote{\fbackref{8:7} Lit. \fbib{sent his hand against}} the Jewish people. \v{8}Now, in the name of the king, you write what seems good to you concerning the Jewish people, and seal it with the king's signet ring, for a document written in the king's name and sealed with the king's signet ring cannot be revoked.''

\v{9}The king's scribes were summoned at that time, on the twenty-third day of the third month, which is the month Sivan, and everything that Mordecai commanded the Jewish people, the regional authorities,\fnote{\fbackref{8:9} Lit. \fbib{satraps}; Persian government officials similar in authority to a governor} the governors, and the provincial officials of the 127 provinces from India to Cush\fnote{\fbackref{8:9} I.e. the upper Nile region} was written down for each province according to its script, for each people according to their language, and for the Jewish people according to their script and language. \v{10}He wrote in the name of King Ahasuerus and sealed it with the king's signet ring. He sent the letters by couriers on horseback, riding steeds especially bred for the king.\fnote{\fbackref{8:10} Or \fbib{sired by the royal stud}; i.e. the horses were strong and fast and specially bred for the task}

\v{11}What the king granted the Jewish people in every town was the right\fnote{\fbackref{8:11} The Heb. lacks \fbib{the right}} to assemble and defend themselves,\fnote{\fbackref{8:11} Lit. \fbib{and stand; or their lives}} to annihilate, kill, and destroy every armed force of a people or a province that was hostile to them, including children and women, and to plunder their property.\fnote{\fbackref{8:11} Lit. \fbib{spoil}} \v{12}Throughout all the provinces of King Ahasuerus, the one day for the Jewish people to do this was the thirteenth day of the twelfth month, which is the month Adar. \v{13}A copy of the document was to be issued as law in each and every province and published for all people, indicating that the Jewish people were to be ready to take vengeance on their enemies on that day. \v{14}The couriers, mounted on the royal steeds, left quickly, urged on by the king's command. The edict was also issued in Susa the capital.
\passage{The Jewish People Celebrate the King's Edict}

\v{15}Mordecai left the king's presence in royal robes of blue and white, wearing a large golden crown and a purple robe made of fine linen; and the city of Susa shouted with joy. \v{16}For the Jewish people, there was light and joy, gladness and honor. \v{17}In each and every province, and in each and every city, in the places where the king's order and edict reached, there was gladness and joy among the Jewish people, along with a festival and a holiday. Many of the people of the land became\fnote{\fbackref{8:17} Or \fbib{professed to be}} Jews, because they had come to fear the Jewish people.
\labelchapt{9}
\passage{The Jewish People Defeat Their Enemies}

\chapt{9}
\v{1}On the thirteenth day of the twelfth month, which is the month Adar, when the king's order and edict was about to be carried out, on the day when the enemies of the Jewish people expected to prevail over them, things were turned around so that the Jewish people themselves prevailed over those who hated them.

\v{2}The Jewish people assembled in their towns throughout the provinces of King Ahasuerus to strike out against those who intended to harm them, and no one could oppose them because all the people had come to fear the Jews.\fnote{\fbackref{9:2} Lit. \fbib{them}} \v{3}All the provincial officials, the regional authorities,\fnote{\fbackref{9:3} Lit. \fbib{satraps}; Persian government officials similar in authority to a governor} the governors, and those doing the king's work supported the Jewish people because the fear of Mordecai had come over\fnote{\fbackref{9:3} Lit. \fbib{fallen on}} them. \v{4}Indeed, Mordecai was a powerful\fnote{\fbackref{9:4} Or \fbib{great}} official in the palace and his fame spread throughout the provinces. Indeed, the man Mordecai grew more and more powerful.\fnote{\fbackref{9:4} Or \fbib{greater and greater}}

\v{5}The Jewish people struck down all their enemies with the sword, killing and destroying them, and they did with their enemies as they pleased. \v{6}In Susa the capital the Jewish people killed and destroyed 500 people. \v{7}They killed Parshandatha, Dalphon, Aspatha, \v{8}Poratha, Adalia, Aridatha, \v{9}Parmashta, Arisai, Aridai, and Vaizatha, \v{10}the ten sons of Hammedatha's son Haman, the enemy of the Jewish people, but they did not lay their hands on the spoils.

\v{11}On that day the number of those slain in Susa the capital was reported to the king. \v{12}The king told Queen Esther, ``In Susa the capital the Jewish people have killed and destroyed 500 people, including Haman's ten sons. What have they done in the rest of the king's provinces? Now what's your petition? It will be given to you. What's your further request? It will be done.''

\v{13}Then Esther said, ``If it pleases the king, let it also be granted to the Jewish people in Susa to do tomorrow what the edict allowed them to do today,\fnote{\fbackref{9:13} Lit. \fbib{according to today's edict}} and let Haman's ten sons be hanged\fnote{\fbackref{9:13} Or \fbib{impaled}} on poles.''

\v{14}The king said, ``Let this be done.'' So an edict was issued in Susa, and Haman's ten sons were hanged\fnote{\fbackref{9:14} Or \fbib{impaled}} on poles. \v{15}The Jewish people in Susa assembled again on that day, the fourteenth of Adar, and they killed 300 people in Susa, but they did not lay their hands on the spoils.
\passage{The Festival of Purim is Celebrated}

\v{16}The rest of the Jewish people in the king's provinces assembled to defend\fnote{\fbackref{9:16} Lit. \fbib{stand; or their lives}} themselves, and they gained relief from their enemies, killing 75,000 of those who hated them. But they did not lay their hands on the spoils. \v{17}They did this on the thirteenth day of Adar and rested on the fourteenth day, making it a day of feasting and joy. \v{18}The Jewish people in Susa assembled on the thirteenth day and again on the fourteenth, and then rested on the fifteenth day and made it a day of feasting and joy. \v{19}Therefore the Jewish people in the rural areas who live in unwalled towns make the fourteenth day of the month Adar a holiday for joy and feasting, and people send presents\fnote{\fbackref{9:19} Or \fbib{portions of food}} to one another.
\passage{Official Instructions for Celebrating Purim}

\v{20}Mordecai wrote these instructions and sent letters to all the Jewish people in all the provinces of King Ahasuerus, both near and far, \v{21}establishing that they should celebrate the fourteenth and fifteenth days of the month Adar every year, \v{22}as the days on which the Jewish people enjoyed relief\fnote{\fbackref{9:22} Or \fbib{the Jews rested}} from their enemies. It was a month when things turned around for them, from sorrow to joy and from mourning to a holiday. They were to celebrate these days as days of feasting and joy, and they were to send presents\fnote{\fbackref{9:22} Or \fbib{portions of food}} to one another and gifts to the poor. \v{23}So the Jewish people made a tradition\fnote{\fbackref{9:23} Lit. \fbib{the Jews accepted}} out of what they had begun to do and of what Mordecai had written to them, \v{24}since Hammedatha's son Haman, the enemy of the Jewish people, had plotted against the Jewish people to destroy them, and he had cast the pur (that is, the lot) to determine when\fnote{\fbackref{9:24} The Heb. lacks \fbib{to determine when}} to confuse and destroy them.

\v{25}But when Esther came before the king, he ordered through a letter that the evil plot that Haman\fnote{\fbackref{9:25} Lit. \fbib{he}} had devised against the Jewish people be rescinded,\fnote{\fbackref{9:25} Lit. \fbib{be turned back on his own head}} and that he and his sons be hanged on poles. \v{26}Therefore these days were called Purim, from the word pur. Because of all that was written in this letter, because of what they experienced in this matter, and because of what happened to them, \v{27}the Jewish people established this celebration, making it a tradition\fnote{\fbackref{9:27} Lit. \fbib{people accepted it}} for themselves, for their descendants, and for all who joined with them\fnote{\fbackref{9:27} I.e. those who became Jews} that they should not fail to observe these two days each year, based on the written instructions, and at the prescribed time. \v{28}These days should be remembered and observed in every generation by each family in every province and town. These days of Purim should not be neglected by\fnote{\fbackref{9:28} Lit. \fbib{should not pass by}} the Jewish people, and that they should not be forgotten by their descendants.
\passage{Queen Esther Confirms the Instructions for Purim}

\v{29}Queen Esther, the daughter of Abihail, and Mordecai the Jew wrote with full authority confirming this second letter about Purim. \v{30}Letters containing wishes for peace and stability were sent to all the Jewish people, to the 127 provinces of Ahasuerus' kingdom, \v{31}establishing these days of Purim at the prescribed time, just as Mordecai the Jew and Queen Esther had established, and just as the Jewish people\fnote{\fbackref{9:31} Lit. \fbib{they}} had established for themselves and for their descendants. The letter included instructions for their fasting\fnote{\fbackref{9:31} Lit. \fbib{descendants, instructions for their fasting}} and lamentations. \v{32}The order of Esther established these instructions for Purim, and it was officially recorded.\fnote{\fbackref{9:32} Lit. \fbib{recorded in a record}}
\labelchapt{10}
\passage{Other Things King Ahasuerus Did}

\chapt{10}
\v{1}King Ahasuerus imposed tribute\fnote{\fbackref{10:1} I.e. a kind of tax} on the land and on the islands of the sea. \v{2}Now as to all the powerful and great deeds of Ahasuerus, along with an exact statement about the high position\fnote{\fbackref{10:2} Lit. \fbib{greatness}} of Mordecai to which the king promoted him, these things are written in the Book of the Chronicles of the Kings of Media and Persia, are they not? \v{3}Indeed, Mordecai the Jew was second in authority only\fnote{\fbackref{10:3} Lit. \fbib{second to King}} to King Ahasuerus and was a powerful official\fnote{\fbackref{10:3} Or \fbib{great}} among the Jewish people. Mordecai\fnote{\fbackref{10:3} Lit. \fbib{He}} was accepted favorably by his many kinsmen, and he sought the good of his countrymen and spoke out for the welfare of all his people.\fnote{\fbackref{10:3} Lit. \fbib{his see}}
