\bookheader{Song of Songs}
\labelbook{Song}

\bookpretitle{The Most Beautiful}
\booktitle{Song of Songs}

\labelchapt{1}
\passage{Title}

\chapt{1}
\v{1}The Most Beautiful Song by Solomon.
\passage{The Loved One}

\begin{poetry}
\poeml \v{2}Let him kiss me over and over again!\fnote{\fbackref{1:2} Lit. \fbib{me with the kisses of his mouth}} \\
\poemll    Your love is better than wine. \\
\poeml \v{3}The fragrance of your perfumed oil is wonderful. \\
\poemll    Your name is perfume poured out. \\
\poemlll       Therefore the young women love you. \\
\poeml \v{4}Take me with\fnote{\fbackref{1:4} Lit. \fbib{Draw me after}} you! Let's run away! \\
\poemll    Let the king bring me into his private chambers.
\passage{The Young Women}
\poeml The daughters\fnote{\fbackref{1:4} Or \fbib{The young women}} of Jerusalem\fnote{\fbackref{1:4} The Heb. lacks \fbib{daughters of Jerusalem} (cf. v. 5)} will rejoice and be happy for you. \\
\poemll    We will value your love more than wine. \\
\poemlll       They love you appropriately.
\passage{The Loved One}
\poeml \v{5}The daughters\fnote{\fbackref{1:5} Or \fbib{The young women}} of Jerusalem, I'm dark and lovely \\
\poemll    like the tents of Kedar, \\
\poemlll       like the curtains of Solomon. \\
\poeml \v{6}Don't stare at me because I'm dark; \\
\poemll    the sun has tanned me. \\
\poeml My mother's sons were angry with me. \\
\poemll    They made me the caretaker of the vineyards, \\
\poemlll       but I didn't take care of my own vineyard. \\
\poeml \v{7}Tell me, you whom I love, \\
\poemll    where do you graze your flock? \\
\poemlll       Where do you make your flock lie down\fnote{\fbackref{1:6} Lit. \fbib{where do you make to lie down}} at noon? \\
\poeml Why should I be considered a veiled woman\fnote{\fbackref{1:6} I.e. a prostitute} \\
\poemll    beside the flocks of your companions?
\passage{The Lover}
\poeml \v{8}If you don't know, most beautiful of women, \\
\poemll    go out after the flock and graze your young goats beside the shepherd's tents. \\
\poeml \v{9}My darling, I compare you to a\fnote{\fbackref{1:9} Lit. \fbib{my}} mare \\
\poemll    among Pharaoh's chariots. \\
\poeml \v{10}Your cheeks are lovely with ornaments, \\
\poemll    your neck with strings of jewels.
\passage{The Young Women}
\poeml \v{11}We will make ornaments of gold for you, \\
\poemll    accented with silver.
\passage{The Loved One}
\poeml \v{12}While the king was sitting at his table, \\
\poemll    my perfume sent forth its fragrance. \\
\poeml \v{13}My beloved is to me a pouch of myrrh\fnote{\fbackref{1:13} I.e. a fragrant spice used as a perfume} \\
\poemll    that lies between my breasts all night. \\
\poeml \v{14}My beloved is to me a cluster of henna\fnote{\fbackref{1:14} I.e. an aromatic shrub used cosmetically} blossoms \\
\poemll    in the vineyards of En-gedi.
\passage{The Lover}
\poeml \v{15}Look at you! You are beautiful, my darling. \\
\poemll    Look at you! You are so beautiful. \\
\poemlll       Your eyes are doves.
\passage{The Loved One}
\poeml \v{16}Look at you! You are handsome, my beloved, truly lovely. \\
\poemll    How lush is our couch. \\
\poeml \v{17}The beams of our house are cedar,\fnote{\fbackref{1:17} I.e. a genus of coniferous evergreen in the family \fbib{Pinaceae}; and so throughout the book} \\
\poemll    our rafters are pine.
\labelchapt{2}
\poeml \chapt{2}
\v{1}I'm a flower\fnote{\fbackref{2:1} Or \fbib{crocus}} from Sharon, \\
\poeml a lily of the valleys.
\passage{The Lover}
\poeml \v{2}Like a lily among thorns, \\
\poemll    so is my darling among the young women.
\passage{The Loved One}
\poeml \v{3}Like an apple\fnote{\fbackref{2:3} Or \fbib{apricot}} tree among the trees of the forest, \\
\poemll    so is my beloved among the young men. \\
\poeml In his shade I take delight and sit down, \\
\poemll    and his fruit is sweet to my taste.\fnote{\fbackref{2:3} Lit. \fbib{palate}} \\
\poeml \v{4}He has brought me to the banquet hall, \\
\poemll    and his banner over me is love. \\
\poeml \v{5}Sustain me with raisin cakes, \\
\poemll    refresh me with apples,\fnote{\fbackref{2:5} Or \fbib{apricots}} \\
\poemlll       for I'm weak with love. \\
\poeml \v{6}I wish that his left hand were under my head, \\
\poemll    and that his right hand were embracing me! \\
\poeml \v{7}Swear to me, young women of Jerusalem, \\
\poemll    by the gazelles or by the does of the field, \\
\poeml that you won't awaken or arouse love \\
\poemll    before its proper time!\fnote{\fbackref{2:7} Lit. \fbib{until it pleases}} \\
\poeml \v{8}The voice of my beloved! \\
\poeml Look! He's coming, \\
\poemll    leaping over the mountains, \\
\poemlll       bounding over the hills. \\
\poeml \v{9}My beloved is like a gazelle or a young stag. \\
\poeml Look, there he stands behind our wall, \\
\poemll    looking through the windows, \\
\poemlll       gazing through the lattice.
\passage{The Lover}
\poeml \v{10}My beloved spoke to me: \\
\poemll    ``Get up, my darling, my beautiful one, and come on. \\
\poeml \v{11}Look! Winter is past. \\
\poemll    The rain is over and gone. \\
\poeml \v{12}Blossoms have appeared in the land. \\
\poemll    The season of songbirds\fnote{\fbackref{2:12} Or \fbib{for pruning}} has arrived, \\
\poemlll       and cooing of turtledoves is heard in our land. \\
\poeml \v{13}The fig tree has produced its fruit,\fnote{\fbackref{2:13} Lit. \fbib{has ripened}} \\
\poemll    the grapevines have blossomed and exude their fragrance. \\
\poeml ``Get up, my darling, my beautiful one, and come on. \\
\poeml \v{14}My dove, in the hidden places of the rocks, \\
\poemll    in the secret places of the cliffs, \\
\poeml show me your form, and let me hear your voice. \\
\poemll    For your voice is pleasant, \\
\poemlll       and your shape is lovely. \\
\poeml \v{15}Catch the foxes for us, \\
\poemll    the little foxes that destroy the vineyards, \\
\poemlll       our vineyards that are in bloom.''
\passage{The Loved One}
\poeml \v{16}My beloved belongs to me and I belong to him. \\
\poemll    He is the one who shepherds his flock among the lilies. \\
\poeml \v{17}Until the day breaks\fnote{\fbackref{2:17} Or \fbib{until the cool of the day}; lit. \fbib{until the day breathes}} and the shadows flee, \\
\poemll    turn around, my beloved, \\
\poeml and be like a gazelle or a young stag \\
\poemll    on the rugged mountains.\fnote{\fbackref{2:17} Or \fbib{the mountains of Bether}}
\end{poetry}
\labelchapt{3}
\passage{The Loved One}

\begin{poetry}
\poeml \chapt{3}
\v{1}Night after night on my bed, \\
\poeml I sought the one I love; \\
\poeml I sought him, but didn't find him. \\
\poeml \v{2}I'll get up and go all around the city, \\
\poemll    throughout the streets, and in the squares. \\
\poeml I'll seek the one I love. \\
\poemll    I sought him, but didn't find him. \\
\poeml \v{3}The watchmen who go all around the city found me. \\
\poemll    I asked,\fnote{\fbackref{3:3} The Heb. lacks \fbib{I asked}} ``Have you seen the one I love?'' \\
\poeml \v{4}I had just passed them \\
\poemll    when I found the one I love. \\
\poeml I held him and wouldn't let him go \\
\poemll    until I brought him to my mother's house, \\
\poemlll       to the room of the one who conceived me. \\
\poeml \v{5}Swear to me, young women of Jerusalem, \\
\poemll    by the gazelles or by the does of the field, \\
\poeml that you won't awaken or arouse love \\
\poemll    before its proper time!\fnote{\fbackref{3:5} Lit. \fbib{until it pleases}}
\passage{The Lover Arrives}
\poeml \v{6}What\fnote{\fbackref{3:6} Or \fbib{Who}} is this coming up from the desert \\
\poemll    like columns of smoke, \\
\poeml perfumed with myrrh\fnote{\fbackref{3:6} Myrrh was a fragrant spice used as a perfume.} and incense\fnote{\fbackref{3:6} Or \fbib{frankincense}} \\
\poemll    from all the fragrant powders of the merchant? \\
\poeml \v{7}Look! It's Solomon's sedan chair,\fnote{\fbackref{3:7} Or \fbib{palanquin}; i.e. a portable compartment for carrying important people.} \\
\poemll    with 60 of the best soldiers in Israel\fnote{\fbackref{3:7} Lit. \fbib{mighty men from the mighty men of Israel}} surrounding it. \\
\poeml \v{8}All of them are wearing swords and are \\
\poemll    experienced in battle. \\
\poeml Each has his sword on his thigh, \\
\poemll    ready for the terror of the night.\fnote{\fbackref{3:8} Lit. \fbib{thigh, from the terror of the night}} \\
\poeml \v{9}King Solomon made the sedan chair for himself \\
\poemll    from the trees of Lebanon. \\
\poeml \v{10}He made its posts of silver, \\
\poemll    its back of gold. \\
\poeml Its seat was purple, \\
\poemll    and its interior was lovingly inlaid \\
\poemlll       by the young women of Jerusalem. \\
\poeml \v{11}Come out, young women of Zion, \\
\poemll    and see King Solomon with the crown \\
\poeml with which his mother crowned him \\
\poemll    on his wedding day--- \\
\poemlll       his day of great delight.\fnote{\fbackref{3:11} Lit. \fbib{day of the rejoicing of his heart}}
\end{poetry}
\labelchapt{4}
\passage{The Lover}

\begin{poetry}
\poeml \chapt{4}
\v{1}Look at you! You are beautiful, my darling. \\
\poeml Look at you! You are so beautiful. \\
\poeml Your eyes behind your veil are doves, \\
\poemll    your hair is like a flock of goats \\
\poemlll       coming down from Mt. Gilead. \\
\poeml \v{2}Your teeth are like a flock of sheep about to be sheared,\fnote{\fbackref{4:2} Or \fbib{like sheared sheep}} \\
\poemll    who are coming up from being washed.\fnote{\fbackref{4:2} Lit. \fbib{from washing}} \\
\poeml All of them are twins, \\
\poemll    not one has lost\fnote{\fbackref{4:2} Lit. \fbib{been bereaved of}; i.e. her teeth match} her young. \\
\poeml \v{3}Your lips are like a scarlet thread, \\
\poemll    and your mouth is lovely. \\
\poeml Behind your veil, \\
\poemll    your temple is like a slice of pomegranate. \\
\poeml \v{4}Your neck is like the tower of David, \\
\poemll    built with rows of stones. \\
\poeml A thousand shields are hung upon it, \\
\poemll    all the shields of the warriors. \\
\poeml \v{5}Your two breasts are like two fawns, \\
\poemll    twins of a gazelle grazing among the lilies. \\
\poeml \v{6}Until the day breaks\fnote{\fbackref{4:6} Or \fbib{until the cool of the day;} lit. \fbib{until the day breathes};} and the shadows flee, \\
\poemll    I'll go to the mountain of myrrh\fnote{\fbackref{4:6} Myrrh was a fragrant spice used as a perfume.} \\
\poemlll       and to the hill of frankincense.\fnote{\fbackref{4:6} Frankincense was a fragrant spice used to make incense and perfume.} \\
\poeml \v{7}My darling, you are altogether beautiful \\
\poemll    and there is no blemish in you. \\
\poeml \v{8}Come with me from Lebanon, my bride, \\
\poemll    come with me from Lebanon. \\
\poeml May you journey from the top of Amana, \\
\poemll    from the tops of Senir and Hermon, \\
\poeml from the dens of lions, \\
\poemll    from the mountain lairs of leopards. \\
\poeml \v{9}You have made my heart beat faster, my sister, my\fnote{\fbackref{4:9} The Heb. lacks \fbib{my}} bride. \\
\poemll    You have made my heart beat faster \\
\poemlll       with one glance of your eyes, \\
\poemll    with one strand of your necklace. \\
\poeml \v{10}How beautiful is your love, my sister, my\fnote{\fbackref{4:10} The Heb. lacks \fbib{my}} bride. \\
\poemll    How much better is your love than wine, \\
\poeml and the fragrance of your perfume \\
\poemll    than all kinds of spices. \\
\poeml \v{11}Your lips drip honey, my\fnote{\fbackref{4:11} The Heb. lacks \fbib{my}} bride; \\
\poemll    milk and honey are under your tongue. \\
\poeml The scent of your garments \\
\poemll    is like the fragrance of Lebanon. \\
\poeml \v{12}My sister, my\fnote{\fbackref{4:12} The Heb. lacks \fbib{my}} bride, is a locked garden \\
\poemll    a locked rock garden, a sealed up spring. \\
\poeml \v{13}Your shoots are an orchard\fnote{\fbackref{4:13} Or \fbib{a park}} of pomegranates, \\
\poemll    with choice fruit, henna with nard, \\
\poeml \v{14}nard and saffron, \\
\poemll    calamus and cinnamon, \\
\poeml with all the trees of frankincense, \\
\poemll    along with myrrh and aloes, and all the finest spices.\fnote{\fbackref{4:14} All the spices listed in verses 13 and 14 were used for perfume.} \\
\poeml \v{15}You are a garden spring, \\
\poemll    a well of fresh water, \\
\poemlll       streams flowing from Lebanon.
\passage{The Loved One}
\poeml \v{16}Awake, north wind, and come, south wind. \\
\poemll    Make my garden breathe out, \\
\poemlll       let its fragrance\fnote{\fbackref{4:16} Or \fbib{its spices}} flow. \\
\poeml Let my beloved come into his garden, \\
\poemll    and let him eat its choicest fruits.
\end{poetry}
\labelchapt{5}
\passage{The Lover}

\begin{poetry}
\poeml \chapt{5}
\v{1}I've come into my garden, my sister, my\fnote{\fbackref{5:1} The Heb. lacks \fbib{my}} bride; \\
\poeml I've gathered my myrrh with my spices. \\
\poeml I've eaten my honeycomb with my honey. \\
\poeml I've drunk my wine with my milk. \\
\poeml Eat, friends! \\
\poemll    Drink and become drunk with love.
\passage{The Loved One}
\poeml \v{2}I was asleep, but my heart was awake. \\
\poemll    There's a sound! My beloved is knocking.
\passage{The Lover}
\poeml ``Open up for me, my sister, my darling, \\
\poemll    my dove, my perfect one. \\
\poeml For my head is drenched with dew, \\
\poemll    my hair with the dampness of the night.''
\passage{The Loved One}
\poeml \v{3}``I've taken off my clothes\fnote{\fbackref{5:3} Or \fbib{my tunic}}--- \\
\poemll    am I supposed to put them on again?\fnote{\fbackref{5:3} Lit. \fbib{How shall I put them on again?}} \\
\poeml I've washed my feet--- \\
\poemll    am I supposed to\fnote{\fbackref{5:3} Lit. \fbib{How shall I}} get them dirty again?'' \\
\poeml \v{4}My beloved reached out his hand for the latch.\fnote{\fbackref{5:4} Or \fbib{keyhole}} \\
\poemll    My feelings for him were aroused. \\
\poeml \v{5}I got up to open the door,\fnote{\fbackref{5:5} The Heb. lacks \fbib{the door}} \\
\poemll    and my hands dripped with myrrh, \\
\poeml my fingers with liquid myrrh, \\
\poemll    on the handle of the lock. \\
\poeml \v{6}I opened the door\fnote{\fbackref{5:6} The Heb. lacks \fbib{the door}} for my beloved, \\
\poemll    but my beloved had turned away; he was gone! \\
\poeml My very life went out when he departed.\fnote{\fbackref{5:6} Or \fbib{when he spoke}} \\
\poemll    I searched for him, \\
\poemlll       but couldn't find him. \\
\poemll    I called out to him, \\
\poemlll       but he didn't answer. \\
\poeml \v{7}The watchmen making their\fnote{\fbackref{5:7} The Heb. lacks \fbib{their}} rounds \\
\poemll    through the city found me. \\
\poeml They beat me, they bruised me. \\
\poemll    Those guarding the walls took my cloak\fnote{\fbackref{5:7} Or \fbib{shawl}} from me. \\
\poeml \v{8}I charge you, young women of Jerusalem, \\
\poemll    ``If you find my beloved, what are you to tell him? \\
\poemlll       Tell him\fnote{\fbackref{5:7} The Heb. lacks \fbib{Tell him}} that I'm weak with love.''
\passage{The Young Women}
\poeml \v{9}What is so special about your beloved,\fnote{\fbackref{5:9} Or \fbib{How is your beloved better than other beloveds}} \\
\poemll    most beautiful of women? \\
\poeml What is so special about your beloved, \\
\poemll    that you charge us like this?
\passage{The Loved One}
\poeml \v{10}My beloved is dazzling, \\
\poemll    with a dark and healthy complexion, \\
\poemlll       outstanding among ten thousand. \\
\poeml \v{11}His head is pure gold, \\
\poemll    his hair is wavy, black like a raven. \\
\poeml \v{12}His eyes are like doves \\
\poemll    by streams of water, \\
\poeml bathed in milk, \\
\poemll    mounted like jewels. \\
\poeml \v{13}His cheeks are like beds of spices, \\
\poemll    like towers of perfume. \\
\poeml His lips are lilies, \\
\poemll    dripping with liquid myrrh. \\
\poeml \v{14}His hands are rods of gold, \\
\poemll    set with beryl. \\
\poeml His stomach is carved ivory, \\
\poemll    inlaid with sapphires.\fnote{\fbackref{5:14} Or \fbib{lapis lazuli}} \\
\poeml \v{15}His legs are pillars of marble \\
\poemll    set on bases of pure gold. \\
\poeml His appearance is like Lebanon, \\
\poemll    choice like its cedars. \\
\poeml \v{16}His mouth\fnote{\fbackref{5:16} Lit. \fbib{palate}} is sweetness, \\
\poemll    and all of him is desirable. \\
\poeml This is my beloved, this is my friend, \\
\poemll    young women of Jerusalem!
\end{poetry}
\labelchapt{6}
\passage{The Young Women}

\begin{poetry}
\poeml \chapt{6}
\v{1}Where did your beloved go, \\
\poeml most beautiful of women? \\
\poeml Where did your beloved turn, \\
\poeml so we may look for him with you?
\passage{The Loved One}
\poeml \v{2}My beloved has gone down to his garden, \\
\poemll    to the beds of spices, \\
\poeml to graze his flock\fnote{\fbackref{6:2} The Heb. lacks \fbib{his flock}} in the gardens \\
\poemll    and gather lilies. \\
\poeml \v{3}I belong to my beloved, and my beloved belongs to me. \\
\poemll    He is the one who grazes his flock\fnote{\fbackref{6:3} The Heb. lacks \fbib{his flock}} among the lilies.
\passage{The Lover}
\poeml \v{4}You are beautiful, my darling, like Tirzah,\fnote{\fbackref{6:4} I.e. a prominent city in the northern kingdom of Israel} \\
\poemll    lovely like Jerusalem, \\
\poemlll       as awesome as an army with banners. \\
\poeml \v{5}Turn your eyes from me, \\
\poemll    for they excite me.\fnote{\fbackref{6:5} Or \fbib{overwhelm me}} \\
\poeml Your hair is like a flock of goats \\
\poemll    coming down from Mt. Gilead. \\
\poeml \v{6}Your teeth are like a flock of ewes \\
\poemll    coming up from being washed.\fnote{\fbackref{6:6} Lit. \fbib{from washing}} \\
\poeml All of them are twins, \\
\poemll    not one has lost\fnote{\fbackref{6:6} Lit. \fbib{been bereaved of}; i.e. her teeth match} her young. \\
\poeml \v{7}Your temple\fnote{\fbackref{6:7} Or \fbib{brow}} behind your veil \\
\poemll    is like a slice of pomegranate. \\
\poeml \v{8}There are sixty queens and eighty mistresses,\fnote{\fbackref{6:8} Or \fbib{concubines}; i.e. secondary wives} \\
\poemll    and too many young women to count, \\
\poeml \v{9}but my dove, my perfect one, is unique. \\
\poeml She's unique to her mother, \\
\poemll    she's pure to the one who gave birth to her. \\
\poeml Young women see her and call her blessed, \\
\poemll    queens and mistresses praise her.
\passage{The Young Women}
\poeml \v{10}Who is this who appears like the dawn, \\
\poemll    beautiful as the moon, \\
\poeml bright as the sun, \\
\poemll    awesome as an army with banners?
\passage{The Loved One}
\poeml \v{11}I went down to the walnut orchard, \\
\poemll    to look at the green sprouts in the valley, \\
\poeml to see whether the vine had budded, \\
\poemll    whether the pomegranates had blossomed. \\
\poeml \v{12}Before I knew it,\fnote{\fbackref{6:12} Lit. \fbib{I did not know}; i.e. was daydreaming} I imagined myself \\
\poemll    among the chariots of my noble people.\fnote{\fbackref{6:12} Or \fbib{of Amminadib}}
\passage{The Young Women}
\poeml \v{13}\fnote{\fbackref{6:13} This v. is 7:1 in MT}Return, return, Shulammite, \\
\poemll    return, return, so we may look at you!
\passage{The Lover}
\poeml Why should you look at the Shulammite,\fnote{\fbackref{6:13} Or \fbib{look at Shulamit}} \\
\poemll    like you watch\fnote{\fbackref{6:13} The Heb. lacks \fbib{you watch}} the dance of the two camps?\fnote{\fbackref{6:13} Or \fbib{the dance of Mahanaim}}
\labelchapt{7}
\poeml \chapt{7}
\v{1}\fnote{\fbackref{7:1} This v. is 7:2 in MT}How beautiful are your feet in sandals, \\
\poeml noble lady.\fnote{\fbackref{7:1} Or \fbib{prince's daughter}} \\
\poeml The curves of your thighs are like ornaments, \\
\poemll    the work of a skilled artist's hands. \\
\poeml \v{2}Your navel is a rounded goblet \\
\poemll    that never lacks mixed wine. \\
\poeml Your abdomen is a bundle of wheat, \\
\poemll    enclosed by lilies. \\
\poeml \v{3}Your two breasts are like two fawns, \\
\poemll    twins of a gazelle. \\
\poeml \v{4}Your neck is like a tower of ivory. \\
\poemll    Your eyes are like the\fnote{\fbackref{7:4} The Heb. lacks \fbib{like the}} pools in Heshbon, \\
\poemlll       beside the gate of Beth-rabbim. \\
\poeml Your nose is like the tower of Lebanon, \\
\poemll    which faces Damascus. \\
\poeml \v{5}Your head crowns you\fnote{\fbackref{7:5} Lit. \fbib{is on you}} like Mount Carmel. \\
\poemll    Your flowing locks\fnote{\fbackref{7:5} Lit. \fbib{the hair of your head}} are like purple, \\
\poemlll       and a king could be captured in the dangling tresses. \\
\poeml \v{6}How beautiful and lovely you are, \\
\poemll    you are love with its exquisite delights.\fnote{\fbackref{7:6} Or \fbib{my love with exquisite delights}; Lit. \fbib{Love with exquisite delights}} \\
\poeml \v{7}Your stature\fnote{\fbackref{7:7} Lit. \fbib{This stature of yours}} is like a palm tree, \\
\poemll    and your breasts are like clusters of fruit.\fnote{\fbackref{7:7} The Heb. lacks \fbib{of fruit}} \\
\poeml \v{8}I told myself, ``I'll go up the palm tree, \\
\poemll    and take hold of its fruit. \\
\poeml May your breasts be like clusters of the vine,\fnote{\fbackref{7:8} I.e. grapes} \\
\poemll    and the smell of your breath like apples.\fnote{\fbackref{7:8} Or \fbib{apricots}} \\
\poeml \v{9}May your mouth\fnote{\fbackref{7:9} Lit. \fbib{palate}} be like good wine.
\passage{The Loved One}
\poeml May it go down smoothly to my beloved, \\
\poemll    gliding gently over the lips of the sleeping ones. \\
\poeml \v{10}I belong to my beloved, \\
\poemll    and his desire is for me. \\
\poeml \v{11}Come, my beloved. \\
\poemll    Let us go out to the country, \\
\poemlll       let us spend the night in the villages.\fnote{\fbackref{7:10} Or \fbib{night among the henna blossoms}} \\
\poeml \v{12}Let us go early to the vineyards, \\
\poemll    let us see whether the vine has budded, \\
\poeml whether the blossom has opened, \\
\poemll    whether the pomegranates have bloomed. \\
\poemlll       There I'll give you my love. \\
\poeml \v{13}The mandrakes give off their\fnote{\fbackref{7:13} The Heb. lacks \fbib{their}} fragrance, \\
\poemll    and over our doors are all the choice fruits, \\
\poeml both old and new, \\
\poemll    that I've stored up for you, my beloved.
\labelchapt{8}
\poeml \chapt{8}
\v{1}If only you were like a brother to me, \\
\poeml like\fnote{\fbackref{8:1} The Heb. lacks \fbib{like} } one who nursed at my mother's breasts. \\
\poeml If I found you outside I would kiss you, \\
\poeml and no one would view me with contempt.\fnote{\fbackref{8:1} I.e. think badly of me} \\
\poeml \v{2}I would lead you, \\
\poemll    I would bring you \\
\poeml to the house of my mother \\
\poemll    who used to teach me. \\
\poeml I would give you some spiced wine to drink, \\
\poemll    from the juice of my pomegranates.\fnote{\fbackref{8:2} Or \fbib{some of my sweet pomegranate wine}} \\
\poeml \v{3}Let his left hand be under my head, \\
\poemll    and let his right hand embrace me. \\
\poeml \v{4}Swear to me, young women of Jerusalem, \\
\poemll    that you won't awaken or arouse love \\
\poemlll       before its proper time!\fnote{\fbackref{8:3} Lit. \fbib{until it pleases}}
\passage{The Young Women}
\poeml \v{5}Who is this coming up from the desert, \\
\poemll    leaning on her beloved?
\passage{The Loved One}
\poeml Under the apple tree\fnote{\fbackref{8:5} Or \fbib{apricot tree}} I awakened you. \\
\poemll    There your mother had gone into labor with you; \\
\poemlll       there she went into labor and gave birth to you. \\
\poeml \v{6}Set me like a seal\fnote{\fbackref{8:6} I.e. a cylinder seal or a signet ring used to indicate ownership or authority} over your heart, \\
\poemll    like a seal on your arm. \\
\poeml For love is as strong as death, \\
\poemll    passion\fnote{\fbackref{8:6} Or \fbib{jealousy}} as intense as Sheol.\fnote{\fbackref{8:6} I.e. the realm of the dead} \\
\poeml The flames of love\fnote{\fbackref{8:6} Lit. \fbib{its flames}} are flames of fire, \\
\poemll    a blaze that comes from the \divine{Lord}.\fnote{\fbackref{8:6} Or \fbib{an intense blaze}} \\
\poeml \v{7}Mighty bodies of water cannot extinguish love, \\
\poemll    rivers cannot put it out. \\
\poeml If a man were to give all the wealth of his house for love, \\
\poemll    he would surely be viewed with contempt.
\passage{The Loved One's Brothers}
\poeml \v{8}We\fnote{\fbackref{8:8} I.e. the bride's brothers} have a little sister, \\
\poemll    and she has not yet reached maturity.\fnote{\fbackref{8:8} Lit. \fbib{has no breasts}} \\
\poeml What will we do for our sister \\
\poemll    to prepare her for\fnote{\fbackref{8:8} Lit. \fbib{sister on the day when she's spoken for}} her engagement?\fnote{\fbackref{8:8} Or \fbib{marriage}} \\
\poeml \v{9}If she's a wall, \\
\poemll    on her we will build a battlement of silver. \\
\poeml If she's a door, \\
\poemll    we will enclose her with planks of cedar.
\passage{The Loved One}
\poeml \v{10}I was a wall and my breasts were like towers. \\
\poemll    Then I became like one who finds peace\fnote{\fbackref{8:10} Or \fbib{contentment}} in his eyes. \\
\poeml \v{11}Solomon has a vineyard in Baal-hamon. \\
\poemll    He entrusted the vineyard to caretakers--- \\
\poeml each one is to bring 1,000 pieces of silver \\
\poemll    in exchange for its fruit. \\
\poeml \v{12}My vineyard belongs to me and is at my disposal.\fnote{\fbackref{8:12} Lit. \fbib{is before me}} \\
\poemll    The 1,000 are for you, Solomon, \\
\poemlll       and 200 are for those who take care of its fruit.
\passage{The Lover}
\poeml \v{13}You who sit in the gardens, \\
\poemll    companions are listening for your voice, \\
\poemlll       but let me hear it.
\passage{The Loved One}
\poeml \v{14}Come quickly, my beloved, and be like a gazelle \\
\poemll    or a young stag on the mountains of spices.\end{poetry}
