\bookheader{Hosea}
\labelbook{Hos}

\bookpretitle{The Book of the Prophet}
\booktitle{Hosea}

\labelchapt{1}
\passage{The Word of the \divine{Lord} to Hosea}

\chapt{1}
\v{1}A message from the \divine{Lord} came\fnote{\fbackref{1:1} Lit. \fbib{that came}} to Beeri's son Hosea\fnote{\fbackref{1:1} The Heb. name \fbib{Hosea} means \fbib{salvation}} during the reigns of\fnote{\fbackref{1:1} Lit. \fbib{in the days of}} Uzziah, Jotham, Ahaz, and Hezekiah, kings of Judah, and during the reign of\fnote{\fbackref{1:1} Lit. \fbib{in the days of}} Joash's son Jeroboam, who was king of Israel.
\passage{Hosea's Wife and Family}

\v{2}When a message from the \divine{Lord} came to Hosea, the \divine{Lord} told him,\fnote{\fbackref{1:2} Lit. \fbib{Hosea}} ``Go marry a prostitute and have children with her,\fnote{\fbackref{1:2} Lit. \fbib{children of prostitution}} because the land is prostituting itself by departing from the \divine{Lord}.'' \v{3}So he went out and married Diblaim's daughter Gomer. She conceived with him and gave birth to a son.
\passage{Naming the Children}

\v{4}The \divine{Lord} told Hosea,\fnote{\fbackref{1:4} Lit. \fbib{him}} ``Name the child\fnote{\fbackref{1:4} Lit. \fbib{Name him}} `Jezreel,'\fnote{\fbackref{1:4} The Heb. name \fbib{Jezreel} means ``God sows''} because in a little while I'll avenge the blood that was shed by Jehu's dynasty at Jezreel. I'll put an end to the kingdom of the house of Israel. \v{5}At that time I'll shatter the military strength\fnote{\fbackref{1:5} Lit. \fbib{the bow}} of Israel in the Valley of Jezreel.''

\v{6}Gomer\fnote{\fbackref{1:6} Lit. \fbib{She}} conceived again and gave birth to a daughter, so the \divine{Lord}\fnote{\fbackref{1:6} Lit. \fbib{he}} told Hosea,\fnote{\fbackref{1:6} Lit. \fbib{him}} ``Name her `Lo-ruhamah,'\fnote{\fbackref{1:6} The Heb. name \fbib{Lo-ruhamah} means \fbib{No mercy}} because I will no longer be showing mercy to the house of Israel, nor will I forgive them. \v{7}But I'll have mercy on the house of Judah, and I'll save them by the \divine{Lord} their God---I will not save them by the bow, by the sword, by battle, by horses, or by cavalry.''

\v{8}After Gomer\fnote{\fbackref{1:8} Lit. \fbib{she}} had weaned Lo-ruhamah, she conceived again and gave birth to a son, \v{9}so the \divine{Lord}\fnote{\fbackref{1:9} Lit. \fbib{he}} told Hosea,\fnote{\fbackref{1:9} Lit. \fbib{him}} ``Name him `Lo-ammi,'\fnote{\fbackref{1:9} The Heb. name \fbib{Lo-ammi} means \fbib{Not my people}} because you are not my people, and I will not be your God.\fnote{\fbackref{1:9} Lit. \fbib{be for you}} \v{10}\fnote{\fbackref{1:10} This vs. is 2:1 in MT}Despite this, the number of the people of Israel will be like ocean sand, which can neither be measured nor counted. And the time will come when instead of it being said,\fnote{\fbackref{1:10} Lit. \fbib{told them}} `You are not my people,' it will be said,\fnote{\fbackref{1:10} Lit. \fbib{told them}} `You are children of the living God.' \v{11}\fnote{\fbackref{1:11} This vs. is 2:2 in MT}And the people of Judah and the people of Israel will be united as one. They will appoint for themselves a single leader and will take dominion over\fnote{\fbackref{1:11} Lit. \fbib{will go up from}} the land, for great will be the day of Jezreel.\chapt{2}
\v{1}\fnote{\fbackref{2:1} This vs. is 2:3 in MT, and so throughout the chapter.}So call your brothers `Ammi,'\fnote{\fbackref{2:1} Lit. \fbib{My People}} and your sisters `Ruhamah.'\,''\fnote{\fbackref{2:1} Lit. \fbib{Beloved}; i.e. in contrast to the names \fbib{Lo-Ruhamah} and \fbib{Lo-Ammi} in vv. 2:6-9}
\labelchapt{2}
\passage{Gomer is Rebuked}

\begin{poetry}
\poeml \v{2}``Call your mother to account, call her--- \\
\poemll    for she is not my wife, \\
\poemlll       and I'm not her husband. \\
\poeml Let her do away with her seductive looks \\
\poemll    and remove her adultery from between her breasts. \\
\poeml \v{3}Otherwise, I'll strip her naked--- \\
\poemll    as she was on the day she was born--- \\
\poeml make her like a wilderness, \\
\poemll    turn her into a parched land, \\
\poemlll       and cause her to die of thirst. \\
\poeml \v{4}Furthermore, I'll not show pity on her children, \\
\poemll    since they are children of prostitution. \\
\poeml \v{5}Indeed, their mother has committed prostitution--- \\
\poemll    the one who has been conceiving them has acted disgracefully--- \\
\poeml when she said, \\
\poemll    `I'm going after my lovers, \\
\poemlll       who provide me food and water, \\
\poemll    as well as my wool, my flax, \\
\poemlll       my oil, and my wine.' \\
\poeml \v{6}``Look how I'm blocking her\fnote{\fbackref{2:6} So with LXX. MT reads \fbib{your}} path with thorns \\
\poemll    and building a wall to hinder\fnote{\fbackref{2:6} Lit. \fbib{wall against}} her, \\
\poemlll       so she can't find her way. \\
\poeml \v{7}She will pursue her lovers, \\
\poemll    but she won't catch up with them. \\
\poeml She will seek them, \\
\poemll    but she won't find them. \\
\poeml Then she will say, \\
\poemll    `I'll go back and return to my first husband, \\
\poemlll       because it was better for me then than now.' \\
\poeml \v{8}She didn't recognize \\
\poemll    that it was I who provided her grain, wine, and oil, \\
\poeml and it was I who gave her silver, \\
\poemll    while they crafted gold for Baal. \\
\poeml \v{9}``Therefore I'll return \\
\poemll    and take back my grain at harvest time \\
\poemlll       and my new wine in its season. \\
\poeml I'll take back my wool and my flax \\
\poemll    that was to have covered her nakedness. \\
\poeml \v{10}So now I'll reveal her lewdness to the eyes of her lovers, \\
\poemll    and no man will rescue her from my control.\fnote{\fbackref{2:10} Lit. \fbib{hand}} \\
\poeml \v{11}I'll put a stop to her mirth, \\
\poemll    along with her celebrations, her New Moons, her Sabbaths,\fnote{\fbackref{2:11} So with LXX. MT reads \fbib{Sabbath}} and all of her festive assemblies. \\
\poeml \v{12}I'll destroy her vines and her fig trees, \\
\poemll    about which she said, \\
\poeml `These are the earnings that my lovers paid me. \\
\poemll    I'll make them grow into a forest, \\
\poemlll       and the wild animals will eat from them.' \\
\poeml \v{13}I'll punish her for the time she has devoted to the Baals,\fnote{\fbackref{2:13} Lit. \fbib{for the days of the Baalim}; i.e. Canaanite deities} \\
\poemll    to whom she burned incense, \\
\poeml and for whom she put on her earrings and jewels \\
\poemll    so she could go after her lovers and forget me,'' \\
\poemlll       declares the \divine{Lord}.
\passage{Alluring a Wayward Wife}
\poeml \v{14}``Therefore, look! I will now allure her. \\
\poemll    I will make her go out to the wilderness, \\
\poemlll       and will speak to her heart. \\
\poeml \v{15}There I will restore her vineyards to her, \\
\poemll    and the Valley of Achor will become a doorway to hope. \\
\poeml There she will respond as she did in her youth, \\
\poemll    when she came up from Egypt.''
\passage{The Restoration of Israel}
\poeml \v{16}``It will come about at that time,'' \\
\poemll    declares the \divine{Lord}, \\
\poeml ``that you will address me as `My husband,' \\
\poemll    and you will no longer call me `My master'.\fnote{\fbackref{2:16} Heb. \fbib{baali}, a word play alluding to the Canaanite deity of a similar name} \\
\poeml \v{17}I will remove the names of the Baals\fnote{\fbackref{2:17} I.e. Canaanite deities} from her vocabulary\fnote{\fbackref{2:17} Lit. \fbib{mouth}}--- \\
\poemll    they will not be remembered by their names anymore. \\
\poeml \v{18}I will make a covenant with them at that time, \\
\poemll    a covenant\fnote{\fbackref{2:18} The Heb. lacks \fbib{a covenant}} with the wild animals of the field, \\
\poeml with the birds of the air, \\
\poemll    and with the creatures of the ground. \\
\poeml I will banish\fnote{\fbackref{2:18} Lit. \fbib{break}} the battle bow, the sword, and war from the earth. \\
\poemll    I will cause my people\fnote{\fbackref{2:18} Lit. \fbib{cause them}} to lie down where it is safe. \\
\poeml \v{19}I will make you my wife forever--- \\
\poeml I will make you my wife in a way that is righteous, \\
\poemll    in a manner that is just, \\
\poeml by a love that is gracious, \\
\poemll    and by a motive that is mercy. \\
\poeml \v{20}I will make you my wife because of my\fnote{\fbackref{2:20} Lit. \fbib{wife in}} faithfulness, \\
\poemll    and you will know the \divine{Lord}. \\
\poeml \v{21}``It will come about at that time that I will respond,'' \\
\poemll    declares the \divine{Lord}, \\
\poeml ``I will respond to the heavens, \\
\poemll    and they will respond to the earth. \\
\poeml \v{22}The earth will respond with grain, new wine, and oil, \\
\poemll    and they will respond to Jezreel.\fnote{\fbackref{2:22} The Heb. name \fbib{Jezreel} means \fbib{God sows}} \\
\poeml \v{23}I will plant my people\fnote{\fbackref{2:23} Lit. \fbib{planted her}} in the land for myself. \\
\poemll    I will show mercy on her who has received no mercy\fnote{\fbackref{2:23} Heb. \fbib{on Lo-Ruhamah}} \\
\poeml I will say to those who are not my people,\fnote{\fbackref{2:23} Heb. \fbib{to Lo-Ammi}} `You are my people!' \\
\poemll    and they will say, `You are\fnote{\fbackref{2:23} The Heb. lacks \fbib{You are}} my God.'\,''
\end{poetry}
\labelchapt{3}
\passage{Hosea Reconciles with His Wife}

\chapt{3}
\v{1}Then the \divine{Lord} told me: ``Go love your wife\fnote{\fbackref{3:1} Or \fbib{love a woman}} again, even though she is being loved by another and is committing adultery. Love her the same way\fnote{\fbackref{3:1} Lit. \fbib{adultery, as}} the \divine{Lord} loves the people of Israel, even though they look to other gods and love raisin cakes.''\fnote{\fbackref{3:1} I.e. cakes used for offerings; cf. 2Sam 6:19} \v{2}So I bought her back for myself for fifteen pieces of silver and one and a half omers\fnote{\fbackref{3:2} I.e. about ten bushels} of barley.

\v{3}I told her, ``You will remain with me a long time,\fnote{\fbackref{3:3} Lit. \fbib{me many days}} you won't be promiscuous, you won't be involved with any man, and I'll do the same.''\fnote{\fbackref{3:3} Lit. \fbib{same with you}}

\v{4}Likewise, the people of Israel will dwell a long time\fnote{\fbackref{3:4} Lit. \fbib{will live many days}} without a king, without a prince, without sacrifice, without sacred\fnote{\fbackref{3:4} The Heb. lacks \fbib{sacred}} pillars, and with neither ephod nor teraphim.\fnote{\fbackref{3:4} I.e. images used for divination} \v{5}Afterward, the people of Israel will return and seek the \divine{Lord} their God and David their king. They will come in awe to the \divine{Lord} and to his goodness in the last days.
\labelchapt{4}
\passage{God Accuses Israel}

\begin{poetry}
\poeml \chapt{4}
\v{1}``Hear this message from the \divine{Lord}, people of Israel. \\
\poemll    Indeed, the \divine{Lord} brings a charge \\
\poemll    against the people who live in the land--- \\
\poeml for there is no truth and no gracious love \\
\poemll    or knowledge of God in the land. \\
\poeml \v{2}Swearing, lying, murder, theft, and adultery are rampant, \\
\poemll    and blood mingles with blood. \\
\poeml \v{3}Therefore the land will mourn, \\
\poemll    and all who live there will languish, \\
\poeml along with the wild animals of the field and the birds of the air. \\
\poemll    Even the fish in the sea will disappear. \\
\poeml \v{4}``Let no one fight or bring charges against another, \\
\poemll    for my dispute is with you, priest. \\
\poeml \v{5}So you will stumble during the day, \\
\poemll    the prophet also will stumble with you at night, \\
\poemlll       and I will destroy your mother.\fnote{\fbackref{4:5} I.e. \fbib{Israel}} \\
\poeml \v{6}My people are destroyed because they lack knowledge of me.\fnote{\fbackref{4:6} The Heb. lacks \fbib{of me}} \\
\poemll    Because you rejected that knowledge, \\
\poemlll       I will reject you as a priest for me. \\
\poeml Since you forget the Law of your God, \\
\poemll    I will also forget your children. \\
\poeml \v{7}``The more they increased in number,\fnote{\fbackref{4:7} The Heb. lacks \fbib{in number}} \\
\poemll    the more they sinned against me, \\
\poemlll       so I will change their glory into shame. \\
\poeml \v{8}They feed on the sin of my people; \\
\poemll    they purpose in their heart to transgress. \\
\poeml \v{9}So it will be: like people, like priest. \\
\poemll    I will punish them for their lifestyles, \\
\poemlll       rewarding them according to their behavior. \\
\poeml \v{10}They will eat, \\
\poemll    but will not be satisfied. \\
\poeml They will engage in prostitution, \\
\poemll    but they won't increase, \\
\poemlll       because they have stopped listening to the \divine{Lord}. \\
\poeml \v{11}``Sexual immorality, wine, and fresh wine seduce the heart of my people.\fnote{\fbackref{4:11} So LXX. The Heb. lacks \fbib{of my people}} \\
\poeml \v{12}My people seek counsel from their piece of wood, \\
\poemll    and their diviner's rod\fnote{\fbackref{4:12} Lit. \fbib{their carved wood}} speaks to them. \\
\poeml For a spirit of prostitution causes them to go astray; \\
\poemll    in their immorality they desert their God. \\
\poeml \v{13}They offer sacrifices on the mountain tops, \\
\poemll    burning offerings on the hills, \\
\poeml under oaks, poplars, and terebinth\fnote{\fbackref{4:13} Or \fbib{great}} trees, \\
\poemll    since their shade is very good. \\
\poeml Therefore your daughters are prostitutes \\
\poemll    and your daughters-in-law commit adultery. \\
\poeml \v{14}However, I'm not going to punish your daughters \\
\poemll    when they commit prostitution, \\
\poeml nor your daughters-in-law \\
\poemll    when they commit adultery, \\
\poeml because their men are themselves immoral--- \\
\poemll    they offer sacrifices with prostitutes. \\
\poeml These people who aren't discerning will stumble. \\
\poeml \v{15}``Even though you prostitute yourself, Israel--- \\
\poemll    let not Judah incur guilt--- \\
\poeml don't go to Gilgal, \\
\poemll    or visit Beth-aven, \\
\poemlll       or swear an oath using the \divine{Lord}'s name.\fnote{\fbackref{4:15} Lit. \fbib{oath, `{\ldots}as the \divine{Lord} lives'}} \\
\poeml \v{16}For Israel is as obstinate as a stubborn mule!\fnote{\fbackref{4:16} Or \fbib{cow}} \\
\poemll    Nevertheless, will not the \divine{Lord} feed them like a lamb in a broad pasture? \\
\poeml \v{17}Ephraim has become entwined with idols; \\
\poemll    leave him alone! \\
\poeml \v{18}While drinking to excess, they prostitute themselves. \\
\poemll    They're in love with dishonor. \\
\poeml \v{19}A wind storm will carry them away in its embrace, \\
\poemll    and their sacrifices will bring them shame.''
\end{poetry}
\labelchapt{5}
\passage{Judgment against Israel}

\begin{poetry}
\poeml \chapt{5}
\v{1}``Hear this, priests, \\
\poemll    pay attention, house of Israel, \\
\poemlll       listen, royal family! \\
\poeml For judgment is coming your way,\fnote{\fbackref{5:1} Lit. \fbib{is toward you}} \\
\poemll    because you have been a trap to Mizpah, \\
\poemlll       a snare spread out on Mount\fnote{\fbackref{5:1} The Heb. lacks \fbib{Mount}} Tabor. \\
\poeml \v{2}The rebels are deep into their slaughter; \\
\poemll    I am punishing them all. \\
\poeml \v{3}I know Ephraim, \\
\poemll    and Israel cannot hide from me, \\
\poeml since you, Ephraim, have been acting like a prostitute, \\
\poemll    defiling Israel. \\
\poeml \v{4}``Their actions hinder them from turning to their God, \\
\poemll    because a spirit of fornication is in their midst, \\
\poemlll       and the \divine{Lord} they do not know. \\
\poeml \v{5}The arrogance of Israel testifies against him; \\
\poemll    therefore Israel and Ephraim will stumble in their iniquity, \\
\poemlll       and Judah with them. \\
\poeml \v{6}They will go with their flocks and herds \\
\poemll    to seek the \divine{Lord}, \\
\poeml but they will not find him; \\
\poemll    he has withdrawn from them. \\
\poeml \v{7}They have been unfaithful to the \divine{Lord}, \\
\poemll    having raised unbelieving children. \\
\poeml In the coming month they will be devoured, \\
\poemll    along with their fields.\fnote{\fbackref{5:7} Or \fbib{inheritance}} \\
\poeml \v{8}``Sound the trumpet in Gibeah, \\
\poemll    and the alarm in Ramah. \\
\poeml Cry out at Beth-aven \\
\poemll    Go out,\fnote{\fbackref{5:8} So LXX.; MT \fbib{After you}} Benjamin! \\
\poeml \v{9}Ephraim will be desolate \\
\poemll    when it is rebuked. \\
\poeml I have made known among the tribes of Israel \\
\poemll    what will surely come about. \\
\poeml \v{10}The princes of Judah have become \\
\poemll    like those who move boundary markers: \\
\poemlll       I will pour out my anger on them like water. \\
\poeml \v{11}Ephraim is crushed, \\
\poemll    broken by judgment, \\
\poemlll       because he\fnote{\fbackref{5:11} I.e. Ephraim as an individual personifying the northern kingdom of Israel.} willingly pursued idols.\fnote{\fbackref{5:11} So LXX. MT \fbib{willingly went away from the command}} \\
\poeml \v{12}Therefore I will consume\fnote{\fbackref{5:12} The Heb. lacks \fbib{will consume}} Ephraim like a moth, \\
\poemll    and the house of Judah as rottenness consumes. \\
\poeml \v{13}When Ephraim examined his illness \\
\poemll    and Judah his injury, \\
\poeml then Ephraim went to Assyria, \\
\poemll    and inquired of the great king; \\
\poeml but he could not cure you \\
\poemll    nor heal your injury. \\
\poeml \v{14}Therefore I will be like a lion to Ephraim, \\
\poemll    and like a young lion to the house of Judah. \\
\poeml I---even I---will tear them\fnote{\fbackref{5:14} The Heb. lacks \fbib{them}} to pieces, \\
\poemll    and then I will leave. \\
\poeml I will take them\fnote{\fbackref{5:14} The Heb. lacks \fbib{them}} away, \\
\poemll    and there will be no rescue. \\
\poeml \v{15}``I will leave and go back to my place \\
\poemll    until they admit their offense \\
\poemlll       and seek my face. \\
\poeml When affliction comes to them, \\
\poemll    they will eagerly seek me.''
\end{poetry}
\labelchapt{6}
\passage{A Call for Israel to Repent}

\begin{poetry}
\poeml \chapt{6}
\v{1}``Come, let us return to the \divine{Lord}; \\
\poemll    even though he has torn us,\fnote{\fbackref{6:1} The Heb. lacks \fbib{us}} \\
\poemlll       he will heal us. \\
\poeml Even though he has wounded us, \\
\poemll    he will bind our wounds.\fnote{\fbackref{6:1} The Heb. lacks \fbib{our wounds}} \\
\poeml \v{2}After two days, he will restore us to life, \\
\poemll    on the third day he will raise us up, \\
\poemlll       and we will live in his presence. \\
\poeml \v{3}Let us know, \\
\poemll    let us pursue knowledge of the \divine{Lord}; \\
\poemlll       his coming is as certain as the dawn. \\
\poeml He will come to us like the rain, \\
\poemll    like the autumn and spring rains come on the earth. \\
\poeml \v{4}``What am I to do with you, Ephraim? \\
\poemll    What am I to do with you, Judah? \\
\poeml Your love is like a morning rain cloud--- \\
\poemll    it passes away like the morning dew. \\
\poeml \v{5}Therefore I cut them\fnote{\fbackref{6:5} The Heb. lacks \fbib{them}} to pieces by the prophets, \\
\poemll    killing them by the words from my mouth. \\
\poemlll       The verdict against you shines like a beacon. \\
\poeml \v{6}For it is love that I seek, \\
\poemll    and not sacrifice; \\
\poemlll       knowledge of God more than burnt offerings. \\
\poeml \v{7}``But like Adam,\fnote{\fbackref{6:7} Or \fbib{men}} they broke the covenant; \\
\poemll    in this they have acted deceitfully against me. \\
\poeml \v{8}Gilead is a lawless town; \\
\poemll    it is polluted by bloodshed. \\
\poeml \v{9}Like a gang of thieves that stalk a man, \\
\poemll    priests commit murder along the road to Shechem, \\
\poemlll       committing shameful crimes. \\
\poeml \v{10}I have seen a horrible evil in the house of Israel--- \\
\poemll    Ephraim's promiscuity. \\
\poemlll       Israel is defiled. \\
\poeml \v{11}``So, Judah, a harvest has been appointed for you \\
\poemll    when I restore my people from captivity.''
\end{poetry}
\labelchapt{7}
\passage{God Accuses Israel}

\begin{poetry}
\poeml \chapt{7}
\v{1}``When I was healing Israel, \\
\poeml Ephraim's sin was uncovered, \\
\poemlll       along with Samaria's wickedness. \\
\poeml While they craft lying schemes, \\
\poemll    the thief invades, \\
\poemlll       and the gang of thieves plunders outside. \\
\poeml \v{2}It never occurs to them that I remember all their sin. \\
\poemll    Now their actions have caught up with them, \\
\poemlll       and they have my attention.\fnote{\fbackref{7:2} Lit. \fbib{they are before my face}} \\
\poeml \v{3}They please the king with their evil, \\
\poemll    and the princes with their dishonesty. \\
\poeml \v{4}All of them are adulterers--- \\
\poemll    they burn like an oven prepared by the baker, \\
\poeml who has ceased stoking it \\
\poemll    until the dough is leavened. \\
\poeml \v{5}``On the king's festival day \\
\poemll    the princes got drunk from wine, \\
\poemlll       so the king\fnote{\fbackref{7:5} Lit. \fbib{so he}} joined the mockers. \\
\poeml \v{6}For they have stirred up themselves\fnote{\fbackref{7:6} Lit. \fbib{up their heart}} like an oven \\
\poemll    as they lie in ambush. \\
\poeml Their baker sleeps through the night; \\
\poemll    in the morning, the oven\fnote{\fbackref{7:6} Lit. \fbib{morning, it}} will be blazing like a fire. \\
\poeml \v{7}They all burn like an oven; \\
\poemll    they have consumed their judges; \\
\poeml all their kings have fallen--- \\
\poemll    not even one of them calls on me. \\
\poeml \v{8}``Ephraim compromises with\fnote{\fbackref{7:8} Or \fbib{dilutes himself among}} the nations; \\
\poemll    he's a half-baked cake.\fnote{\fbackref{7:8} Lit. \fbib{a cake not turned}} \\
\poeml \v{9}Foreigners have consumed his strength, \\
\poemll    and he hasn't noticed. \\
\poeml Furthermore, his head is sprinkled with gray hair, \\
\poemll    but he doesn't realize it. \\
\poeml \v{10}Israel's arrogance testifies against him;\fnote{\fbackref{7:10} Lit. \fbib{testifies in his face}} \\
\poemll    but they do not return to the \divine{Lord} their God, \\
\poemlll       nor seek him in all of this. \\
\poeml \v{11}``Ephraim is also like a silly dove, \\
\poemll    lacking sense:\fnote{\fbackref{7:11} Lit. \fbib{heart}} \\
\poeml They call out to Egypt, \\
\poemll    and turn toward Assyria. \\
\poeml \v{12}When they go, \\
\poemll    I'll cast my net over them. \\
\poeml I'll bring them down, as one shoots\fnote{\fbackref{7:12} The Heb. lacks \fbib{one shoots}} birds in the sky. \\
\poemll    I'll chasten them, \\
\poemlll       as the assembly has already heard. \\
\poeml \v{13}Woe to them--- \\
\poemll    because they have run away from me. \\
\poeml Ruin to them--- \\
\poemll    because they have sinned against me. \\
\poeml Even though I redeemed them, \\
\poemll    they spread lies against me. \\
\poeml \v{14}They will not cry to me from their heart--- \\
\poemll    instead, they wail on their beds. \\
\poeml They gather together to eat and drink,\fnote{\fbackref{7:14} Lit. \fbib{together for grain and fresh wine}} \\
\poemll    turning away from me. \\
\poeml \v{15}``Though I have taught them \\
\poemll    and strengthened their arms, \\
\poemlll       nevertheless they plot evil against me. \\
\poeml \v{16}They return---but not to the Most High. \\
\poemll    They are like a defective weapon.\fnote{\fbackref{7:16} Lit. \fbib{bow}} \\
\poeml Their princes will fall by the sword \\
\poemll    because of their raging tongue, \\
\poemlll       and they will be a laughingstock in the land of Egypt.''
\end{poetry}
\labelchapt{8}
\passage{Reaping the Wind Storm}

\begin{poetry}
\poeml \chapt{8}
\v{1}``Sound the ram's horn! \\
\poeml Like a vulture\fnote{\fbackref{8:1} Or \fbib{eagle}} the enemy\fnote{\fbackref{8:1} Lit. \fbib{he}} will come against the Temple of the \divine{Lord}, \\
\poeml because Israel\fnote{\fbackref{8:1} Lit. \fbib{they}} violated my covenant, \\
\poemll    transgressing my Law. \\
\poeml \v{2}They cry out to me, \\
\poemll    `God, we of Israel acknowledge you.' \\
\poeml \v{3}``Israel has discarded what is good. \\
\poemll    The enemy will pursue them.\fnote{\fbackref{8:3} Lit. \fbib{him}} \\
\poeml \v{4}They set kings in place, \\
\poemll    but not by me. \\
\poeml They established princes, \\
\poemll    whom I did not recognize. \\
\poeml They crafted idols for themselves from their silver and gold; \\
\poemll    as a result, they will be destroyed. \\
\poeml \v{5}Your calf,\fnote{\fbackref{8:5} I.e. the idol they crafted (cf. v. 4)} Samaria, has been thrown away. \\
\poemll    My anger is burning against them. \\
\poemlll       How long until they become pure again? \\
\poeml \v{6}Because from Israel it was fashioned by craftsmen, \\
\poemll    it is not God; \\
\poemlll       therefore Samaria's calf will be broken in pieces. \\
\poeml \v{7}``Because they sow the wind, \\
\poemll    they will reap the wind storm. \\
\poeml The plant has no stalk \\
\poemll    and its bud yields no grain. \\
\poeml Even if there's a harvest, \\
\poemll    foreigners will gobble it up. \\
\poeml \v{8}Israel has been devoured; \\
\poemll    now they will live among the nations \\
\poemlll       like a worthless container. \\
\poeml \v{9}``Because they went over to Assyria, \\
\poemll    they are like a wild donkey alone by itself. \\
\poemlll       Ephraim has hired some lovers. \\
\poeml \v{10}Even though they sold themselves to the nations, \\
\poemll    I will gather them. \\
\poeml They will mourn for a while \\
\poemll    for the burden they were to the king and princes.\fnote{\fbackref{8:10} So LXX. MT \fbib{king of princes}} \\
\poeml \v{11}``The more altars Ephraim builds for sin, \\
\poemll    the more altars there will be for sin. \\
\poeml \v{12}I prescribed great things from my Law for them,\fnote{\fbackref{8:12} Lit. \fbib{him}} \\
\poemll    but they considered them profane. \\
\poeml \v{13}They offer me meat from the sacrifices of my offerings, \\
\poemll    and they eat from it, \\
\poemlll       but the \divine{Lord} does not accept them. \\
\poeml He will now remember their transgression \\
\poemll    and pay them back for their sins; \\
\poemlll       to Egypt they will return. \\
\poeml \v{14}``Israel has neglected its maker in building palaces. \\
\poemll    Judah has multiplied its fortified cities, \\
\poeml but I will send fire to their cities, \\
\poemll    and it will consume their fortresses.''
\end{poetry}
\labelchapt{9}
\passage{Punishment for Israel}

\begin{poetry}
\poeml \chapt{9}
\v{1}``Don't celebrate, Israel, like other nations would rejoice, \\
\poeml because you left your God by committing fornication, \\
\poemlll       loving the profit you gained on all of the threshing floors. \\
\poeml \v{2}Neither threshing floor nor winepress will sustain them, \\
\poemll    and the new wine will disappoint her. \\
\poeml \v{3}They will not live in the \divine{Lord}'s land--- \\
\poemll    Ephraim will return to Egypt, \\
\poemlll       and they will eat unclean food in Assyria. \\
\poeml \v{4}They won't present wine offerings to the \divine{Lord}, \\
\poemll    nor will they please him. \\
\poeml Their sacrifices will seem like food for mourners--- \\
\poemll    everyone who eats them will become unclean; \\
\poemlll       none of them will enter the Temple of the \divine{Lord}. \\
\poeml \v{5}``What will you do on the designated holiday, \\
\poemll    when the \divine{Lord}'s festival comes? \\
\poeml \v{6}Look! They have gone away because of the destruction--- \\
\poemll    Egypt will gather them up, \\
\poemlll       and Memphis\fnote{\fbackref{9:6} So LXX. Heb. \fbib{Moph}; i.e. the capital of Lower Egypt} will bury them. \\
\poeml Weeds will overgrow their inheritance,\fnote{\fbackref{9:6} Lit. \fbib{their precious silver}} \\
\poemll    and thorns will grow\fnote{\fbackref{9:6} Lit. \fbib{be}} in their tents. \\
\poeml \v{7}The time for your judgment has now come; \\
\poemll    payday is here--- \\
\poemlll       and Israel knows it. \\
\poeml The prophet is a fool, \\
\poemll    and the spiritual man is insane. \\
\poeml Because of your great sin, \\
\poemll    the hatred against you\fnote{\fbackref{9:7} The Heb. lacks \fbib{against you}} is great. \\
\poeml \v{8}While Ephraim stands watch with my God, \\
\poemll    the prophet has snares set that will trap his ways, \\
\poemlll       and hostility lodges in the Temple of his God. \\
\poeml \v{9}They have corrupted themselves deeply, \\
\poemll    as did Gibeah\fnote{\fbackref{9:9} Cf. Judg 19:1} in its day. \\
\poeml Therefore God\fnote{\fbackref{9:9} Lit. \fbib{he}} will remember their lawlessness, \\
\poemll    and he will pay them back for their sins. \\
\poeml \v{10}``I found Israel, \\
\poemll    as one finds\fnote{\fbackref{9:10} The Heb. lacks \fbib{finds}} grapes in the wilderness; \\
\poeml Your ancestors seemed to me like the fruit \\
\poemll    gleaned from a fig tree's first harvest. \\
\poeml When they went to Baal-peor,\fnote{\fbackref{9:10} Cf. Num 25:1-3} \\
\poemll    they devoted themselves to that filth, \\
\poeml and they became loathsome, \\
\poemll    like what they loved. \\
\poeml \v{11}The glory of Ephraim will fly away like a bird--- \\
\poemll    no birth, no pregnancy, not even a conception. \\
\poeml \v{12}Even if they rear their children, \\
\poemll    I will, in turn, make them childless--- \\
\poeml in fact, woe to them \\
\poemll    when I turn away from them! \\
\poeml \v{13}Ephraim, as I see it, is like Tyre, \\
\poemll    planted in a comfortable place; \\
\poeml Ephraim will bear children \\
\poemll    but they will be executed.'' \\
\poeml \v{14}Give them, \divine{Lord}--- \\
\poemll    What will you give? \\
\poemlll       You will give them a womb that miscarries and dry breasts. \\
\poeml \v{15}``All of their wickedness started\fnote{\fbackref{9:15} The Heb. lacks \fbib{started}} in Gilgal, \\
\poemll    because I began to hate them there. \\
\poeml Because of the wickedness of their behavior, \\
\poemll    I will drive them from my Temple. \\
\poeml I will not love them anymore; \\
\poemll    all their leaders are rebels. \\
\poeml \v{16}Ephraim is blighted;\fnote{\fbackref{9:16} Or \fbib{stricken}} \\
\poemll    its roots shriveled. \\
\poemlll       It can bear no fruit. \\
\poeml Even if they bear children, \\
\poemll    I will kill their cherished offspring. \\
\poeml \v{17}``My God will reject them, \\
\poemll    because they did not obey him, \\
\poemlll       and they will become wanderers among the nations.''
\end{poetry}
\labelchapt{10}
\passage{The Coming Destruction}

\begin{poetry}
\poeml \chapt{10}
\v{1}``Israel, the overgrown\fnote{\fbackref{10:1} Or \fbib{empty}} vine, bears fruit like itself; \\
\poeml the more fruitful they become, \\
\poemlll       the more altars they build. \\
\poeml The better the land, \\
\poemll    the more ornate the stone idols.\fnote{\fbackref{10:1} Or \fbib{pillars}} \\
\poeml \v{2}Their hearts are divided; \\
\poemll    from now on they are to be found guilty. \\
\poeml God\fnote{\fbackref{10:2} Lit. \fbib{He}} will tear down their altars, \\
\poemll    he will destroy their stone idols.\fnote{\fbackref{10:2} Or \fbib{pillars}} \\
\poeml \v{3}From now on they will say, \\
\poemll    `We have no king, \\
\poemlll       because we did not fear the \divine{Lord}--- \\
\poemll    and what would a king do for us?' \\
\poeml \v{4}Their word is falsely given \\
\poemll    as they make their agreements;\fnote{\fbackref{10:4} Lit. \fbib{make a covenant}} \\
\poeml so judgment springs up \\
\poemll    like poisonous weeds in the furrows of a field.\fnote{\fbackref{10:4} Cf. Deut 29:18} \\
\poeml \v{5}``The residents of Samaria will be terrified \\
\poemll    because of the cows\fnote{\fbackref{10:5} I.e. Canaanite heifer deities} of Beth-aven. \\
\poeml Its people will mourn over Beth-aven,\fnote{\fbackref{10:5} Lit. \fbib{it}} \\
\poemll    along with the priests who will mourn its glory, \\
\poemlll       because that glory has departed.\fnote{\fbackref{10:5} The Heb. verb \fbib{depart} is similar to the Heb. verb \fbib{mourn}} \\
\poeml \v{6}Indeed, that glory\fnote{\fbackref{10:6} Lit. \fbib{Indeed, it}} will be carried to Assyria--- \\
\poemll    it will become a present for an avenging king.\fnote{\fbackref{10:6} Cf. 2Kings 15-16; 2Chr 28:19-20} \\
\poeml Ephraim will be disgraced, \\
\poemll    and Israel will become ashamed of its decision. \\
\poeml \v{7}Samaria's king will float away \\
\poemll    like driftwood on the surface of water. \\
\poeml \v{8}Destroyed will be the high places of Aven, \\
\poemll    that are the sin of Israel. \\
\poemlll       Both thorn and thistle will grow up over their altars. \\
\poeml They will call out to the mountains, `Cover us!' \\
\poemll    and to the hills, `Fall on us!' \\
\poeml \v{9}``From the time of Gibeah, \\
\poemll    you have sinned, Israel; \\
\poeml There they took their stand; \\
\poemll    the battle at Gibeah could not subdue the lawless. \\
\poeml \v{10}When I'm ready, I will chasten them; \\
\poemll    and the people will gather against them, \\
\poemlll       to imprison them for their two unrighteous acts.''\fnote{\fbackref{10:10} So LXX. MT reads \fbib{two eyes}}
\passage{Israel Urged to Sow in Righteousness}
\poeml \v{11}``Ephraim, the well-trained heifer, \\
\poemll    loves to thresh grain, \\
\poeml so I will spare her neck. \\
\poemll    I will turn Ephraim into a pack animal. \\
\poeml Judah will pull the plow, \\
\poemll    and Jacob will turn up the fallow ground. \\
\poeml \v{12}Sow in righteousness in your own interest, \\
\poemll    reap in gracious love, \\
\poemlll       break up your own unprepared ground; \\
\poeml It is now time to inquire of\fnote{\fbackref{10:12} Or \fbib{to worship}} the \divine{Lord}, \\
\poemll    until he comes to pour out righteousness for you. \\
\poeml \v{13}You have plowed\fnote{\fbackref{10:13} Or \fbib{fashioned}} evil; \\
\poemll    you have reaped unrighteousness; \\
\poemlll       you have eaten the fruit of hypocrisy; \\
\poeml because you trusted in your own direction, \\
\poemll    and in the number of your mighty forces. \\
\poeml \v{14}Therefore a disaster will come upon your people, \\
\poemll    and all of your fortresses will be ruined. \\
\poeml As Shalman\fnote{\fbackref{10:14} Possibly Shalmaneser, king of Assyria; cf. 2Kings 17:3} destroyed Beth-arbel in wartime, \\
\poemll    mothers were\fnote{\fbackref{10:14} Lit. \fbib{a mother was}} dashed to pieces \\
\poemlll       along with their children. \\
\poeml \v{15}The same will happen to you, Bethel, \\
\poemll    because of your great evil--- \\
\poemlll       early one morning the king of Israel will be totally silenced.''
\end{poetry}
\labelchapt{11}
\passage{God Loves Israel}

\begin{poetry}
\poeml \chapt{11}
\v{1}``When Israel was a young child I loved him, \\
\poemll    and from Egypt I called my son. \\
\poeml \v{2}The more I called out to them, \\
\poemll    the farther they fled from me;\fnote{\fbackref{11:2} So LXX. MT reads \fbib{They called to them, so they went away from them.}} \\
\poeml they sacrificed to Baals,\fnote{\fbackref{11:2} I.e. Canaanite deities} \\
\poemll    burning incense to carved images. \\
\poeml \v{3}Yet it was I who taught Ephraim to walk, \\
\poemll    supporting them by their arms, \\
\poemlll       but they never knew that I was healing them. \\
\poeml \v{4}I guided them with human kindness, \\
\poemll    with loving reins. \\
\poeml I acted toward them \\
\poemll    like one who removes a yoke from their neck; \\
\poemlll       I bent down and fed them. \\
\poeml \v{5}``They will not return to the land of Egypt; \\
\poemll    instead, the Assyrian will be their king, \\
\poemlll       because they kept refusing to repent. \\
\poeml \v{6}The sword will fall on their cities, \\
\poemll    consuming and devouring their fortified gates,\fnote{\fbackref{11:6} Lit. \fbib{devouring the bars}} \\
\poemlll       despite their planning. \\
\poeml \v{7}My people are determined to turn away from me; \\
\poemll    though they call to the Most High, \\
\poemlll       no one is worshiping. \\
\poeml \v{8}``How can I give up on you, Ephraim? \\
\poemll    I will deliver you, will I not, Israel? \\
\poeml How can I treat you like Admah? \\
\poemll    I can't make you like Zeboim,\fnote{\fbackref{11:8} cf. Deut 29:23} can I? \\
\poeml My heart stirs within me; \\
\poemll    my compassion also fans into flame! \\
\poeml \v{9}I will not act in my anger; \\
\poemll    I will not return to destroy Ephraim, \\
\poeml For I am God, \\
\poemll    and not a human--- \\
\poeml the Holy One among you--- \\
\poemll    so I will not enter the city in anger.\fnote{\fbackref{11:9} The Heb. lacks \fbib{in anger}} \\
\poeml \v{10}They will go after the \divine{Lord}, \\
\poemll    who will roar like a lion; \\
\poeml and when he roars, \\
\poemll    the children will come trembling from the west. \\
\poeml \v{11}Trembling like a bird, they will come out of Egypt, \\
\poemll    and as a dove from the land of Assyria; \\
\poeml and I will settle them in their houses,'' \\
\poemll    declares the \divine{Lord}. \\
\poeml \v{12}\fnote{\fbackref{11:12} This v. is 12:1 in MT}``Ephraim surrounds me with lies, \\
\poemll    and the house of Israel surrounds me\fnote{\fbackref{11:12} The Heb. lacks \fbib{surrounds me}} with deceit, \\
\poeml But Judah still rules with God, \\
\poemll    and remains faithful, along with the godly ones.''
\end{poetry}
\labelchapt{12}
\passage{Israel's Sin}

\begin{poetry}
\poeml \chapt{12}
\v{1}\fnote{\fbackref{12:1} This v. is 12:2 in MT, and so throughout the chapter.}``Ephraim feeds on the wind, \\
\poemll    chasing after the eastern winds, \\
\poemlll       storing up lies and desolation day after day. \\
\poeml They are making a contract with the Assyrians, \\
\poemll    and sending oil to Egypt. \\
\poeml \v{2}The \divine{Lord} accuses Judah, \\
\poemll    and will punish Jacob according to his ways; \\
\poemlll       he will repay him for what he does. \\
\poeml \v{3}He circumvented his brother\fnote{\fbackref{12:3} Lit. \fbib{He grabbed his brother by the heal}} in the womb, \\
\poemll    and as an adult he fought with God. \\
\poeml \v{4}He even fought the angel and won; \\
\poemll    he cried and prayed to him. \\
\poeml Then at Bethel he found him, \\
\poemll    and there he spoke with us--- \\
\poeml \v{5}the \divine{Lord} God of the Heavenly Armies--- \\
\poemll    the \divine{Lord} is his name.\fnote{\fbackref{12:5} Or \fbib{his traditional name}} \\
\poeml \v{6}So you, return to your God; \\
\poemll    guard grace and justice, \\
\poemlll       and look to your God always. \\
\poeml \v{7}``Now as for the merchant,\fnote{\fbackref{12:7} MT word for \fbib{merchant} sounds like \fbib{Canaan}} \\
\poemll    deceitful balances remain in his hand, \\
\poemlll       and he loves to defraud. \\
\poeml \v{8}Ephraim claims, \\
\poemll    `I have become rich, \\
\poeml I have made a fortune! \\
\poemll    Because of all my wealth, \\
\poemlll       no one will find any iniquity or sin in me.' \\
\poeml \v{9}``Yet I remain the \divine{Lord} your God, \\
\poemll    who brought you out of the land of Egypt. \\
\poeml I will make you live in tents again, \\
\poemll    as in the festival of that name.\fnote{\fbackref{12:9} I.e. the Festival of Tents} \\
\poeml \v{10}I spoke to the prophets, \\
\poemll    giving revelation after revelation, \\
\poemlll       and employing parables in the prophetic writings.\fnote{\fbackref{12:10} Lit. \fbib{parables by the hand of the prophets}} \\
\poeml \v{11}``There's iniquity in Gilead, isn't there? \\
\poemll    They have become truly vain. \\
\poeml They sacrifice bulls in Gilgal; \\
\poemll    their altars are like piles of stone in furrowed fields. \\
\poeml \v{12}Jacob fled into the land of Aram;\fnote{\fbackref{12:12} I.e. Syria} \\
\poemll    Israel served there to obtain his wife, \\
\poemlll       tending sheep to gain\fnote{\fbackref{12:12} The Heb. lacks \fbib{to gain}} his wife. \\
\poeml \v{13}``By a prophet the \divine{Lord} brought Israel out of Egypt, \\
\poemll    and by a prophet he\fnote{\fbackref{12:13} I.e. Israel} was rescued. \\
\poeml \v{14}Ephraim has stirred up violent anger; \\
\poemll    therefore the guilt of his blood will remain on him, \\
\poemlll       and his Lord will repay him for his contempt.''
\end{poetry}
\labelchapt{13}
\passage{The \divine{Lord}'s Anger against Israel}

\begin{poetry}
\poeml \chapt{13}
\v{1}``When the tribe of\fnote{\fbackref{13:1} The Heb. lacks \fbib{the tribe of}} Ephraim spoke, there was trembling; \\
\poemll    and it was exalted within Israel. \\
\poeml But when they offended God by Baal, \\
\poemll    they died, \\
\poeml \v{2}but now they are sinning more and more, \\
\poemll    crafting idols from melted silver. \\
\poeml Their idols are made with the most exacting skill, \\
\poemll    all of it the work of craftsmen. \\
\poeml People\fnote{\fbackref{13:2} Lit. \fbib{They}} say about them, \\
\poemll    `They offer human sacrifice, \\
\poemlll       and kiss calf-shaped idols.'\fnote{\fbackref{13:2} Lit. \fbib{kissing calves}} \\
\poeml \v{3}Therefore they will be like morning clouds, \\
\poemll    like early morning dew that evaporates, \\
\poeml like chaff blown away from the threshing floor, \\
\poemll    or like smoke from a chimney.''\fnote{\fbackref{13:3} Or \fbib{window}}
\passage{The \divine{Lord} is Israel's God}
\poeml \v{4}``I am the \divine{Lord} your God \\
\poemll    from the land of Egypt, \\
\poeml and you have known no god except for me, \\
\poemll    because except for me there is no savior. \\
\poeml \v{5}I took care of\fnote{\fbackref{13:5} So LXX. MT reads \fbib{I knew}} you in the wilderness, \\
\poemll    in a waterless land. \\
\poeml \v{6}As their pastures flourished, \\
\poemll    all their desires were met.\fnote{\fbackref{13:6} Lit. \fbib{flourished, they were satiated}} \\
\poeml As they were satiated, \\
\poemll    they became arrogant \\
\poemlll       and therefore ignored me. \\
\poeml \v{7}``So I will be like a lion to them. \\
\poemll    Like a leopard I will stalk them along the road. \\
\poeml \v{8}I will confront them like a bear deprived of her cubs;\fnote{\fbackref{13:8} The Heb. lacks \fbib{of her cubs}} \\
\poemll    I will tear open their ribs. \\
\poeml I will devour them like a lion--- \\
\poemll    the wild beasts will rip them apart. \\
\poeml \v{9}``You have destroyed yourself, Israel, \\
\poemll    although I remain your help. \\
\poeml \v{10}Now where is your king? \\
\poemll    Will he save you in all your cities? \\
\poeml And where are\fnote{\fbackref{13:10} The Heb. lacks \fbib{where are}} your judges, \\
\poemll    about whom you demanded, \\
\poemlll       `Give me a king and officials!'? \\
\poeml \v{11}I gave you a king in my anger, \\
\poemll    and I took him\fnote{\fbackref{13:11} The Heb. lacks \fbib{him}} away in my fury.'' \\
\poeml \v{12}``Ephraim's guilt is on record; \\
\poemll    his sin is stored away. \\
\poeml \v{13}When the time of childbirth comes, \\
\poemll    he will be so foolish \\
\poemlll       that he will refuse to be born.'' \\
\poeml \v{14}``From the power of Sheol I will rescue them, \\
\poemll    from death I will redeem them. \\
\poeml Death, where are\fnote{\fbackref{13:14} Or \fbib{death, I will be}} your plagues? \\
\poemll    Sheol, where is\fnote{\fbackref{13:14} Or \fbib{Sheol, I will be}} your destruction? \\
\poemlll       My eyes will remain closed to your pleas for\fnote{\fbackref{13:14} The Heb. lacks \fbib{your pleas for}} compassion. \\
\poeml \v{15}Even though he is fruitful compared to his relatives, \\
\poemll    an east wind will come, \\
\poeml the \divine{Lord}'s wind storm from the wilderness, \\
\poemll    and his spring will evaporate. \\
\poeml His fountain will dry up, \\
\poemll    and the \divine{Lord}'s\fnote{\fbackref{13:15} Lit. \fbib{and it}} wind storm will plunder \\
\poemlll       all the expensive vessels of the treasury. \\
\poeml \v{16}\fnote{\fbackref{13:16} This v. is 14:1 in MT}Samaria will be held guilty, \\
\poemll    because she has rebelled against her God. \\
\poeml By the sword they will fall--- \\
\poemll    with their infants dashed to pieces, \\
\poemlll       and their pregnant women torn open.''
\end{poetry}
\labelchapt{14}
\passage{A Call to Repentance}

\begin{poetry}
\poeml \chapt{14}
\v{1}\fnote{\fbackref{14:1} This v. is 14:2 in MT, and so throughout the chapter.}``Return, Israel, to the \divine{Lord} your God, \\
\poemll    for you have fallen due to your own iniquity. \\
\poeml \v{2}Bring a prepared speech with you \\
\poemll    as you return to the \divine{Lord}. Say to him: \\
\poeml `Take away all our\fnote{\fbackref{14:2} The Heb. lacks \fbib{our}} iniquity, \\
\poemll    and accept what is good. \\
\poeml Then we will present the fruit\fnote{\fbackref{14:2} So LXX and DSS. MT reads \fbib{bulls}} of our lips. \\
\poeml \v{3}Assyria won't save us; \\
\poemll    we won't be riding on horses, \\
\poeml Nor will we be saying anymore to the work of our hands, \\
\poemll    ``You are\fnote{\fbackref{14:3} The Heb. lacks \fbib{You are}} our God.'' \\
\poemlll       Indeed, in you the orphan finds mercy.' \\
\poeml \v{4}``I will correct their apostasy, \\
\poemll    loving them freely, \\
\poemlll       since my anger will have turned away from them.\fnote{\fbackref{14:4} Lit. \fbib{him}} \\
\poeml \v{5}I will be like the dew to Israel; \\
\poemll    Israel\fnote{\fbackref{14:5} Lit. \fbib{he}} will blossom like a lily, \\
\poemlll       growing roots like the cedars of\fnote{\fbackref{14:5} The Heb. lacks \fbib{the cedars of}} Lebanon. \\
\poeml \v{6}Israel's\fnote{\fbackref{14:6} Lit. \fbib{His}} branches will spread out, \\
\poemll    and its beauty will be like an olive tree, \\
\poemlll       with its scent like that of Lebanon. \\
\poeml \v{7}Those who live under its protection\fnote{\fbackref{14:7} Lit. \fbib{shadow}} will surely return. \\
\poemll    Their grain will flourish; \\
\poeml they will blossom like a vine, \\
\poemll    and Israel's\fnote{\fbackref{14:7} Lit. \fbib{His}} scent will be like wine from Lebanon. \\
\poeml \v{8}``Ephraim, what have I in common with idols? \\
\poemll    I have listened and will pay attention to him. \\
\poeml I am like a flourishing cypress; \\
\poemll    in me will your fruit be found.''
\passage{Concluding Counsel}
\poeml \v{9}Whoever is wise, let him understand these things. \\
\poemll    Whoever is discerning, let him know them. \\
\poeml For the ways of the \divine{Lord} are right: \\
\poemll    the righteous follow his example, \\
\poemlll       but the rebellious stumble in them.\end{poetry}
