\bookheader{Obadiah}
\labelbook{Obad}

\bookpretitle{The Book of the Prophet}
\booktitle{Obadiah}

\passage{Coming Judgment against Edom}

\begin{poetry}
\poeml \v{1}Obadiah's\fnote{\fbackref{1} The Heb. name \fbib{Obadiah} means \fbib{Servant of the \divine{Lord}}} vision: \\
\poeml This is what the Lord \divine{God} has to say about Edom. \\
\poeml We have heard a report from the \divine{Lord}, \\
\poemll    and a messenger has been dispatched among the nations to say\fnote{\fbackref{1} The Heb. lacks \fbib{to say}} \\
\poeml ``Get up! Let us rise up against her to fight!''
\passage{God's Announcement to Edom}
\poeml \v{2}``Look! I will make you insignificant among the nations; \\
\poemll    you will be utterly despised. \\
\poeml \v{3}The arrogance in your heart has deceived you, \\
\poemll    who inhabit hidden places on rocky cliffs, \\
\poemll    whose dwelling is in the heights, \\
\poemll    who say continuously to yourself,\fnote{\fbackref{3} Lit. \fbib{continually in your heart}} \\
\poemlll       `Who will bring me down to the ground?' \\
\poeml \v{4}Though you soar high like the eagle \\
\poemll    and make your nest among the stars, \\
\poeml I will bring you down even from there,'' \\
\poemlll       declares the \divine{Lord}.\fnote{\fbackref{4} Cf. Jer 49:14-16}
\passage{The Harvest from Edom's Arrogance}
\poeml \v{5}``If thieves came against you, \\
\poemll    if marauding gangs by night \\
\poemlll       ---Oh, how you will be destroyed!--- \\
\poeml Would they not steal only until they had enough? \\
\poeml If grape pickers came to you, \\
\poemll    would they not leave some\fnote{\fbackref{5} The Heb. lacks \fbib{some}} grapes to be gleaned? \\
\poeml \v{6}``Oh, how Esau is ransacked, \\
\poemll    how his hidden treasures are thoroughly\fnote{\fbackref{6} The Heb. lacks \fbib{thoroughly}} searched out! \\
\poeml \v{7}All your allies will force you out of the land,\fnote{\fbackref{7} Lit. \fbib{out to the border}} \\
\poemll    your associates will deceive you and prevail against you. \\
\poeml Your friends\fnote{\fbackref{7} Lit. \fbib{Your bread}; i.e. those who eat your bread} will lay out a trap for you, \\
\poemll    and you will\fnote{\fbackref{7} The Heb. lacks \fbib{you will}} never understand it! \\
\poeml \v{8}``In that day,'' declares the \divine{Lord}, \\
\poemll    ``will I not destroy the wise from Edom, \\
\poemll    and those with understanding from Esau's Mountain? \\
\poeml \v{9}Teman, our mighty soldiers will be dismayed, \\
\poemll    so that every man from Esau's Mountain will be slaughtered.''\fnote{\fbackref{9} Lit. \fbib{will be cut off by slaughter}}
\passage{Judgment for Edom's Cruelty to Jacob}
\poeml \v{10}``Shame will overwhelm you \\
\poemll    because of the violence you inflicted on your brother Jacob, \\
\poeml and you will be excluded\fnote{\fbackref{10} Lit. \fbib{cut off}} forever. \\
\poeml \v{11}``On the day you just stood by,\fnote{\fbackref{11} Or \fbib{stood in opposition}} \\
\poeml when\fnote{\fbackref{11} Lit. \fbib{in the day}} strangers carried away Jacob's\fnote{\fbackref{11} Lit. \fbib{his}} wealth \\
\poeml and foreigners entered his gates, \\
\poemll    casting lots for Jerusalem, \\
\poeml you were just like one of them. \\
\poeml \v{12}``You should not have gloated over your brother,\fnote{\fbackref{12} Lit. \fbib{in the day of your brother}} \\
\poemll    on the day of his calamity. \\
\poeml You should not have rejoiced \\
\poemll    when\fnote{\fbackref{12} Lit. \fbib{in the day}} the descendants of Judah were perishing. \\
\poeml You should not have boasted\fnote{\fbackref{12} Lit. \fbib{have let your mouth boast}} \\
\poemll    when\fnote{\fbackref{12} Lit. \fbib{in the day}} they were in distress. \\
\poeml \v{13}``You should not have entered the gate of my people \\
\poemll    on the day of their disaster.\fnote{\fbackref{13} The Heb. words \fbib{their disaster} may be a word play on the Heb. word \fbib{Edom}} \\
\poeml Also, you should not have gloated over Judah's\fnote{\fbackref{13} Lit. \fbib{his}} misfortune \\
\poemll    on the day of his disaster,\fnote{\fbackref{13} The Heb. words \fbib{his disaster} may be a word play on the Heb. word \fbib{Edom}} \\
\poeml nor should you have plundered his wealth \\
\poemll    on the day of his disaster.\fnote{\fbackref{13} The Heb. words \fbib{his disaster} may be a word play on the Heb. word \fbib{Edom}} \\
\poeml \v{14}And you should not have taken your stand at the crossroads \\
\poemll    to cut down his fleeing refugees, \\
\poeml nor should you have handed over his survivors \\
\poemll    on the day of his distress.''
\passage{The \divine{Lord}'s Judgment and Israel's Final Victory}
\poeml \v{15}``Indeed, the Day of the \divine{Lord} approaches all nations. \\
\poemll    As you have done it will be done to you--- \\
\poemlll       your deeds will return to haunt you!\fnote{\fbackref{15} Lit. \fbib{return on your own head}} \\
\poeml \v{16}Just as you have drunk from the cup of my wrath\fnote{\fbackref{16} The Heb. lacks \fbib{from the cup of my wrath}} upon my holy mountain, \\
\poemll    so will all nations drink from the cup of my wrath\fnote{\fbackref{16} The Heb. lacks \fbib{from the cup of my wrath}} perpetually. \\
\poeml They will drink, they will gulp it down, \\
\poemll    and they will be as if they had never existed! \\
\poeml \v{17}``But there will be a delivered remnant on Mount Zion. \\
\poemll    There will be holiness, \\
\poemlll       and the house of Jacob will take back their possessions. \\
\poeml \v{18}``The house of Jacob will be a fire, \\
\poeml and the house of Joseph a flame, \\
\poemll    but the house of Esau will be kindling. \\
\poeml Then Jacob and Joseph\fnote{\fbackref{18} Lit. \fbib{They}} will burn and consume Esau,\fnote{\fbackref{18} Lit. \fbib{them}} \\
\poemll    and no survivor will remain from the house of Esau.'' \\
\poeml Indeed, the \divine{Lord} has spoken it. \\
\poeml \v{19}``Those in the Negev\fnote{\fbackref{19} I.e. the southern regions of the Sinai peninsula; cf. Josh 10:40} will possess Esau's Mountain, \\
\poemll    and those in the Shephelah\fnote{\fbackref{19} I.e. the verdant central lowlands of Israel; cf. Josh 10:40} the Philistines. \\
\poeml They will possess the fields of Ephraim \\
\poemll    and the fields of Samaria, \\
\poeml while Benjamin will possess the territory of\fnote{\fbackref{19} The Heb. lacks \fbib{will possess the territory of}} Gilead. \\
\poeml \v{20}The exiles, the Israeli host, \\
\poemll    will possess the territory of the\fnote{\fbackref{20} The Heb. lacks \fbib{the territory of}} Canaanites all the way to Zarephath. \\
\poeml The exiles of Jerusalem who are in Sepharad\fnote{\fbackref{20} I.e. perhaps Sardis, capital of Lydia, Saparda in eastern Assyria, Sparta in Greece, or a location in Spain (so \fbib{Targ of Jonathan})} \\
\poemll    will possess the cities of the Negev.\fnote{\fbackref{20} I.e. the southern regions of the Sinai peninsula; cf. Josh 10:40} \\
\poeml \v{21}Deliverers will assemble on Mount Zion to judge Esau's Mountain, \\
\poemll    and to the \divine{Lord} will the kingdom belong!''\end{poetry}
