\bookheader{Micah}
\labelbook{Mic}

\bookpretitle{The Book of the Prophet}
\booktitle{Micah}

\labelchapt{1}
\passage{God's Coming Judgment}

\chapt{1}
\v{1}This message from the \divine{Lord} came to Micah\fnote{The Heb. name \fbib{Micah} means \fbib{Who is like the \divine{Lord}?}} of Moresheth during the reigns of\fnote{Lit. \fbib{days}} the Judean kings Jotham, Ahaz, and Hezekiah concerning the vision he saw about Samaria and Jerusalem:

\begin{poetry}
\poeml \v{2}``Listen, people! All of you! \\
\poemll    Earth! Pay attention, and all you inhabitants of it! \\
\poeml May the Lord \divine{God} be a witness against you--- \\
\poemll    the Lord from his holy Temple. \\
\poeml \v{3}Look here! The \divine{Lord} is coming from his place! \\
\poemll    He will come down \\
\poemlll       and will trample down the high places\fnote{I.e. the sites of idol worship} throughout the land. \\
\poeml \v{4}The mountains will melt under him \\
\poemll    and the valleys will split apart, \\
\poeml like wax in the presence of fire \\
\poemll    and like water gushing down a steep incline. \\
\poeml \v{5}All this comes about due to the transgression of Jacob, \\
\poemll    and due to the sins of the house of Israel. \\
\poeml What is Jacob's sin? It's Samaria, isn't it? \\
\poemll    And what's Judah's high place?\fnote{I.e. the sites of idol worship} It's Jerusalem, isn't it? \\
\poeml \v{6}``So I will turn Samaria into a mound of dirt in a field, \\
\poemll    a place to plant vineyards. \\
\poeml And I will dump her building stones into the valley, \\
\poemll    uncovering her foundation. \\
\poeml \v{7}All of her carved images will be crushed to pieces, \\
\poemll    all the earnings of her prostitution will be burned up, \\
\poemlll       and I will destroy all her idols; \\
\poeml because she collected the wages of prostitution, \\
\poemll    and to the wages of prostitution they will return.''
\passage{The Coming Destruction}
\poeml \v{8}``Therefore I will cry out and grieve loudly; \\
\poemll    I will walk around stripped and naked. \\
\poeml I will cry out like a jackal \\
\poemll    and mourn like a company of ostriches. \\
\poeml \v{9}For Samaria's\fnote{Lit. \fbib{her}} injury is fatal, \\
\poemll    reaching all the way to Judah, \\
\poemlll       extending even to the gate of my people---to Jerusalem.'' \\
\poeml \v{10}``Don't discuss it in Gath!\fnote{A city in Philistia} \\
\poemll    Don't cry bitterly in Akim!\fnote{So LXX; MT reads \fbib{Don't cry at all}} \\
\poemlll       Roll in the ashes, Beth-leaphrah! \\
\poeml \v{11}Run away, you residents of Shaphir, \\
\poemll    displaying your shameful nakedness. \\
\poeml Don't come out, you residents of Zaanan!\fnote{A city of Judah} \\
\poemll    Your firm standing will disappear as Beth-ezel mourns. \\
\poeml \v{12}Even though the inhabitants of Maroth long for success, \\
\poemll    nevertheless evil descended from the \divine{Lord} to the gate of Jerusalem. \\
\poeml \v{13}``You inhabitants of Lachish, harness your chariot to your swiftest steed--- \\
\poemll    the daughter of Zion has begun to sin--- \\
\poeml because within you the transgressions of Israel were revealed. \\
\poeml \v{14}Therefore give your gifts to Moresheth-gath; \\
\poemll    that is, the houses of Achzib as a deceitful symbol\fnote{Or \fbib{a lie}} to the kings of Israel. \\
\poeml \v{15}Nevertheless, I will deliver an heir to you, inhabitants of Mareshah--- \\
\poemll    to Adullam the glory of Israel will come. \\
\poeml \v{16}``Shave your head \\
\poemll    and cut off your locks as you mourn your beloved children. \\
\poeml Make yourself bald as an eagle, \\
\poemll    because they will go from you into exile!''
\end{poetry}
\labelchapt{2}
\passage{God's Warning to His People}

\begin{poetry}
\poeml \chapt{2}
\v{1}``Woe to those who are crafting iniquity, \\
\poeml planning evil well into the night!\fnote{Lit. \fbib{evil upon their beds}} \\
\poeml When morning's light comes, \\
\poemll    they carry out their plans\fnote{The Heb. lacks \fbib{their plans}} because they have the power to do so. \\
\poeml \v{2}They covet fields and seize them; \\
\poemll    they covet\fnote{The Heb. lacks \fbib{they covet}} houses, and grab them, too. \\
\poeml They harass the valiant man, along with his household, \\
\poemll    an individual and his estate. \\
\poeml \v{3}``Therefore this is what the \divine{Lord} says, \\
\poeml `I'm crafting evil against this family, \\
\poemll    from which you can't escape.\fnote{Lit. \fbib{can't remove your necks}} \\
\poeml You won't strut around arrogantly, \\
\poemll    because the times are evil.' \\
\poeml \v{4}``When this happens,\fnote{Lit. \fbib{In that day}} someone will compose a proverb about you, lamenting sorrowfully, \\
\poemll    `We are completely ruined! \\
\poemlll       He has given my people's heritage to others.\fnote{The Heb. lacks \fbib{to others}} \\
\poemll    How he has removed it from me, \\
\poemlll       dividing up our fields!' \\
\poeml \v{5}``This is why there will not be left even a single person \\
\poemll    to settle boundary disputes\fnote{Lit. \fbib{to stretch out a measuring line}} in the \divine{Lord}'s community. \\
\poeml \v{6}To those who speak out, they will declare, \\
\poemll    `Don't prophesy to anyone!' \\
\poemlll       so their shame won't go away. \\
\poeml \v{7}``It is said, house of Jacob, \\
\poemll    `The Spirit of the \divine{Lord} is limited, \\
\poemlll       if he acts this way, is he not?' \\
\poeml ``But my words benefit those who live righteously, do they not? \\
\poeml \v{8}Lately my people have acted like an enemy--- \\
\poemll    you strip travelers who thought they were\fnote{The Heb. lacks \fbib{thought they were}} as secure \\
\poemlll       as those who return from war. \\
\poeml \v{9}You have evicted the wives of my people from their dream homes; \\
\poemll    you have removed my majesty from their children permanently. \\
\poeml \v{10}``Get up and go, \\
\poemll    because there's no rest for you here! \\
\poeml Since everything\fnote{Lit. \fbib{it}} is polluted, it can only cause destruction, \\
\poemll    even heavy destruction. \\
\poeml \v{11}Suppose a man who keeps company with a deceiving spirit prophesies like this: \\
\poemll    `Drink wine and strong drink!' \\
\poemlll       Won't the people accept him as a prophet?''
\passage{The Coming Judgment}
\poeml \v{12}``Jacob, how I will surely gather all of you! \\
\poemll    How I will gather the survivors of Israel! \\
\poeml I will gather them together like sheep in a pen,\fnote{Or \fbib{sheep of Bozrah}} \\
\poemll    like the flock in the midst of the sheepfold. \\
\poemlll       There will be a great commotion because of all\fnote{The Heb. lacks \fbib{all}} the people. \\
\poeml \v{13}God will stand up and break through\fnote{Lit. \fbib{The one who breaks through will arise}} in their presence. \\
\poemll    Then they will pass through the gate, going out by it. \\
\poeml Their king will pass in front of them \\
\poemll    with the \divine{Lord} at their head.''
\labelchapt{3}
\poeml \chapt{3}
\v{1}``He will say, `Listen, you leaders of Jacob, \\
\poeml you officials of the house of Israel! \\
\poeml You should know justice, should you not?--- \\
\poeml \v{2}you who despise good and love evil, \\
\poemll    who tear off the skin of my people,\fnote{The Heb. lacks \fbib{of my people}} \\
\poemlll       along with the flesh from their bones. \\
\poeml \v{3}You eat the flesh of my people, \\
\poemll    flaying their skin from them. \\
\poeml You break their bones, \\
\poemll    chopping them in pieces like meat\fnote{The Heb. lacks \fbib{meat}} for a pot, \\
\poemlll       like meat destined for a soup kettle.' \\
\poeml \v{4}``Then they will cry to the \divine{Lord}, \\
\poemll    but he will not listen to them. \\
\poeml In fact, he will hide his face from them at that time, \\
\poemll    because they were so wicked in what they were doing.''
\passage{God's Judgment against False Prophets}
\poeml \v{5}``This is what the \divine{Lord} says about the prophets \\
\poemll    who are causing my people to go astray, \\
\poeml who are calling out `Peace' when they're being fed,\fnote{Lit. \fbib{who bite with their teeth, crying out ``Peace''}} \\
\poemll    but who declare war against those who won't feed them:\fnote{Lit. \fbib{against him who puts nothing in their mouths}} \\
\poeml \v{6}`You will have nights without visions, \\
\poemll    and darkness without prophecy. \\
\poeml The sun will set on the prophets, \\
\poemll    and the day will darken for them. \\
\poeml \v{7}Those who see visions will be put to shame, \\
\poemll    and the diviners will be disgraced---every one of them--- \\
\poeml they will cover their faces,\fnote{Lit. \fbib{beard}} \\
\poemll    because there will be no answer from God.'\,''
\passage{The Message of God's Prophet}
\poeml \v{8}``As for me, I am truly filled with power by the Spirit of the \divine{Lord}, \\
\poemll    filled\fnote{The Heb. lacks \fbib{filled}} with judgment and power \\
\poeml to announce to Jacob his transgression, \\
\poemll    and to Israel his sin. \\
\poeml \v{9}Please listen to this, you leaders of the house of Jacob, \\
\poemll    you officials of the house of Israel, \\
\poeml you who hate administering justice, \\
\poemll    who pervert the very meaning of\fnote{Lit. \fbib{pervert all}} equity, \\
\poeml \v{10}who are building up Zion by means of bloodshed, \\
\poemll    and Jerusalem by means of iniquity. \\
\poeml \v{11}Her leaders judge for the money, \\
\poemll    her priests teach only when they're paid , \\
\poemlll       and her prophets prophesy for cash. \\
\poeml Even so, don't they all rely on the \divine{Lord} as they ask, \\
\poemll    `The \divine{Lord} is among us, is he not? \\
\poemlll       Nothing bad can possibly happen to us!' \\
\poeml \v{12}``Therefore, because of you, Zion will be plowed up like a field, \\
\poemll    and Jerusalem will become heaps of rubble, \\
\poemlll       and the Temple Mount like a forest high place.''
\end{poetry}
\labelchapt{4}
\passage{The Future Reign of God}

\begin{poetry}
\poeml \chapt{4}
\v{1}``But in the last days it will come about \\
\poemll    that the Temple Mount of the \divine{Lord} will be firmly set \\
\poemlll       as the leading mountain. \\
\poeml It will be exalted above its surrounding\fnote{The Heb. lacks \fbib{surrounding}} hills, \\
\poemll    and people will stream toward it. \\
\poeml \v{2}Many nations will approach and say, \\
\poemll    `Come, let's go up to the mountain of the \divine{Lord}, \\
\poemlll       and to the Temple of the God of Jacob. \\
\poemll    He will teach us about his ways, \\
\poemlll       and we will walk according to his directions.' \\
\poeml ``Indeed, the Law will proceed from Zion, \\
\poemll    and the message of the \divine{Lord} from Jerusalem. \\
\poeml \v{3}And he will judge among many people, \\
\poemll    rebuking strong nations far away; \\
\poeml and they will reshape their swords as plowshares \\
\poemll    and their spears as pruning hooks. \\
\poeml No nation will threaten another,\fnote{Lit. \fbib{A nation will not lift up a sword against a nation}} \\
\poemll    nor will they train for war anymore. \\
\poeml \v{4}Instead, each man will sit in the shade of\fnote{Lit. \fbib{sit under}} his grape vines \\
\poemll    and beneath the shade of\fnote{The Heb. lacks \fbib{the shade of}} his fig tree,'' \\
\poemlll       since the\fnote{Lit. \fbib{the mouth of the}} \divine{Lord} of the Heavenly Armies has spoken. \\
\poeml \v{5}``Because all of the people will walk, \\
\poemll    each person in the name of his God, \\
\poemlll       and we will walk in the name of the \divine{Lord} our God forever and ever. \\
\poeml \v{6}``At that time,'' declares the \divine{Lord}, \\
\poemll    ``I will assemble the lame; \\
\poeml and I will gather those whom I have scattered, \\
\poemll    along with those whom I have afflicted. \\
\poeml \v{7}I will transform the lame into survivors, \\
\poemll    and those who were scattered into a strong nation; \\
\poeml and the \divine{Lord} will reign over them in Mount Zion, \\
\poemll    now and forever.''
\passage{Zion's Captivity}
\poeml \v{8}``And you, watchtower of the flock, \\
\poemll    you stronghold of the daughter of Zion, \\
\poeml it will happen even to you: \\
\poemlll       The former dominion, even the kingdom of the daughter of Jerusalem, will come. \\
\poeml \v{9}Why are you crying so loud now? \\
\poemll    There's no king among you, is there? \\
\poeml Perhaps your advisor has died? \\
\poemll    For pain has overtaken you like a woman in labor. \\
\poeml \v{10}Be in pain! Be in labor, you daughter of Zion, \\
\poemll    like a woman about to give birth, \\
\poeml because now you will depart from the city, \\
\poemll    living in the open fields. \\
\poeml To Babylon you will go. \\
\poemll    There you will be delivered, \\
\poemlll       there the \divine{Lord} will rescue you from the power\fnote{Lit. \fbib{hand}} of your enemies.''
\passage{The Nations Despise Zion}
\poeml \v{11}``Now many nations have gathered against you, saying, \\
\poemll    `Let her be defiled,' and \\
\poemlll       `Let's look down on Zion.' \\
\poeml \v{12}But they don't know the thoughts of the \divine{Lord}, \\
\poemll    and they don't understand his tactics, \\
\poemlll       for he will gather them like harvested grain to his threshing floor. \\
\poeml \v{13}Get up and smash\fnote{Lit. \fbib{thresh}} them to pieces, daughter of Zion, \\
\poemll    because I will make your horn like iron \\
\poemlll       and your hooves like bronze! \\
\poeml And you will beat many people to pieces, \\
\poemll    and I will consecrate their dishonest\fnote{The Heb. lacks \fbib{dishonest}} gain to the \divine{Lord} \\
\poemlll       and their assets to the Lord of the entire earth.''
\labelchapt{5}
\poeml \chapt{5}
\v{1}\fnote{5:1 is 4:14 in MT}``Now marshal yourselves as troops.\fnote{Lit. \fbib{yourselves, you daughter of mobs}} \\
\poeml He has laid siege to us. \\
\poemll    They will strike the judge\fnote{Or \fbib{ruler}} of Israel on the cheek with a rod.''
\passage{The Ruler from Bethlehem}
\poeml \v{2}\fnote{5:2 is 5:1 in MT, 5:3 is 5:2 in MT, and so throughout the chapter}``As for you, Bethlehem of Ephrathah, \\
\poemll    even though you remain least among the clans\fnote{Or \fbib{thousands}} of Judah, \\
\poeml nevertheless, the one who rules in Israel for me \\
\poemll    will emerge from you. \\
\poeml His existence has been\fnote{The Heb. lacks \fbib{has been}} from antiquity, \\
\poemll    even from eternity. \\
\poeml \v{3}Therefore that ruler\fnote{Lit. \fbib{Therefore he}} will abandon them \\
\poemll    until the woman in labor gives birth. \\
\poeml Then the rest of his countrymen will return to the Israelis.'' \\
\poeml \v{4}``Then he will take his stand, \\
\poemll    shepherding by means of the strength of the \divine{Lord}, \\
\poemlll       by the power\fnote{Or \fbib{majesty}} of the name of the \divine{Lord} his God. \\
\poeml And they will be firmly\fnote{The Heb. lacks \fbib{firmly}} established; \\
\poemll    indeed, from then on he will become great--- \\
\poemlll       to the ends of the earth. \\
\poeml \v{5}And he will be our peace.''
\passage{God's Judgment on Assyria}
\poeml ``When the Assyrian invades our land, \\
\poemll    trampling through our palaces, \\
\poeml we will raise up seven shepherds against him, \\
\poemll    even eight significant men. \\
\poeml \v{6}The shepherds\fnote{Lit. \fbib{And they}} will devastate the land of Assyria with the sword, \\
\poemll    along with the entrances to the land of Nimrod. \\
\poeml ``This is how he will vanquish\fnote{Lit. \fbib{will deliver from}} Assyria \\
\poemll    when he invades our land, \\
\poemlll       trampling within our borders: \\
\poeml \v{7}The survivors of Jacob will live among many nations, \\
\poemll    as dew from the \divine{Lord}, \\
\poemlll       as showers upon the grass. \\
\poeml They will look to no one, \\
\poemll    and will place no hope in human beings. \\
\poeml \v{8}The survivors of Jacob will live among the nations; \\
\poemll    they will live among many nations, \\
\poemlll       like a lion among flocks of sheep, \\
\poeml who, if he passes through, \\
\poemll    will trample and tear down \\
\poemlll       with no one to deliver. \\
\poeml \v{9}You will turn your power\fnote{Lit. \fbib{hand}} against your adversaries, \\
\poemll    and all of your enemies will be cut down.''
\passage{God Removes Idolatry}
\poeml \v{10}``It will come about at that time,'' declares the \divine{Lord}, \\
\poemll    ``I will tear away your horses from you, \\
\poemlll       and I will destroy your chariots. \\
\poeml \v{11}I will cut off the cities of your land, \\
\poemll    and I will tear down all of your fortifications. \\
\poeml \v{12}I will render your witchcraft powerless,\fnote{Lit. \fbib{will cut witchcraft out of your hands}} \\
\poemll    and mediums will no longer exist among you. \\
\poeml \v{13}I will separate you from your carved images and sacred pillars, \\
\poemll    and you no longer will worship \\
\poemlll       what you've made with your hands. \\
\poeml \v{14}I will uproot your cultic gods\fnote{Heb. \fbib{Asherim}} from you, \\
\poemll    and I will tear down your cities. \\
\poeml \v{15}I will execute vengeance, anger, and fury \\
\poemll    on the nations who do not obey.''
\end{poetry}
\labelchapt{6}
\passage{The \divine{Lord}'s Indictment against Israel}

\chapt{6}
\v{1}Please hear what the \divine{Lord} says:

\begin{poetry}
\poemll    ``Get up and make your case before the mountains, \\
\poemlll       and let the hills listen to your voice. \\
\poeml \v{2}Listen, you mountains, to the \divine{Lord}'s argument! \\
\poemll    Listen, you\fnote{The Heb. lacks \fbib{Listen, you}} strong foundations of the earth, \\
\poeml because the \divine{Lord} has a dispute with his people, \\
\poemll    and he will set out his case before Israel. \\
\poeml \v{3}``My people, what have I done to you, \\
\poemll    and how have I offended you? \\
\poemlll       Answer me! \\
\poeml \v{4}For I brought you up from the land of Egypt, \\
\poemll    and ransomed you from the house of slavery, \\
\poemlll       sending Moses, Aaron, and Miriam into your presence. \\
\poeml \v{5}``My people, recall how king Balak of Moab deliberated, \\
\poemll    and how Beor's son Balaam counseled him from Shittim to Gilgal, \\
\poemlll       so that you may know the righteousness of the \divine{Lord}.''
\passage{The Nature of True Righteousness}
\poeml \v{6}How am I to present myself in the \divine{Lord}'s presence \\
\poemll    and bow in the presence of the High God? \\
\poeml Should I present myself with burnt offerings, \\
\poemll    with year-old calves? \\
\poeml \v{7}Will the \divine{Lord} be pleased with thousands of rams, \\
\poemll    or with endless\fnote{Lit. \fbib{with ten thousand}} rivers of oil? \\
\poeml Am I to give my firstborn to pay for\fnote{The Heb. lacks \fbib{to pay for}} my rebellion, \\
\poemll    the fruit of my body in exchange for\fnote{The Heb. lacks \fbib{in exchange for}} my soul? \\
\poeml \v{8}He has made it clear to you, mortal man, what is good \\
\poemll    and what the \divine{Lord} is requiring from you--- \\
\poeml to act with justice, \\
\poemll    to treasure the \divine{Lord}'s\fnote{The Heb. lacks \fbib{the \divine{Lord}'s}} gracious love, \\
\poemlll       and to walk humbly in the company of your God.
\passage{A Call to Honest Business Practices}
\poeml \v{9}The voice of the \divine{Lord} cries out to the city--- \\
\poemll    wisdom fears your name: \\
\poemlll       ``Heed the rod, and the one who prepared it! \\
\poeml \v{10}Are there still wicked treasures in the house of the wicked, \\
\poemll    along with deceitful and abominable measuring standards?\fnote{Lit. \fbib{with short, detestable ephahs}} \\
\poeml \v{11}Will I tolerate those who maintain deceptive standards\fnote{Lit. \fbib{maintain evil balances}} \\
\poemll    and who use deceitful weights in their business?\fnote{The Heb. lacks \fbib{in their business}} \\
\poeml \v{12}Her rich people are filled with violence, \\
\poemll    and her inhabitants tell lies--- \\
\poemlll       their tongues speak deceitfully! \\
\poeml \v{13}``Therefore I will make you ill when I attack you; \\
\poemll    I will bring you to ruin because of your offenses. \\
\poeml \v{14}You'll eat, \\
\poemll    but you won't have enough; \\
\poemlll       and hunger will be common among you. \\
\poeml You'll horde things, \\
\poemll    but you won't save them, \\
\poeml and what you preserve \\
\poemll    I'll give over to the sword. \\
\poeml \v{15}You'll plant, \\
\poemll    but you won't reap. \\
\poeml You'll crush the olive harvest, \\
\poemll    but you'll have no oil to anoint yourself. \\
\poeml You'll tread out the grapes, \\
\poemll    but you'll never drink wine. \\
\poeml \v{16}You keep Omri's\fnote{I.e. King Omri of Israel, father of Ahab (cf. 2Kings 8:26; 2Chr 22:2)} statutes \\
\poemll    and observe the customs of the house of Ahab. \\
\poeml Because you live according to their advice, \\
\poemll    I'll make you desolate \\
\poemlll       and turn your\fnote{Lit. \fbib{its}} inhabitants into an object of scorn. \\
\poeml Therefore you will bear the shame of my people.''
\end{poetry}
\labelchapt{7}
\passage{The Evil Behavior of the People}

\begin{poetry}
\poeml \chapt{7}
\v{1}Poor me! \\
\poeml I feel like those who harvest summer fruit, \\
\poemlll       or like those who pick grapes--- \\
\poeml there are no clusters to eat \\
\poemll    or any fresh fruit that I want. \\
\poeml \v{2}The faithful have died off, \\
\poemll    and there is not one upright human being in the land. \\
\poeml They all stalk one another\fnote{The Heb. lacks \fbib{one another}} with lethal intent, \\
\poemll    a man will even hunt his own brother with a net. \\
\poeml \v{3}And speaking of evil, \\
\poemll    they practice it eagerly---with both hands! \\
\poeml Both leader and judge demand a bribe, \\
\poemll    the famous confess their perverted desires, \\
\poemlll       and they scheme together. \\
\poeml \v{4}The best of them is like a thorn, \\
\poemll    and their most upright like a hedge of thorns. \\
\poeml The day announced by\fnote{Lit. \fbib{day of}} your watchmen--- \\
\poemll    and by your own calculations---approaches. \\
\poemlll       Now it's your time to be\fnote{The Heb. lacks \fbib{time to be}} confused! \\
\poeml \v{5}Don't trust your friends, \\
\poemll    don't confide in a companion, \\
\poemlll       watch what you say to your wife.\fnote{Lit. \fbib{her who lies near your bosom}} \\
\poeml \v{6}The son disrespects his father, \\
\poemll    the daughter rebels against her mother, \\
\poeml the daughter-in-law against her mother-in-law, \\
\poemll    and a man's enemies are the people of his own house.\fnote{Cf. Matt 10:34-36}
\passage{Micah Looks to God}
\poeml \v{7}But as for me, I will look to the \divine{Lord}; \\
\poemll    I will wait for the God who will deliver me. \\
\poemlll       My God will hear me. \\
\poeml \v{8}Don't be glad on my account, my enemy. \\
\poemll    When I fall, I'll get up. \\
\poeml Though I sit in darkness, \\
\poemll    the \divine{Lord} is a light for me. \\
\poeml \v{9}I will endure the \divine{Lord}'s anger--- \\
\poemll    since I have sinned against him--- \\
\poeml until he takes over my defense, \\
\poemll    administers justice on my behalf, \\
\poeml and brings me out to the light, \\
\poemll    where I will gaze on his righteousness. \\
\poeml \v{10}Then my enemy will observe it, \\
\poemll    and shame will engulf the ones\fnote{Lit. \fbib{will bury her}} who asked me, \\
\poemlll       `Where is the \divine{Lord} your God?' \\
\poeml My own eyes will see them,\fnote{Lit. \fbib{her}} \\
\poemll    they\fnote{Lit. \fbib{she}} will be trampled on like mud in the streets.
\passage{A Word of Restoration}
\poeml \v{11}When the time comes\fnote{Lit. \fbib{In the day}} for rebuilding your walls, \\
\poemll    that time\fnote{Lit. \fbib{in the day}} will surely be extended.\fnote{Or \fbib{that day the decree will be removed far away}} \\
\poeml \v{12}At that time\fnote{Lit. \fbib{At that day}} armies\fnote{Lit. \fbib{they}} will invade you from Assyria, \\
\poemll    from Egyptian cities to the Euphrates\fnote{The Heb. lacks \fbib{Euphrates}} River, \\
\poeml from sea to sea \\
\poemll    and from mountain to mountain. \\
\poeml \v{13}The land will become desolate \\
\poemll    because of its inhabitants, \\
\poemlll       and as a result of their behavior. \\
\poeml \v{14}Use your rod to shepherd your people, \\
\poemll    the flock that belongs to you, \\
\poemlll       that lives alone in the forest of Carmel. \\
\poeml Let them find pasture in Bashan and Gilead, \\
\poemll    as they did long ago. \\
\poeml \v{15}As I did when\fnote{Lit. \fbib{As in the days}} you came out of the land of Egypt, \\
\poemll    I will show you\fnote{Lit. \fbib{him}} awesome things. \\
\poeml \v{16}The nations will look on \\
\poemll    and will be ashamed in spite of all their power; \\
\poeml they will cup their hands over their mouths, \\
\poemll    and their ears will be deaf. \\
\poeml \v{17}They will lick the dust like a serpent; \\
\poemll    they will crawl from their strongholds like snakes. \\
\poeml They will fear the \divine{Lord} our God. \\
\poemll    They will be terrified because of you.
\passage{Who is like God?}
\poeml \v{18}Is there any God like you, \\
\poemll    forgiving iniquity, \\
\poemlll       passing over transgressions by the survivors who are your\fnote{Lit. \fbib{his}} heritage?\fnote{I.e. God's people} \\
\poeml He is not angry forever, \\
\poemll    because he delights in gracious love. \\
\poeml \v{19}He will again show us compassion; \\
\poemll    he will subdue our iniquities. \\
\poeml You will hurl all their sins into the deepest sea. \\
\poeml \v{20}You will remain true to Jacob, \\
\poemll    and merciful to Abraham, \\
\poeml as you promised our ancestors long ago.\end{poetry}
