\bookheader{Zephaniah}
\labelbook{Zeph}

\bookpretitle{The Book of the Prophet}
\booktitle{Zephaniah}

\labelchapt{1}
\passage{The Imminent Destruction of Judah}

\chapt{1}
\v{1}This message from the \divine{Lord} came to Cushi's son Zephaniah,\fnote{\fbackref{1:1} The Heb. name \fbib{Zephaniah} means \fbib{The \divine{Lord} has treasured}} the grandson of Gedaliah and great-grandson of Hezekiah's son Amariah, during the reign\fnote{\fbackref{1:1} Lit. \fbib{in the days of}} of Amon's son Josiah, king of Judah:

\begin{poetry}
\poeml \v{2}``I'll utterly sweep away everything from\fnote{\fbackref{1:2} Lit. \fbib{from the face of}} the land,'' \\
\poemll    declares the \divine{Lord}. \\
\poeml \v{3}``I'll consume both human beings and animals--- \\
\poemll    I'll consume the birds of the sky, \\
\poeml the fish in the sea, \\
\poemll    and the wicked along with their sin,\fnote{\fbackref{1:3} Lit. \fbib{with the heap of rubble}} \\
\poeml when I eliminate human beings from the land,'' \\
\poemlll       declares the \divine{Lord}. \\
\poeml \v{4}``I will also stretch out my hand against Judah, \\
\poemll    and upon all inhabitants of Jerusalem. \\
\poeml I'll wipe out every trace of Baal from this place, \\
\poemll    and the name\fnote{\fbackref{1:4} Or \fbib{authority}} of the pagan priests\fnote{\fbackref{1:4} Lit. \fbib{the Chemarim}; i.e. idol worshiping priests}, \\
\poemlll       along with my own\fnote{\fbackref{1:4} The Heb. lacks \fbib{my own}} priests. \\
\poeml \v{5}I'll wipe out\fnote{\fbackref{1:5} The Heb. lacks \fbib{I'll wipe out}} those who worship the stars \\
\poemll    that they view\fnote{\fbackref{1:5} The Heb. lacks \fbib{that they view}} from their housetops, \\
\poeml those who bow down and swear to the \divine{Lord} \\
\poemll    and who also swear by Milcom,\fnote{\fbackref{1:5} I.e. the national idol of the Ammonites} \\
\poeml \v{6}those who turn away from the \divine{Lord}, \\
\poemll    don't seek the \divine{Lord}, \\
\poemlll       and never ask for his help.\fnote{\fbackref{1:6} Lit. \fbib{and don't inquire of him}}''
\passage{The Approaching Day of the \divine{Lord}}
\poeml \v{7}Remain silent in the presence of the Lord \divine{God}, \\
\poemll    because the Day of the \divine{Lord} approaches, \\
\poeml and because the \divine{Lord} has prepared a sacrifice \\
\poemll    for those whom he has invited to be consecrated. \\
\poeml \v{8}``It will come about during\fnote{\fbackref{1:8} Lit. \fbib{in the day of}} the \divine{Lord}'s sacrifice \\
\poemll    that I'll punish the officials,\fnote{\fbackref{1:8} Lit. \fbib{princes}} the royal descendants, \\
\poemlll       and all who wear foreign clothing.\fnote{\fbackref{1:8} Or \fbib{wear cultic vestments}} \\
\poeml \v{9}At the same time, I'll punish every idol worshipper,\fnote{\fbackref{1:9} Lit. \fbib{everyone who leaps over the threshold}; cf. 1Sam 5:5} \\
\poemll    especially those who are filling their master's temple with violence and deceit. \\
\poeml \v{10}When all of this happens,''\fnote{\fbackref{1:10} Lit. \fbib{It will come about in that day}} declares the \divine{Lord}, \\
\poemll    ``a loud shriek will come\fnote{\fbackref{1:10} The Heb. lacks \fbib{will come}} from the Fish Gate, \\
\poeml and howling from the Mishneh\fnote{\fbackref{1:10} Or \fbib{from The Second}; i.e. a section of Jerusalem} Quarter, \\
\poemll    along with great destruction from the hills.''
\passage{Divine Judgment on the Business Community}
\poeml \v{11}``Wail, you who live in the market district, \\
\poemll    because all of the merchants will be crushed \\
\poemlll       and all of their customers\fnote{\fbackref{1:11} Lit. \fbib{all who carry silver}} will be eliminated.\fnote{\fbackref{1:11} Lit. \fbib{be cut off}} \\
\poeml \v{12}And it will come about that I will search Jerusalem with candles,\fnote{\fbackref{1:12} Or \fbib{with oil lamps}} \\
\poemll    punishing the self-satisfied and complacent, \\
\poeml who say to themselves, \\
\poemll    `The \divine{Lord} will do neither good nor evil.' \\
\poeml \v{13}Therefore their possessions will be seized as plunder \\
\poemll    and their homes left in ruins. \\
\poeml They may build houses, \\
\poemll    but they won't live in them. \\
\poeml They may plant vineyards, \\
\poemll    but they won't drink their wine.''
\passage{Zephaniah's Description of the Day of the \divine{Lord}}
\poeml \v{14}``The great Day of the \divine{Lord} approaches--- \\
\poemll    How it comes, hurrying faster and faster! \\
\poeml The sound of the Day of the \divine{Lord} there \\
\poemll    includes the bitter cry of the mighty soldier. \\
\poeml \v{15}That day will be filled with wrath, \\
\poemll    a day of trouble and tribulation, \\
\poeml a day of desolation and devastation, \\
\poemll    a day of doom and gloom, \\
\poeml a day of clouds and shadows,\fnote{\fbackref{1:15} Cf. Joel 2:2a} \\
\poeml \v{16}a day of trumpet and battle cry \\
\poemlll       against fortified cites and watch\fnote{\fbackref{1:16} Lit. \fbib{corner}} towers. \\
\poeml \v{17}``And I'll bring so much distress to people \\
\poemll    that they will walk around like the blind. \\
\poeml Because they have sinned against the \divine{Lord}, \\
\poemll    their blood will be poured out like dust \\
\poemlll       and their intestines will spill out\fnote{\fbackref{1:17} The Heb. lacks \fbib{will spill out}} like manure. \\
\poeml \v{18}Neither their silver nor their gold will deliver them \\
\poemll    in the Day of the \divine{Lord}'s wrath; \\
\poeml but the entire land will be consumed \\
\poemll    by the fire of his jealousy, \\
\poeml for he will bring the inhabitants of the land to a sudden end.''
\end{poetry}
\labelchapt{2}
\passage{A Plea for Repentance to the People}

\begin{poetry}
\poeml \chapt{2}
\v{1}``Gather together! \\
\poeml Yes, indeed, gather together, \\
\poemlll       you shameless nation! \\
\poeml \v{2}Before the decree is carried out, \\
\poemll    before the day flies away like chaff, \\
\poeml before the fierce anger of the \divine{Lord} visits you, \\
\poemll    before the Day of the \divine{Lord}'s wrath surprises\fnote{\fbackref{2:2} Lit. \fbib{wrath comes upon}} you, \\
\poeml \v{3}seek the \divine{Lord}, all you humble people of the land, \\
\poemll    who do what he commands. \\
\poeml Seek righteousness! \\
\poemll    Seek humility! \\
\poemlll       Maybe you will be protected in the Day of the \divine{Lord}'s anger.''
\passage{The Coming Destruction of Philistine Cities}
\poeml \v{4}``For Gaza will be forsaken,\fnote{\fbackref{2:4} MT word \fbib{forsaken} sounds like The Heb. name \fbib{Gaza.}} \\
\poemll    and Ashkelon deserted--- \\
\poeml Ashdod will be emptied at high noon; \\
\poemll    even Ekron will be uprooted.\fnote{\fbackref{2:4} MT word \fbib{uprooted} sounds like The Heb. name \fbib{Ekron}.} \\
\poeml \v{5}Woe to those who live along the coast, \\
\poemll    the people of Philistia!\fnote{\fbackref{2:5} Lit. \fbib{of the Cherethites}} \\
\poeml This message from the \divine{Lord} is being spoken against you, \\
\poemll    Canaan, land of the Philistines: \\
\poemlll       `I'll destroy you until no one lives there!' \\
\poeml \v{6}The Philistine\fnote{\fbackref{2:6} Lit. \fbib{Cherethite}} coast will become meadows \\
\poemll    for shepherds and sheep pens. \\
\poeml \v{7}The survivors of Judah will find pasture on it; \\
\poemll    at twilight they will lie down in the houses of Ashkelon, \\
\poeml for the \divine{Lord} their God will visit them, \\
\poemll    restoring their prosperity.''\fnote{\fbackref{2:7} Or \fbib{them, turning away their captivity}}
\passage{The Lord's Rebuke to Moab and Ammon}
\poeml \v{8}``I've heard Moab's insults \\
\poemll    and the curses from those Ammonites\fnote{\fbackref{2:8} Lit. \fbib{the descendants of Ammon}; so also in v. 9.} \\
\poeml by which they defame my people \\
\poemll    and boast about their territory. \\
\poeml \v{9}Therefore as I'm alive and living,'' declares the \divine{Lord} of the Heavenly Armies, the God of Israel, \\
\poemll    ``Moab will surely become like Sodom, \\
\poemlll       and the Ammonites\fnote{\fbackref{2:9} Lit. \fbib{the descendants of Ammon}; so also in v. 9.} like Gomorrah, \\
\poeml a place overrun by weeds and salty marshes, \\
\poemll    unpopulated forever. \\
\poeml The survivors of my people will confiscate their property, \\
\poemll    and those who remain of my nation will inherit what was theirs.\fnote{\fbackref{2:9} Cf. Obad 19-21} \\
\poeml \v{10}This they\fnote{\fbackref{2:10} I.e. the people of Edom and Moab} will have in lieu of their pride, \\
\poemll    because they have insulted and mocked the people of the \divine{Lord} of the Heavenly Armies. \\
\poeml \v{11}The \divine{Lord} will incite them to terror, \\
\poemll    because he will cause all the gods of the earth to waste away. \\
\poeml They will worship him, \\
\poemll    every person in his own home, \\
\poemlll       including even the coastlands of the nations.''
\passage{The \divine{Lord}'s Rebuke to Cush and Nineveh}
\poeml \v{12}``Now as for you, Cush,\fnote{\fbackref{2:12} I.e. Sudan, Ethiopia, or southern Iraq} \\
\poemll    you\fnote{\fbackref{2:12} Lit. \fbib{they}} will surely be slain by my sword! \\
\poeml \v{13}``And the \divine{Lord}\fnote{\fbackref{2:13} Lit. \fbib{And he}} will attack\fnote{\fbackref{2:13} Lit. \fbib{stretch out his hand toward}} the north, destroying Assyria. \\
\poemll    He will turn Nineveh into a desolate ruin, \\
\poemlll       as dry as a desert wilderness. \\
\poeml \v{14}Flocks will lie down in her midst, \\
\poemll    along with animals of every kind. \\
\poeml Desert owls and screeching owls will nest at the top of the pillars, \\
\poemll    hooting through the vacant\fnote{\fbackref{2:14} The Heb. lacks \fbib{vacant}} windows, \\
\poemlll       `Ruin sits at these doorsills,' \\
\poeml for he will expose even the cedar framework. \\
\poeml \v{15}This is that carefree city that lived irresponsibly, \\
\poemll    that told herself, `Me first!'\fnote{\fbackref{2:15} Lit. \fbib{said in her heart, `I'}} \\
\poemlll       and, `There will be no one else!' \\
\poeml How ruined she has become--- \\
\poemll    a habitat for wild animals! \\
\poeml Everyone who passes by her will sneer at her \\
\poemll    and make obscene gestures!''\fnote{\fbackref{2:15} Lit. \fbib{and wag his hand}}
\end{poetry}
\labelchapt{3}
\passage{The \divine{Lord}'s Rebuke to Jerusalem}

\begin{poetry}
\poeml \chapt{3}
\v{1}Woe to this filthy, polluted, and oppressive city! \\
\poeml \v{2}It won't obey anyone.\fnote{\fbackref{3:2} Lit. \fbib{obey the voice}} \\
\poeml It won't accept discipline. \\
\poemll    It does not trust in the \divine{Lord}. \\
\poemlll       It does not approach God. \\
\poeml \v{3}Its national officials\fnote{\fbackref{3:3} Lit. \fbib{Her princes}} are roaring lions; \\
\poemll    its judges are like wolves of the night \\
\poemlll       that don't leave the bones for the morning. \\
\poeml \v{4}Its prophets are arrogant and treacherous. \\
\poemll    Its priests have contaminated the sanctuary. \\
\poemlll       They give perverse interpretations of the Law.\fnote{\fbackref{3:4} Lit. \fbib{They do violence to the Law}} \\
\poeml \v{5}The righteous \divine{Lord} who lives\fnote{\fbackref{3:5} The Heb. lacks \fbib{within her}} within her will do no wrong; \\
\poemll    he will bring justice to light morning by morning. \\
\poeml He never fails, \\
\poemll    but the unjust are shameless.
\passage{The Integrity of God's Justice}
\poeml \v{6}``I have destroyed\fnote{\fbackref{3:6} Lit. \fbib{have cut off}} nations--- \\
\poemll    their fortifications are deserted. \\
\poeml I have turned their main thoroughfares into wastelands \\
\poemll    where no one will travel. \\
\poeml Their cities are desolate; \\
\poemll    as a result, not one man remains--- \\
\poemlll       no, not even a single resident. \\
\poeml \v{7}I have said, `If only you would fear me, \\
\poemll    if only you would take my instructions to heart.' \\
\poemlll       Then their houses would not have been torn down. \\
\poeml I have chastened them, \\
\poemll    but they were eager to corrupt everything they were doing.''
\passage{The Future Deliverance of Israel}
\poeml \v{8}``Just you wait!'' declares the \divine{Lord}. \\
\poemll    ``The day will come when I stand up as a prosecutor,\fnote{\fbackref{3:8} Or \fbib{witness}} \\
\poeml for I am determined to assemble the nations \\
\poemll    and to gather the kingdoms, \\
\poeml in order to pour out my indignation upon them--- \\
\poemll    all of my fierce anger. \\
\poeml All the earth will be consumed by the fire of my jealousy. \\
\poeml \v{9}Indeed, then I will return my people to a pure language\fnote{\fbackref{3:9} Or \fbib{to pure lips}} \\
\poemll    so that they all may call upon the name of the \divine{Lord}, \\
\poemlll       serving him with a united will.\fnote{\fbackref{3:9} Lit. \fbib{shoulder}} \\
\poeml \v{10}``From beyond the rivers of Sudan\fnote{\fbackref{3:10} Or Ethiopia; Lit. \fbib{Cush}} my worshipers\fnote{\fbackref{3:10} Lit. \fbib{incense burners}}--- \\
\poemll    including my dispersed people--- \\
\poemlll       will present offerings to me. \\
\poeml \v{11}When this happens,\fnote{\fbackref{3:11} Lit. \fbib{On that day}} you will not be ashamed of your actions \\
\poemll    by which you sinned against me, \\
\poeml because I will remove from among you those who revel in pride. \\
\poemll    Arrogance will have no place in my holy mountain. \\
\poeml \v{12}I will keep a humble and gentle people among you, \\
\poemll    and they will trust in the name of the \divine{Lord}. \\
\poeml \v{13}The survivors of Israel will not practice lawlessness \\
\poemll    nor tell lies, \\
\poeml nor will a deceitful message be found in their mouths, \\
\poemll    because they will eat and rest, \\
\poemlll       with no one to cause fear.''
\passage{Israel's Future Joys}
\poeml \v{14}``Sing aloud, daughter of Zion! \\
\poemll    Shout out, Israel! \\
\poemlll       Rejoice with all of your heart, daughter of Jerusalem! \\
\poeml \v{15}The \divine{Lord} has acquitted you;\fnote{\fbackref{3:15} Lit. \fbib{has withdrawn your judgments}} \\
\poemll    turning back your adversaries. \\
\poeml Israel's king, the \divine{Lord}, is among you; \\
\poemll    you will not fear disaster anymore. \\
\poeml \v{16}``When all of this happens,\fnote{\fbackref{3:16} Lit. \fbib{In that day}} it will be told Jerusalem, \\
\poemll    `Don't be afraid!'' \\
\poeml and to Zion, \\
\poemll    `Don't lose courage!''\fnote{\fbackref{3:16} Lit. \fbib{Don't let your hands drop}} \\
\poeml \v{17}The \divine{Lord} your God among you is powerful--- \\
\poemll    he will save \\
\poeml and he will take joyful delight in you. \\
\poemll    In his love he will renew you\fnote{\fbackref{3:17} The Heb. lacks \fbib{you}} with his love; \\
\poemlll       he will celebrate with singing because of you. \\
\poeml \v{18}I will gather the afflicted from the solemn assembly; \\
\poemll    those who were with you, \\
\poemlll       who were bearing a burden of disgrace. \\
\poeml \v{19}``Watch how I deal with everyone who oppresses you! \\
\poemll    At that time I will rescue the one who is lame, \\
\poemlll       and I will draw to me\fnote{\fbackref{3:19} The Heb. lacks \fbib{to me}} the one who has been driven away. \\
\poeml I will honor\fnote{\fbackref{3:19} Lit. \fbib{give}} them with praise \\
\poemll    and with a good reputation in every land \\
\poemlll       where they have been put to shame. \\
\poeml \v{20}At that time I will gather you; \\
\poemll    at that time I will bring you home.\fnote{\fbackref{3:20} The Heb. lacks \fbib{home}} \\
\poeml Indeed, I will give you a good reputation, \\
\poemll    making you praiseworthy among all of the people of the world, \\
\poemlll       when I restore your prosperity before your eyes,'' says the \divine{Lord}.\end{poetry}
