\bookheader{Zechariah}
\labelbook{Zech}

\bookpretitle{The Book of the Prophet}
\booktitle{Zechariah}

\labelchapt{1}
\passage{A Call to Return}

\chapt{1}
\v{1}In the eighth month of the second year\fnote{I.e. sometime from mid-October to mid-November 520 BC} of the reign of\fnote{The Heb. lacks \fbib{the reign of}} Darius, this message from the \divine{Lord} came to Berechiah's son Zechariah,\fnote{The Heb. name \fbib{Zechariah} means \fbib{the \divine{Lord} remembers}} the grandson of Iddo the prophet: \v{2}``The \divine{Lord} was very angry with your ancestors. \v{3}So tell them, `This is what the \divine{Lord} of the Heavenly Armies says: ``Return to me,'' declares the \divine{Lord} of the Heavenly Armies, ``and I will return to you.\fnote{Lit. \fbib{you,'' declares the \divine{Lord} of the Heavenly Armies}} \v{4}Don't be like your ancestors, to whom the former prophets proclaimed: `This is what the \divine{Lord} of the Heavenly Armies says: ``It's time to turn from your evil lifestyles\fnote{Lit. \fbib{paths}} and from your evil actions,'' `but they would neither listen nor pay attention to me,'\,'' declares the \divine{Lord}.' \v{5}``Your ancestors---where are they? And the prophets---do they live forever? \v{6}But my words and my statutes that I gave as commands to my servants the prophets---did they not overwhelm your ancestors? And they returned to me:\fnote{The Heb. lacks \fbib{to me}} `The \divine{Lord} of the Heavenly Armies acted toward us just as he planned to do---in keeping with our lifestyles\fnote{Lit. \fbib{paths}} and in keeping with our actions.'\,''
\passage{The Vision of Horses}

\v{7}On the twenty-fourth day of the eleventh month (the month Shebat) in the second year of the reign of\fnote{The Heb. lacks \fbib{the reign of}} Darius, this message from the \divine{Lord} came to Berechiah's son Zechariah, the grandson of Iddo the prophet: \v{8}``I stared into the night, and there was a man mounted on a red horse! The horse\fnote{Lit. \fbib{He}} was standing among the myrtle trees in a ravine. Behind him there were red, brown,\fnote{Or \fbib{sorrel}} and white horses.''

\v{9}Then I asked, ``Who are these, sir?''\fnote{Lit. \fbib{My lord}}

The messenger who was talking to me answered, ``I will tell you who these are.''

\v{10}The man who stood among the myrtle trees answered, ``These are the ones whom the \divine{Lord} sent out to wander throughout the earth.''

\v{11}Then they reported to the angel of the \divine{Lord} who stood among the myrtle trees, ``We have wandered throughout the earth---and look!---the entire earth is at rest. Everything is quiet and peaceful.''\fnote{Lit. \fbib{is rest and remains}}

\v{12}And the angel of the \divine{Lord} replied, ``\divine{Lord} of the Heavenly Armies, how long will it be until you show mercy to Jerusalem and to the cities of Judah, with whom you have been angry for these past seventy years?''

\v{13}So the \divine{Lord} answered the angel who was speaking to me with kind and comforting words.
\passage{The \divine{Lord}'s Concern for Zion}

\v{14}Then the angel who was speaking to me told me, ``Announce this: `This is what the \divine{Lord} of the Heavenly Armies says: ``I have a deep concern for Jerusalem, a great concern for Zion. \v{15}I am deeply angry with the nations who are complacent, with whom I was only a little displeased---but they made things worse!'' \v{16}`Therefore this is what the \divine{Lord} says: ``I have returned to Jerusalem with compassionate intentions. My Temple will be rebuilt there,'' declares the \divine{Lord} of the Heavenly Armies, ``and the measuring line will be stretched out over Jerusalem.''\,'\,''
\passage{The Future Prosperity of Zion}

\v{17}``Also announce the following: `This is what the \divine{Lord} of the Heavenly Armies says: ``My cities will again overflow with prosperity. The \divine{Lord} will comfort Zion once more and will choose Jerusalem again.''\,'\,''
\passage{The Vision of Four Horns}

\v{18}\fnote{This v. is 2:1 in MT, 1:19 is 2:2 in MT, and so through v. 21}Then I looked up and saw four horns. \v{19}I asked the angel who was talking to me, ``What are those?''

So he answered me, ``Those are the forces\fnote{Lit. \fbib{horns}} that have dispersed Judah, Israel, and Jerusalem.''
\passage{The Vision of Four Artisans}

\v{20}Then the \divine{Lord} showed me four artisans.

\v{21}Then I asked, ``What have they come to do?''

He answered, ``Those horns that dispersed Judah---so that no one could lift up his head---those artisans\fnote{The Heb. lacks \fbib{artisans}} are coming to disrupt the power\fnote{Lit. \fbib{to terrify the horns}} of nations, tearing them down now that they've come to power and dispersed the land of Judah.''
\labelchapt{2}
\passage{The Vision of the Measuring Line}

\chapt{2}
\v{1}\fnote{This v. is 2:5 in MT, and so throughout the chapter}Then I looked up and saw a man with a measuring line in his hand. \v{2}I asked, ``Where are you going?''

He responded, ``To measure Jerusalem in order to determine its width and length.''

\v{3}Look! That angel who was talking to me left, and another angel came forward to meet him. \v{4}That other angel told him, ``Run and tell that young man: `Jerusalem will be an inhabited city without walls due to the number of people and livestock within it. \v{5}I myself will be an encircling rampart of fire,' declares the \divine{Lord}, `and I will be the glory in her midst.'

\v{6}```Come now! Come now! Flee from the land of the north,' declares the \divine{Lord}, `for I have dispersed you like the four winds of heaven,' declares the \divine{Lord}.

\v{7}```Come now, Zion! Escape, you who are living with the residents of Babylon. \v{8}For this is what the \divine{Lord} of the Heavenly Armies says: ``In pursuit of glory I was sent to the nations who plundered you, because whoever injures you injures the pupil of my eye. \v{9}And look, I will shake my fist over them, and they will become plunder for their former\fnote{The Heb. lacks \fbib{former}} slaves. And you will know that the \divine{Lord} of the Heavenly Armies sent me.''\,'\,''
\passage{The \divine{Lord} will Live in Zion}

\v{10}``Sing and rejoice, daughter of Zion! Take note! I am coming to live in your midst,'' declares the \divine{Lord}. \v{11}``Many nations will cling to the \divine{Lord} at that time\fnote{Lit. \fbib{day}} and will become my people. I will live in your midst, and you will know that the \divine{Lord} of the Heavenly Armies has sent me to you. \v{12}The \divine{Lord} will take possession of Judah as his own property in the Holy Land---and he will choose Jerusalem again. \v{13}Be silent, every living thing, in the presence of the \divine{Lord}, because he is emerging\fnote{Lit. \fbib{awakened}} from his Holy Place.''
\labelchapt{3}
\passage{The Vision of the High Priest}

\chapt{3}
\v{1}Then I saw Joshua the High Priest standing in the presence of the angel of the \divine{Lord}, with Satan\fnote{The Heb. name \fbib{Satan} means \fbib{The Opponent} or \fbib{The Accuser}} standing at his right to oppose him.

\v{2}The \divine{Lord} told Satan, ``The \divine{Lord} rebuke you, Satan---in fact, may the \divine{Lord} who has chosen Jerusalem rebuke you! This man is a burning brand plucked from the fire, is he not?''

\v{3}Now Joshua was wearing filthy clothes as he stood in the presence of the angel.

\v{4}So the angel\fnote{Lit. \fbib{So he}} continued to tell those who were standing in his presence, ``Remove his filthy clothes.''

And he told Joshua,\fnote{Lit. \fbib{him}} ``Look how I've removed your iniquity. Now I'm clothing you with fine garments.''

\v{5}Then I said, ``Let them place a pure diadem\fnote{Or \fbib{clean turban}} on his head.''

So they placed the pure diadem\fnote{Or \fbib{clean turban}} on his head and clothed him with fine garments while the angel of the \divine{Lord} was standing beside them.\fnote{The Heb. lacks \fbib{beside them}}
\passage{The \divine{Lord}'s Charge to Joshua}

\v{6}Then the angel of the \divine{Lord} gave this charge to Joshua: \v{7}``This is what the \divine{Lord} of the Heavenly Armies says: `If you will live according to my ways, and if you will keep what I have entrusted to you,\fnote{The Heb. lacks \fbib{to you}} then you will also administer my Temple, and you will also guard my courtyards. And I will give you access to these who serve\fnote{Lit. \fbib{stand}} here.

\v{8}```Listen, High Priest Joshua, you and those companions of yours who sit with you,\fnote{Lit. \fbib{sit in your presence}} for these men are a sign that I am presenting my servant, the Branch.\fnote{Cf. Ps 132:17; Jer 23:5; 33:15} \v{9}Look, the stone that I put in place in Joshua's presence---on that one stone are seven eyes.\fnote{Or \fbib{facets}} And look, I will do the engraving myself,' declares the \divine{Lord} of the Heavenly Armies, `and I will remove the perversity of that land in a single day. \v{10}At that time,'\fnote{Lit. \fbib{day}} declares the \divine{Lord} of the Heavenly Armies, `you will invite each of your neighbors to join you\fnote{The Heb. lacks \fbib{to join you}} under the vine and fig tree.'\,''
\labelchapt{4}
\passage{The Vision of the Golden Menorah}

\chapt{4}
\v{1}Then the angel who had been speaking with me returned and woke me up as if I had been asleep. \v{2}Then he asked me, ``What do you see?''

So I said, ``I have been watching---and look!---there is a menorah made completely of gold with a bowl on top of it. And there are seven lights on it, along with seven feeder channels to the lamps, which are also on top of it. \v{3}Two olive trees are near it, one on the right side of the bowl and one on the left.''

\v{4}Then I asked the angel who had been speaking with me, ``Sir,\fnote{Lit. \fbib{My lord}} what are these?''

\v{5}The angel who had been speaking with me answered by asking, ``You don't know what these mean, do you?''

So I responded, ``No, sir.''
\passage{The \divine{Lord}'s First Charge to Zerubbabel}

\v{6}Then he replied to me, ``This is this message from the \divine{Lord} to Zerubbabel: `Not by valor nor by strength, but only by my Spirit,' says the \divine{Lord} of the Heavenly Armies. \v{7}`Who are you, great mountain? You will become a plain in Zerubbabel's presence, and he will position the capstone, exulting over it, ``How beautiful! How beautiful!''\,'\,''
\passage{The \divine{Lord}'s Second Charge to Zerubbabel}

\v{8}Then this message from the \divine{Lord} came to me again: \v{9}``Zerubbabel's hands have laid the foundation of this Temple, and his hands will finish it, so that you will know that the \divine{Lord} of the Heavenly Armies has sent me to all of you. \v{10}For who has despised the time\fnote{Lit. \fbib{day}} of insignificant things? They will rejoice to see the plumb line in the hand of Zerubbabel.\fnote{I.e. building the temple} These seven lights\fnote{The Heb. lacks \fbib{lights}} represent the eyes of the \divine{Lord}, which are looking throughout all of the earth.''
\passage{The Two Anointed Ones}

\v{11}Then I asked the angel,\fnote{Lit. \fbib{him}} ``What are these two olive trees, one\fnote{The Heb. lacks \fbib{one}} on the right of the menorah and one\fnote{The Heb. lacks \fbib{one}} on the left?'' \v{12}I also asked him a second question: ``What are these two olive branches on either side of\fnote{Lit. \fbib{that are at the hand of}} the two golden feeder channels that carry the golden oil to the seven lights?''\fnote{The Heb. lacks \fbib{oil to the seven lights}}

\v{13}The angel\fnote{Lit. \fbib{He}} replied, ``You don't know what these are, do you?''

I responded to him, ``No, sir.''

\v{14}He said, ``These are the two anointed ones,\fnote{Lit. \fbib{sons of fresh oil}} who stand continuously beside the Lord of the whole earth.''
\labelchapt{5}
\passage{The Vision of the Flying Scroll}

\chapt{5}
\v{1}Then I looked up and saw a flying scroll! \v{2}And the angel\fnote{Lit. \fbib{he}} asked me, ``What do you see?''

I answered him, ``I'm looking at a flying scroll. It's 20 cubits\fnote{I.e. about 30 feet; a cubit was about eighteen inches} long and ten cubits\fnote{I.e. about fifteen feet; a cubit was about eighteen inches} wide.''

\v{3}He responded to me, ``This is the curse that is going out over the surface of the whole earth, because, according to this side of the scroll,\fnote{The Heb. lacks \fbib{side of the scroll}} all thieves will be ejected, and according to the other side of the scroll,\fnote{The Heb. lacks \fbib{side of the scroll}} all liars will be ejected.''

\v{4}``I am bringing this about,'' declares the \divine{Lord} of the Heavenly Armies. ``The curse\fnote{Lit. \fbib{it}} will enter the house of the thief and the house of the one who lies in my name. The curse\fnote{Lit. \fbib{It}} will remain in his house until that house\fnote{Lit. \fbib{it}} has been completely destroyed, right down to its timber and stones.''
\passage{The Vision of the Basket}

\v{5}Then the angel who had been talking with me stepped forward and told me, ``Please look up and see what's going out.''

\v{6}So I asked, ``What is it?''\fnote{Lit. \fbib{is she}}

He replied, ``This is a basket\fnote{Lit. \fbib{ephah}; i.e. a measure of dry grain about 20 dry quarts in capacity; and so throughout the chapter} making its appearance.'' He also said, ``This is what it appears to be in\fnote{Lit. \fbib{is the eyes in}} the entire land.''
\passage{The Vision of the Woman in the Basket}

\v{7}Look, a round lead cover was being lifted, and there was one woman seated inside the basket! \v{8}And the angel\fnote{Lit. \fbib{he}} said, ``This is evil!'' So he shoved her back into the basket and snapped the round, lead cover over the opening.
\passage{The Vision of the Two Winged Women}

\v{9}Then I looked up to see two women coming forward with the wind filling their wings. (They had wings like those of a stork.) They took up the basket, holding it between the earth and sky.

\v{10}So I asked the angel who was talking to me, ``Where are they taking the basket?''

\v{11}He answered me, ``To the land of Shinar,\fnote{I.e. the Babylon area} so they can build a temple to the woman in the basket.\fnote{Lit. \fbib{to her}} Then when its preparations are complete, the basket\fnote{Lit. \fbib{complete, it}} will be set there in its place.''
\labelchapt{6}
\passage{The Vision of the Four Chariots}

\chapt{6}
\v{1}Then I looked up and saw four chariots coming out from between two mountains! And the mountains were made of brass! \v{2}Attached to the first chariot were red horses. Attached to the second chariot were black horses. \v{3}Attached to the third chariot were white horses. Attached to the fourth chariot were speckled horses and gray\fnote{Or strong} horses. \v{4}In response, I asked the angel who had been talking with me, ``Sir, what are these?''

\v{5}The angel told me, ``These are four heavenly spirits that are going out on patrol after having presented themselves to the Lord of the whole earth. \v{6}The black horses are headed into the north country, and the white ones are headed out after them. The speckled horses are headed toward the south country.''

\v{7}Then the gray horses went out. They were eager to go out on patrol throughout the earth, so the angel\fnote{Lit. \fbib{he}} said, ``Go patrol the earth.'' So they went out on patrol throughout the earth.

\v{8}Then he called to me, ``Look! The horses that went north have caused my spirit to rest in the north country.''
\passage{The Future Prosperity of the Branch}

\v{9}Then this message from the \divine{Lord} came to me: \v{10}``Go take up an offering\fnote{The Heb. lacks \fbib{an offering}} from those who came out of\fnote{The Heb. lacks \fbib{those who came out of}} the Babylonian\fnote{The Heb. lacks \fbib{Babylonian}} exile, that is, from Heldai, from Tobijah, and from Jedaiah. Go along with them today into the house of Zephaniah's son Josiah, who returned from Babylon. \v{11}Take silver and gold and fashion crowns to set upon the head of Joshua son of Johozadak, the High Priest. \v{12}Then tell him, `This is what the \divine{Lord} of the Heavenly Armies says: ``Here is the man whose name is The Branch.\fnote{i.e. Zerubbabel; cf. Zech 3:8} He will branch out from where he is and will rebuild the Temple of the \divine{Lord}. \v{13}Yes, he will indeed rebuild the Temple of the \divine{Lord}, and he will exalt its majesty by sitting and ruling on his throne. He will serve as priest on his throne, and no contention\fnote{Lit. \fbib{and peaceful counsel}} will exist between them. \v{14}The crowns will go to Helem, to Tobijah, to Jedaiah, and to Zephaniah's son Hen, as a memorial in the Temple of the \divine{Lord}. \v{15}Those who are now far away will come and do reconstruction work in the Temple of the \divine{Lord}. Then you will know that the \divine{Lord} of the Heavenly Armies has sent me to you. This will come about if you diligently obey the voice of the \divine{Lord} your God.''\,'\,''
\labelchapt{7}
\passage{A Rebuke about Selfish Fasts}

\chapt{7}
\v{1}During the fourth year of the reign of\fnote{The Heb. lacks \fbib{the reign of}} King Darius, a message from the \divine{Lord} came to Zechariah on the fourth day of the ninth month Kislev.\fnote{I.e. c. 7 December 518 BC} \v{2}The people of\fnote{The Heb. lacks \fbib{The people of}} Bethel were sending\fnote{Lit. \fbib{was sending}} Sharezer, Regem-melech, and their men to pray in the \divine{Lord}'s presence \v{3}and to speak to the priests assigned\fnote{The Heb. lacks \fbib{assigned}} to the Temple of the \divine{Lord} of the Heavenly Armies along with the prophets, asking, ``Am I to go about mourning, denying myself throughout the fifth month,\fnote{I.e. the anniversary month of the Jerusalem temple's destruction} as I have these many years?''

\v{4}Then this message from the \divine{Lord} of the Heavenly Armies came to me: \v{5}``Talk to everyone in the land, as well as to the priests. Ask them, `When you were fasting and mourning during the fifth and seventh months\fnote{The Heb. lacks \fbib{months}} for the past seventy years, were you really fasting for me? \v{6}And when you eat and drink, you're eating and drinking for your own benefit, aren't you? \v{7}Isn't this what the \divine{Lord} proclaimed through the former prophets, when a prosperous Jerusalem was inhabited, as were its surrounding cities, the Negev,\fnote{I.e. the southern regions of the Sinai peninsula; cf. Josh 10:40} and the Shephelah?'\,''\fnote{I.e. the verdant central lowlands of Israel; cf. Josh 10:40}
\passage{The Consequence of Turning from God}

\v{8}This message from the \divine{Lord} came to Zechariah again: \v{9}``This is what the \divine{Lord} of the Heavenly Armies says: `Administer true justice, and show gracious love and mercy toward each other.\fnote{Lit. \fbib{toward his brother}} \v{10}You are not to wrong the widow, orphans, the foreigner, or the poor, and you are not to plan evil against each other.\fnote{Lit. \fbib{toward his brother}} \v{11}But they refused to pay attention, turned their backs, and stopped listening. \v{12}They made their hearts hard like a diamond, to keep from obeying the Law and the messages that the \divine{Lord} of the Heavenly Armies sent by his Spirit through the former prophets. \v{13}Therefore, just as when I\fnote{Lit. \fbib{he}} cried out and they would not listen, so also they will cry out, and I will not listen,' says the \divine{Lord} of the Heavenly Armies. \v{14}`I will scatter them to all of the nations, which they have not known.'\,''

Now the earth was left desolate after them. As a result, no one came or went because they had turned a pleasant land into a desert.
\labelchapt{8}
\passage{What the \divine{Lord} will Do for Zion}

\chapt{8}
\v{1}This is this message from the \divine{Lord} of the Heavenly Armies:

\v{2}``This is what the \divine{Lord} of the Heavenly Armies says: `I'm greatly jealous about Zion, and that makes me furious about her.'

\v{3}``This is what the \divine{Lord} says: `I will return to Zion and I will live in the midst of Jerusalem. And Jerusalem will be called, ``The City of Truth'' and the mountain of the \divine{Lord} of the Heavenly Armies will be called,\fnote{The Heb. lacks \fbib{will be called}} ``The Holy Mountain''.'

\v{4}``This is what the \divine{Lord} of the Heavenly Armies says: `There will yet be old men and old women sitting in the parks\fnote{Or \fbib{streets}} of Jerusalem, each one of them holding canes in their hands due to their old age! \v{5}The city parks\fnote{Or \fbib{streets}} will be filled with boys and girls. They will play in the city's\fnote{Lit. \fbib{in its}} open parks.'

\v{6}``This is what the \divine{Lord} of the Heavenly Armies says: `It may seem impossible to the survivors of this people, but is it impossible for me?' declares the \divine{Lord} of the Heavenly Armies.

\v{7}``This is what the \divine{Lord} of the Heavenly Armies says: `Look! After having saved my people from the land of the east and from the land of the west, \v{8}I will also bring them back to live in the midst of Jerusalem. They will be my people and I will truly be their righteous God.'\,''
\passage{An Exhortation to Rebuild the Temple}

\v{9}``This is what the \divine{Lord} of the Heavenly Armies says: `Be strong so the Temple can be built, you who are now listening to this message spoken by the prophets when the foundation was laid to the Temple of the \divine{Lord} of the Heavenly Armies. \v{10}Before then, everyone was unemployed---even the draft animals---and no one was safe coming or going because of the enemy, because I caused everyone to oppose each other.\fnote{Lit. \fbib{I set everyone against his neighbor}}

\v{11}```But now I will not treat the survivors of this people as I did formerly,' declares the \divine{Lord} of the Heavenly Armies. \v{12}`For there will be a sowing of peace: the vine will produce its fruit, the earth will produce its full\fnote{The Heb. lacks \fbib{full}} yield, the sky will produce its dew, and I will make the survivors of this people inherit all these things. \v{13}Furthermore, house of Judah and house of Israel, even though you used to be a curse among the nations, now I will surely save you, and you will be a blessing. Stop being afraid. Instead, grow stronger.'\,''\fnote{Lit. \fbib{your hands become strong.}}
\passage{The \divine{Lord} will Do Good to Jerusalem and Judah}

\v{14}``This is what the \divine{Lord} of the Heavenly Armies says: `Just as I intended to bring harm to you when your ancestors angered me,' says the \divine{Lord} of the Heavenly Armies, `and I did not relent, \v{15}so I have decided at this time\fnote{Lit. \fbib{at these days}} to do good things for Jerusalem and for the house of Judah. So stop being afraid. \v{16}Here's what you must do: tell the truth to your neighbors, administer true and peaceful justice in your courtrooms,\fnote{Lit. \fbib{gates}} \v{17}don't plot evil in your heart against a neighbor, and don't tolerate\fnote{Lit. \fbib{and love no}} false testimonies,\fnote{Or \fbib{oaths}} because I hate all these things,' declares the \divine{Lord}.''

\v{18}Once again this message from the \divine{Lord} come to me: \v{19}``This is what the \divine{Lord} of the Heavenly Armies says: `The fasts that occur in the fourth, fifth, seventh, and tenth months will be joyful and glad times for the house of Judah, replete with\fnote{The Heb. lacks \fbib{replete with}} cheerful festivals. Therefore, love truth and peace.'\,''
\passage{Future Visits by Nations to Jerusalem}

\v{20}``This is what the \divine{Lord} of the Heavenly Armies says: `In the future, people will come, including residents of many cities, \v{21}and they will travel from one place to another place and say, ``Let's go quickly to pray in the presence of the \divine{Lord} and to inquire of the \divine{Lord} of the Heavenly Armies.' And I will go, too. \v{22}Many people and powerful nations will come to inquire of the \divine{Lord} of the Heavenly Armies in Jerusalem, and to pray in the presence of the \divine{Lord}.''\,'\,''

\v{23}``This is what the \divine{Lord} of the Heavenly Armies says: `In the future, ten men speaking\fnote{The Heb. lacks \fbib{speaking}} all the languages of the nations will grab hold of one Jewish person by the hem of his garment and say, ``Let us go up to Jerusalem\fnote{The Heb. lacks \fbib{to Jerusalem}} with you, because we heard that God is with you.''\,'\,''
\labelchapt{9}
\passage{Judgment on Israel's Enemies}

\begin{poetry}
\poeml \chapt{9}
\v{1}A declaration: this message from the \divine{Lord} in the land of Hadrach,\fnote{I.e. a district near Damascus} \\
\poemll    with Damascus its confederate,\fnote{Cf. Isa 7:2; or \fbib{resting place}} \\
\poeml because the eyes of humanity\fnote{Or \fbib{Aram}} will look to\fnote{Lit. \fbib{will be on}} the \divine{Lord}, \\
\poemll    along with those of\fnote{The Heb. lacks \fbib{those of}} all the tribes of Israel. \\
\poeml \v{2}Also Hamath, which borders on it--- \\
\poemll    along with Tyre and Sidon--- \\
\poemlll       indeed they are very wise. \\
\poeml \v{3}``Tyre built itself a fortification, \\
\poemll    hoarding silver like dust \\
\poemlll       and pure gold like mud in a street. \\
\poeml \v{4}Look! The Lord will evict her, \\
\poemll    striking at her power in the sea, \\
\poemlll       and she will be consumed by fire. \\
\poeml \v{5}Ashkelon will see it happen and will be terrified; \\
\poemll    Gaza will tremble greatly. \\
\poeml Ekron will be ashamed of her expectations, \\
\poemll    Gaza's king will perish, \\
\poemlll       and Ashkelon will become uninhabited. \\
\poeml \v{6}A strange people\fnote{Lit. \fbib{A bastard}} will inhabit Ashdod, \\
\poemll    and I will eliminate the arrogance of Philistia. \\
\poeml \v{7}I will remove the blood from its mouth, \\
\poemll    along with its abominations from between its teeth. \\
\poeml Its survivors will be dedicated to our God; \\
\poemll    It will be like a clan of Judah, \\
\poemlll       and Ekron will be as a Jebusite.\fnote{I.e. a Jerusalemite} \\
\poeml \v{8}I will set a garrison around my Temple, \\
\poemll    to hinder those who might come and go, \\
\poeml and to guard against oppressors who intend to invade; \\
\poemll    for I have taken note of this with my eyes.''
\passage{The Coming of Zion's King}
\poeml \v{9}``Rejoice greatly, daughter of Zion; \\
\poemll    cry out, daughter of Jerusalem! \\
\poeml Look! Your king is coming to you. \\
\poemll    He is righteous, \\
\poemlll       and he is able to save. \\
\poeml He is humble, \\
\poemll    and is riding on a colt, \\
\poemlll       the foal of a donkey. \\
\poeml \v{10}I will banish\fnote{Lit. \fbib{will cut off}} chariots from Ephraim \\
\poemll    and horses from Jerusalem. \\
\poeml War weapons\fnote{Lit. \fbib{The war bow}} will be banished, \\
\poemll    and your king\fnote{Lit. \fbib{and he}} will speak peace to the nations. \\
\poeml His dominion will extend from sea to sea, \\
\poemll    and from the River to the farthest portion of the earth. \\
\poeml \v{11}Now concerning you and my blood covenant with you, \\
\poemll    I have liberated your prisoners \\
\poemlll       from a waterless pit. \\
\poeml \v{12}Return to your fortress, you prisoners who have hope. \\
\poemll    Even today I am telling you: \\
\poemlll       In return I will repay you double. \\
\poeml \v{13}For I have bent Judah as if it were my bow, \\
\poemll    loading it with Ephraim. \\
\poeml I raised up your sons, Zion, \\
\poemll    against your sons, Greece, \\
\poemlll       wielding you like a mighty warrior's sword.''
\passage{Israel's \divine{Lord} Appears}
\poeml \v{14}The \divine{Lord} will appear over them--- \\
\poemll    his arrow will shoot like lightning. \\
\poeml The Lord \divine{God} will blow the trumpet, \\
\poemll    and go out with the southern windstorm. \\
\poeml \v{15}The \divine{Lord} of the Heavenly Armies will defend them; \\
\poemll    they will devour and conquer those who sling stones. \\
\poeml They will drink and be boisterous like those who are drunk. \\
\poemll    They will be filled to the brim with blood,\fnote{The Heb. lacks \fbib{with blood}} \\
\poemlll       like the corners of the altar. \\
\poeml \v{16}The \divine{Lord} their God will save them at that time\fnote{Lit. \fbib{day}} \\
\poemll    as his very own flock of people, \\
\poeml because they are his crown jewels, \\
\poemll    exalted throughout his land. \\
\poeml \v{17}For how great is his goodness, \\
\poemll    and how great is his beauty! \\
\poeml Grain will make the young men thrive, \\
\poemll    and new wine the virgins.
\end{poetry}
\labelchapt{10}
\passage{The \divine{Lord}'s Care for Judah}

\begin{poetry}
\poeml \chapt{10}
\v{1}``Ask the \divine{Lord} for rain in the spring\fnote{Lit. \fbib{the season of the latter rain}; i.e. the March and April rains}--- \\
\poemll    the \divine{Lord} who fashions lightning thunderstorms, \\
\poeml giving rain showers to mankind\fnote{Lit. \fbib{them}} \\
\poemll    along with grain in the fields.\fnote{Lit. \fbib{fields to a man}} \\
\poeml \v{2}Truly the family idols\fnote{Lit. \fbib{the teraphim;} i.e. household images apparently used for divination} talk nonsense \\
\poemll    and the diviners discern lies, \\
\poemlll       describing delusional dreams. \\
\poeml Since their comfort is vacuous, \\
\poemll    they wander off on their own like sheep, \\
\poemlll       because there is no shepherd. \\
\poeml \v{3}``Against the shepherds my anger rises--- \\
\poemll    I am punishing the leaders\fnote{Lit. \fbib{rams}} also, \\
\poeml because the \divine{Lord} of the Heavenly Armies has visited his flock, the house of Judah, \\
\poemll    appointing them as his royal war horse for battle. \\
\poeml \v{4}From them arises\fnote{The Heb. lacks \fbib{arises}} the cornerstone and tent peg, \\
\poemll    from them the battle bow, \\
\poemlll       from them arise all sorts of oppressive rulers. \\
\poeml \v{5}They will be like mighty soldiers \\
\poemll    who trample mud in the streets during battle. \\
\poeml They will fight because the \divine{Lord} is with them, \\
\poemll    and the opposing\fnote{The Heb. lacks \fbib{opposing}} horsemen will be confused. \\
\poeml \v{6}``I will fortify the house of Judah, \\
\poemll    and the house of Joseph I will save. \\
\poeml I will surely bring them back, \\
\poemll    because I care about them. \\
\poeml They will be as if I had never cast them away. \\
\poemll    Since I am the \divine{Lord} their God, \\
\poemlll       I will answer them. \\
\poeml \v{7}``The people of Ephraim will become like mighty soldiers; \\
\poemll    they will be glad, like those who have wine. \\
\poeml Their children will see this and rejoice; \\
\poemll    their hearts will find joy in the Lord. \\
\poeml \v{8}I will whistle for them, gathering them together, \\
\poemll    because I have redeemed them, \\
\poemlll       and they will multiply as they were before.\fnote{Lit. \fbib{will increase as they increased}} \\
\poeml \v{9}I will scatter them among the nations, \\
\poemll    and so they will remember me in distant countries. \\
\poeml They will rear their children, \\
\poemll    and they will return. \\
\poeml \v{10}I will bring them once again out of the land of Egypt, \\
\poemll    gathering them from Assyria. \\
\poeml I will bring them into the land of Gilead and Lebanon, \\
\poemll    but there will not be enough space for them. \\
\poeml \v{11}They\fnote{So LXX. MT reads \fbib{He will}} will pass through the sea of affliction, \\
\poemll    and they\fnote{So LXX. MT reads \fbib{He will}} will strike the waves\fnote{I.e. perhaps reminiscent of Elijah striking the Jordan River; cf. 2Kings 2:8} in that sea. \\
\poeml All of the depths of the Nile will evaporate, \\
\poemll    Assyria's arrogance will be brought down low, \\
\poemlll       and the ruling power\fnote{Lit. \fbib{the scepter}} of Egypt will disappear. \\
\poeml \v{12}``I will strengthen them in the \divine{Lord}, \\
\poemll    and they will march here and there in his name,'' \\
\poemlll       declares the \divine{Lord}.
\end{poetry}
\labelchapt{11}
\passage{Destruction of Lebanon and Bashan}

\begin{poetry}
\poeml \chapt{11}
\v{1}Open your doors, Lebanon, \\
\poemll    and fire will consume your cedars.\fnote{I.e. a genus of coniferous evergreen in the family \fbib{Pinaceae}; and so throughout the chapter} \\
\poeml \v{2}Wail, cypress tree, \\
\poemll    for the cedar has fallen \\
\poemlll       while the stately trees are destroyed. \\
\poeml Wail, oak trees of Bashan, \\
\poemll    for the old growth forest has been cut down. \\
\poeml \v{3}Hear\fnote{The Heb. lacks \fbib{Hear}} the wailing of the shepherds, \\
\poemll    for the magnificence of the forest\fnote{The Heb. lacks \fbib{of the forest}} is ruined! \\
\poeml Hear\fnote{The Heb. lacks \fbib{Hear}} the roar of the lions, \\
\poemll    for the Jordan's arrogance is ruined!
\end{poetry}
\passage{The Shepherd}

\v{4}This is what the \divine{Lord} my God says: ``Shepherd the flock marked for\fnote{Lit. \fbib{flock of the}} slaughter. \v{5}Their buyers slaughter them without being punished, continuing to sell them as they say, `Bless the \divine{Lord}!' and, `I'm rich!' Meanwhile, their shepherds show them no compassion. \v{6}Therefore I will no longer show compassion upon those who live in the land,'' declares the \divine{Lord}. ``Look! I will deliver every single person into the control\fnote{Lit. \fbib{hand}} of his neighbor and into the control\fnote{Lit. \fbib{hand}} of the king. Even though they assault the land, I will not deliver it from their control.''\fnote{Lit. \fbib{hand}}

\v{7}So I became shepherd of the flock marked for\fnote{Lit. \fbib{flock of the}} slaughter, paying attention to the oppressed of the flock. I took two staffs---naming one ``Pleasant'' and the other one ``Union''---and then I pastured the flock. \v{8}In a single month I got rid of three shepherds because I grew tired of them, and they despised me. \v{9}So I said, ``I will no longer be your shepherd. Let those who are about to die perish, and let what is about to be destroyed be destroyed. As for the survivors, let them devour each other.''

\v{10}Then I took the staff that I had named ``Pleasant'' and broke it, showing I was breaking my covenant that I had made with all of the people. \v{11}It was broken at that time\fnote{Lit. \fbib{day}} so the oppressed of the flock who were observing me would know that it had been a message from the \divine{Lord}.

\v{12}I told them, ``If it's alright with you, pay me what I've earned. But if it isn't, don't.''

So they paid out what I had earned---30 pieces of silver.\fnote{The Heb. lacks a specific unit of measurement}

\v{13}Then the \divine{Lord} told me, ``Throw the money\fnote{The Syr reads \fbib{Throw it}} into the treasury\fnote{According to Syr}---that magnificent value they placed on me!''

So I took the 30 shekels of\fnote{The Heb. lacks a specific unit of measurement} silver and threw them into the treasury\fnote{According to Syr} of the Temple of the \divine{Lord}. \v{14}Then I broke my second staff---the one I had named ``Union''---breaking the union between the house of Judah and the house of Israel.
\passage{God's Curse on the Worthless Shepherd}

\v{15}The \divine{Lord} told me, ``Pick up the tools of a worthless shepherd again, \v{16}for I am now raising up a shepherd in the land who will neither search for the lost, nor care for the young, nor fix the broken, nor sustain the healthy. Instead, he will devour the meat of the best of the sheep, tearing off their hoofs.''

\begin{poetry}
\poeml \v{17}``Woe to the worthless shepherd, \\
\poemll    who deserts the flock! \\
\poeml May the sword strike his arm \\
\poemll    and his right eye. \\
\poeml May his arm wither \\
\poemll    and his right eye be completely blind.''
\end{poetry}
\labelchapt{12}
\passage{The \divine{Lord} will Deliver Jerusalem}

\chapt{12}
\v{1}A declaration: a message from the \divine{Lord} to Israel. ``The \divine{Lord}, who stretches out the heavens, who lays the foundation of the earth, and who frames the spirit of man within himself, declares, \v{2}`Look, I am making Jerusalem an unstable cup\fnote{I.e. one that cannot be set down without spilling its contents} toward all of its surrounding armies when they lay siege against Judah and Jerusalem. \v{3}It will come about at that time\fnote{Lit. \fbib{day}} that I will make Jerusalem a heavy weight; so everyone who burdens themselves with it will be crushed,\fnote{Lit. \fbib{sliced}} even though all of the nations of the earth gather themselves against it. \v{4}At that time,'\fnote{Lit. \fbib{day}} declares the \divine{Lord}, `I will strike every horse with panic and every rider with insanity. I will keep my eyes on the house of Judah, but I will blind every horse of the invading\fnote{The Heb. lacks \fbib{invading}} armies. \v{5}The leaders of Judah will say to themselves, ``Those who live in Jerusalem are my strength through the \divine{Lord} of the Heavenly Armies, their God.'' \v{6}`At that time,\fnote{Lit. \fbib{day}} I will make the leaders of Judah like a brazier filled with blazing wood, or like a torch setting fire to harvested grain. They will devour all the invading\fnote{Lit. \fbib{surrounding}} armies, both on the right hand and on the left. As a result, Jerusalem will again be inhabited in its rightful place---as the real\fnote{The Heb. lacks \fbib{the real}} Jerusalem.'\,''

\v{7}The \divine{Lord} will deliver the tents of Judah first, so that neither the glory of the house\fnote{Or \fbib{family}} of David nor the glory of the inhabitants of Jerusalem overshadows Judah. \v{8}At that time,\fnote{Lit. \fbib{day}} the \divine{Lord} will defend those who live in Jerusalem, and the one who is feeble among them at that time will be like David. The entire house of David will be like God---indeed, like the angel of the \divine{Lord} in their midst!

\v{9}```At that time,\fnote{Lit. \fbib{day}} I will search out and destroy all of the nations who have come against Jerusalem. \v{10}I will pour out on the house of David and on the residents of Jerusalem a spirit of grace and of supplications, and they will look to me---the one whom they pierced.'\,''\fnote{The quotation possibly continues through 13:1}
\passage{Mourning in Jerusalem}

Then they will mourn for him, as for an only son. They will grieve bitterly for him, as for a firstborn son. \v{11}At that time,\fnote{Lit. \fbib{day}} Jerusalem will mourn deeply, like the mourning about Hadad-rimmon\fnote{Cf. 2Chr 35:20-25} that took place in the plain of Megiddo. \v{12}And so the land will mourn, families by families, alone by themselves---the family of the house of David by itself with their wives by themselves, the family of the house of Nathan by itself with their wives by themselves, \v{13}the family of the house of Levi by itself with their wives by themselves, the family of Shimei by itself with their wives by themselves---\v{14}all of the surviving families by themselves, along with their wives by themselves.\chapt{13}
\v{1}At that time,\fnote{Lit. \fbib{day}} a fountain will be opened for the house of David and for those who live in Jerusalem so they can be cleansed from\fnote{The Heb. lacks \fbib{so they can be cleansed from}} sin and ceremonial impurity.
\labelchapt{13}
\passage{Cessation of Prophecy}

\v{2}``At that time,''\fnote{Lit. \fbib{day}} declares the \divine{Lord} of the Heavenly Armies, ``I will eliminate the names of the idols from the land, and they will not be remembered anymore. I will also force both prophet and demon\fnote{Lit. \fbib{and unclean spirit}} to leave the land. \v{3}It will also come about that if any man would dare to\fnote{Lit. \fbib{would still}} prophesy, then his father and his mother who bore him will respond to him, `You will not live, because you are speaking lies in the name of the \divine{Lord}.' Then his father and mother who bore him will stab him for prophesying. \v{4}Furthermore, it will come about at that time\fnote{Lit. \fbib{day}} that every prophet will become ashamed of his vision as he prophesies. They will wear no rough garments intended to deceive others.''\fnote{The Heb. lacks \fbib{others}}
\passage{The Injured Servant of Mankind}

\v{5}``He\fnote{MT subject is third sing.; i.e. the \fbib{\divine{Lord}.}} will say, `I am no mere\fnote{The Heb. lacks \fbib{mere}} prophet. A servant of mankind am I, because a man dedicated to this\fnote{The Heb. lacks \fbib{to this}} have I been from my youth.'

\v{6}``Someone will say to him, `What are these injuries to your hands?'

``He will reply, `{\ldots}what I received at my friend's house.'

\begin{poetry}
\poeml \v{7}``Arise, sword, against my shepherd, \\
\poemll    against the mighty one who is related to me,'' \\
\poemlll       declares the \divine{Lord} of the Heavenly Armies. \\
\poeml ``Strike the shepherd, \\
\poemll    the sheep will be scattered, \\
\poemlll       and I will turn against the insignificant ones. \\
\poeml \v{8}It will come about in all of the land,'' \\
\poemll    declares the \divine{Lord}, \\
\poeml ``that two thirds of the people living there will die, \\
\poemll    but a third will survive who live there. \\
\poeml \v{9}And I will bring that surviving third through, \\
\poemll    testing them as if through fire, \\
\poeml purifying them like silver, \\
\poemll    assaying them like gold. \\
\poeml They will call on my name, \\
\poemll    and I will answer them. \\
\poeml I will say, `This is my people,' \\
\poemll    and they will say, `The \divine{Lord} is my God.'\,''
\end{poetry}
\labelchapt{14}
\passage{The \divine{Lord} Comes to the Mount of Olives}

\chapt{14}
\v{1}``Look! A day is coming for the \divine{Lord}, when your plunder will be divided among you. \v{2}I will gather all the nations against Jerusalem, to lay siege against it. The city will be captured, the houses will be ransacked, the women raped, and half of the city will go into exile, but the remaining people will not be cut off from the city. \v{3}Then the \divine{Lord} will go out to battle against those nations, waging war as in a day of battle. \v{4}His feet will stand in that day on the Mount of Olives, east of Jerusalem. Then the Mount of Olives will be split in two from east to west, forming\fnote{The Heb. lacks \fbib{forming}} a very large valley, with half of the mountain moving toward the north and half toward the south. \v{5}You will run away through my mountain valley, because the valley of the mountains will extend as far as Azal. You will flee, as you fled from the earthquake during the reign of King Uzziah of Judah. And so the \divine{Lord} my God will come, and all his holy ones will be accompanying you.''
\passage{A Unique Day}

\v{6}``At that time,\fnote{Lit. \fbib{day}} the daylight will be neither bright nor overcast. \v{7}It will be a unique day, known only to the \divine{Lord}---neither daytime nor nighttime---and it will come about at twilight there will be light! \v{8}At that time,\fnote{Lit. \fbib{day}} flowing waters will run perennially\fnote{Lit. \fbib{waters in summer and in winter will be}} from Jerusalem, half toward the Dead\fnote{Lit. \fbib{eastern}} Sea and half to the Mediterranean\fnote{Lit. \fbib{western}} Sea. \v{9}The \divine{Lord} will be king over all the earth at that time.\fnote{Lit. \fbib{day}} There\fnote{Or \fbib{earth. At that time, there}} will be one \divine{Lord}, and his name the only one. \v{10}The entire land will become like the Arabah plain from Geba\fnote{I.e. an ancient city about 6 miles northeast of Jerusalem} to Rimmon, south of Jerusalem. It will be raised up and inhabited where it is, from the Gate of Benjamin to the First Gate, then to the Corner Gate, to the Hananel Tower, and to the king's winepresses. \v{11}People\fnote{Lit. \fbib{They}} will live there, there will be no more destruction, and Jerusalem will be safely inhabited.''
\passage{God's Judgment on Jerusalem's Attackers}

\v{12}``This will be the plague with which the \divine{Lord} inflicts all of the people who have attacked Jerusalem: he will cause their flesh to rot away, even while they're standing on their feet. He will cause their eyes to rot away in their sockets, and their tongues to rot away in their mouths. \v{13}At that time,\fnote{Lit. \fbib{day}} they will be stricken with a terrible panic from the \divine{Lord}. Everyone will attack each other. \v{14}Judah, too, will fight at Jerusalem. Then the wealth of the surrounding nations will be gathered up: gold, silver, and clothing in great abundance. \v{15}A similar plague will also strike horses, mules, camels, donkeys, and all of the animals in those camps.''
\passage{Discipline of the Nations}

\v{16}``It will come about that all of the survivors of the nations who came against Jerusalem will come there from year to year to worship the King, the \divine{Lord} of the Heavenly Armies, and to observe the Festival of Tents. \v{17}If anyone from the families of the earth will not come to Jerusalem to worship the King, the \divine{Lord} of the Heavenly Armies, there will be no rain for them. \v{18}If the people of Egypt do not come to Jerusalem\fnote{The Heb. lacks \fbib{to Jerusalem}} to take part, they will have no annual Nile overflow.\fnote{The Heb. lacks \fbib{annual Nile overflow}} A plague will come from the \divine{Lord} to strike the nations who do not come to observe the Festival of Tents. \v{19}This will be the punishment for Egypt and all nations who do not come to observe the Festival of Tents.''
\passage{Holiness to the \divine{Lord}}

\v{20}``At that time,\fnote{Lit. \fbib{day}} there will be written on the bells of the horses:

\divine{Holiness to the Lord}

and the pots in the Temple of the \divine{Lord} will be like the bowls in front of the altar---\v{21}every pot in Jerusalem and in Judah will be consecrated to the \divine{Lord} of the Heavenly Armies. Everyone who offers sacrifices will come, will take them, and will cook in them. Furthermore, at that time,\fnote{Lit. \fbib{day}} there will no longer be a Canaanite in the Temple of the \divine{Lord} of the Heavenly Armies.''
