\bookheader{Malachi}
\labelbook{Mal}

\bookpretitle{The Book of the Prophet}
\booktitle{Malachi}

\labelchapt{1}
\passage{God's First Complaint: Against His People---A Despised Love}

\chapt{1}
\v{1}A declaration: a message from the \divine{Lord} to Israel by Malachi.\fnote{The Heb. name \fbib{Malachi} means \fbib{My messenger}}

\v{2}``I've loved you,'' says the \divine{Lord}. ``But you ask, `How have you loved us?'

``Was not Esau Jacob's brother?'' declares the \divine{Lord}, ``yet I loved Jacob, \v{3}rejected\fnote{Lit. \fbib{hated}} Esau, turned his mountains into a wasteland, and gave\fnote{The Heb. lacks \fbib{gave}} his inheritance to desert jackals. \v{4}Even though Edom may claim, `We were crushed, but we will return and rebuild the ruins,' this is what the \divine{Lord} of the Heavenly Armies says: ``They may rebuild, but I'll demolish. People\fnote{Lit. \fbib{they}} will call them, `The Wicked Land,' and, `The People With Whom the \divine{Lord} is Forever Angry.' \v{5}Your own eyes will see this, and you will say, `Great is the \divine{Lord} even beyond the borders of Israel!'\,''
\passage{God's Second Complaint: Against His Priests---A Despised Offering}

\v{6}``A son honors his father and a servant his master. So if I'm a father, where is my honor? And if I'm a master, where is my respect?'' says the \divine{Lord} of the Heavenly Armies to you priests who are despising my name. ``But you ask, `How have we despised your name?' \v{7}By presenting defiled food on my altar. And you ask, `How have we defiled you?' By saying, `The Table of the \divine{Lord} is contemptible.' \v{8}When you bring blind animals for sacrifice, is that not wrong? And when you sacrifice crippled or diseased animals,\fnote{The Heb. lacks \fbib{animals}} is that not wrong? Offer that to your governor---would he be pleased with you or receive you favorably?'' asks the \divine{Lord} of the Heavenly Armies. \v{9}``And now, go ahead and implore God by saying, `Be gracious to us.' Will he receive you favorably and accept offerings like that from your hand?'' asks the \divine{Lord} of the Heavenly Armies.
\passage{Useless Offerings and Useless Altar Fires}

\v{10}``Oh, that one of you would shut the Temple\fnote{The Heb. lacks \fbib{temple}} doors and not light useless fires on the altar! I'm not pleased with you,'' says the \divine{Lord} of the Heavenly Armies, ``and I'll accept no offerings from you.\fnote{Lit. \fbib{from your hands}} \v{11}Even so, from where the sun rises to where it sets my name will be great among the Gentiles. Incense will be brought to me\fnote{Lit. \fbib{to my name}} everywhere, along with pure offerings, because my name will be great among the Gentiles,'' says the \divine{Lord} of the Heavenly Armies. \v{12}``But you are profaning my name\fnote{Lit. \fbib{it}} by saying that the Table of the \divine{Lord} is defiled and that its fruit and its food are contemptible.

\v{13}``And you say, `What a burden!' and sniff contemptuously at it,'' says the \divine{Lord} of the Heavenly Armies, ``when you present maimed, crippled, and diseased animals,\fnote{The Heb. lacks \fbib{animals}} and when you bring the offering. Should I accept this from your hand?'' asks the \divine{Lord}. \v{14}``Cursed is the deceiver who has an acceptable\fnote{The Heb. lacks \fbib{acceptable}} male in his flock, and vows to give it,\fnote{The Heb. lacks \fbib{to give it}} but sacrifices a mutilated one to the \divine{Lord}. Indeed, I am a great king,'' says the \divine{Lord} of the Heavenly Armies, ``and my name is feared among the Gentiles.''
\labelchapt{2}
\passage{God's Third Complaint: Against His Priests---Failing to Honor Him}

\chapt{2}
\v{1}Now this commandment is for you priests: \v{2}``If you don't listen, and if you don't choose\fnote{Lit. \fbib{don't set your hearts}} to give honor to my name,'' says the \divine{Lord} of the Heavenly Armies, ``then I'll curse both you and your blessings.\fnote{Lit. \fbib{I'll send on you the curse and I'll curse your blessings}} I've even cursed them\fnote{Lit. \fbib{it}} already, because none of you are taking it to heart. \v{3}Look! I'm rebuking your descendants because of you, and I'll spread waste\fnote{I.e. the parts of the sacrificial animal that were discarded after the offering had been presented.} on your faces, the waste of your festival sacrifices, and you will be carried off with it.

\v{4}``You will know that I sent this commandment to you in order to continue my covenant with Levi,'' says the \divine{Lord} of the Heavenly Armies. \v{5}``My covenant with him was for life and peace, and I gave the commandments\fnote{Lit. \fbib{them}} to him so he would fear me. He did fear me as he stood in my presence.\fnote{Lit. \fbib{in the presence of my name}} \v{6}True teachings were in his mouth, and falsehood was not found on his lips. He walked with me peacefully and righteously, and he turned many from sin. \v{7}For the lips of a priest should preserve knowledge, and people should seek instruction from his mouth, because he's the messenger of the \divine{Lord} of the Heavenly Armies.

\v{8}``But you priests\fnote{The Heb. lacks \fbib{priests}} turned aside from the way, and by your teaching you caused many to stumble. You have violated the covenant of Levi,'' says the \divine{Lord} of the Heavenly Armies. \v{9}``So I also made you despised, humiliating you before all of the people, because you aren't following my ways and are showing partiality when you teach.''
\passage{A Plea and a Prayer from Malachi}

\v{10}Do we not have one father?\fnote{The reference is either to Abraham as ancestor of Israel or to God as heavenly Father.} Has not one God created us? Why does each of us act deceitfully, each man against his own brother, to profane the covenant of our ancestors? \v{11}Judah has become unfaithful, and a detestable thing was committed in Israel and Jerusalem. Indeed, Judah profaned the Holy Place of the \divine{Lord}, which he loves, and married a daughter of a foreign god. \v{12}May the \divine{Lord} exclude from the community\fnote{Lit. \fbib{tents}} of Jacob any man who does this, whoever he may be,\fnote{Lit. \fbib{he who awakens and he who answers}} even though he brings offerings to the \divine{Lord} of the Heavenly Armies.
\passage{God's Fourth Complaint: Against His Priests---Marital Abuses}

\v{13}``This is another thing you do: you flood the altar of the \divine{Lord} with tears, weeping and wailing because he no longer pays attention to your offering nor takes pleasure in it from your hand. \v{14}Yet you ask, `For what reason?' Because the \divine{Lord} acts as a witness between you and the wife of your youth, because you were unfaithful to her, your partner, the wife of your covenant. \v{15}Did he not make them\fnote{The Heb. lacks \fbib{them}} one? And the vestige of the spirit remains in him. And why did he make them one? He was seeking godly offspring. So guard yourselves in your spirit, and don't be unfaithful to the wife of your youth.

\v{16}``Indeed, the \divine{Lord} God of Israel says that he hates divorce, along with the one who conceals his violence by outward appearances,''\fnote{Lit. \fbib{by his garments}} says the \divine{Lord} of the Heavenly Armies. ``So guard yourselves carefully,\fnote{Lit. \fbib{yourselves in your spirit}} and don't be unfaithful.''
\passage{God's Fifth Complaint: Against His People---Complaining about God.}

\v{17}``You have wearied the \divine{Lord} with your words. You ask, `How have we wearied you?' By your saying, `All who do evil are good in the eyes of the \divine{Lord} and he's pleased with them,' or `Where is the God of justice?'\,''
\labelchapt{3}
\passage{The Coming of the Messenger}

\chapt{3}
\v{1}``Watch out! I'm sending my messenger, and he will prepare the way before me. Then suddenly, the \divine{Lord} you are looking for will come to his Temple. He is the messenger of the covenant whom you desire. Watch out! He is coming!'' says the \divine{Lord} of the Heavenly Armies.

\v{2}But who will survive the day when he comes? Or who can stand when he appears? Because he's like a refiner's fire and a launderer's soap, \v{3}he will sit refining and purifying silver, purifying the descendants of Levi, refining them like gold and silver. Then they'll bring a righteous offering to the \divine{Lord}. \v{4}Then the offering to the \divine{Lord} by Judah and Jerusalem will be acceptable as it was in the past, even as in former years.
\passage{The Judgment of God}

\v{5}``I'll come near to you for judgment. I'll be a witness, quick to speak against sorcerers, against adulterers, against those who swear falsely, against those who defraud the laborer of his wage, against those who defraud\fnote{The Heb. lacks \fbib{against those who defraud}} the widow and the orphan, against those who deprive the alien of justice, and against those who don't fear me,'' says the \divine{Lord} of the Heavenly Armies. \v{6}``Because I the \divine{Lord} don't change; therefore you children of Jacob aren't destroyed.''
\passage{God's Sixth Complaint: Against His People---Gifts and Offerings}

\v{7}``Ever since the time of your ancestors, you have turned away from my decrees and haven't kept them. Return to me and I'll return to you,'' says the \divine{Lord} of the Heavenly Armies. ``But you ask, `How will we return?' \v{8}``Will a person rob God? Yet you are robbing me! But you ask, `How are we robbing you?' ``By the tithe and the offering. \v{9}You are cursed under the curse---the entire nation---because you are robbing me!

\v{10}``Bring the entire tithe into the storehouse that there may be food in my house. So put me to the test in this right now,'' says the \divine{Lord} of the Heavenly Armies, ``and see if I won't throw open the windows\fnote{Or \fbib{floodgates}} of heaven for you and pour out on you blessing without measure. \v{11}And I'll prevent the devourer from harming you,\fnote{Lit. \fbib{I'll rebuke the devourer from you}} so that he does not destroy the crops of your land. Nor will the vines in your fields drop their fruit,'' says the \divine{Lord} of the Heavenly Armies.

\v{12}``Then all the nations will call you blessed, for you will be a land of delight,'' says the \divine{Lord} of the Heavenly Armies.
\passage{God's Seventh Complaint: Against His People---Slandering God}

\v{13}``You have spoken arrogant words against me,'' says the \divine{Lord}. ``Yet you ask, `What did we say against you?' \v{14}You said, `It is futile to serve God,' and, `What did we get out of it\fnote{Lit. \fbib{What gain}} when we carried out his requirements and went about like mourners in the presence of the \divine{Lord} of the Heavenly Armies?' \v{15}and, `Now we call the arrogant one blessed. Those who do evil prosper and those who challenge God escape the consequences.'\,''\fnote{The Heb. lacks \fbib{the consequences}}
\passage{The Repentance of the Righteous}

\v{16}Then those who feared the \divine{Lord} talked with each other. The \divine{Lord} listened and took note,\fnote{Lit. \fbib{heard}} and a scroll of remembrance was written in his presence about those who fear the \divine{Lord} and honor His name. \v{17}``They'll be mine,'' says the \divine{Lord} of the Heavenly Armies, ``in the day when I prepare my treasured possession. I'll spare them, just as a man spares his own son who serves him. \v{18}When you return, you will see the difference between the righteous and the wicked, and between the one who serves God and the one who does not.''
\labelchapt{4}
\passage{The Great Day of the \divine{Lord}\passagenote{Chap 4:1-6 is chapter 3:19-24 in MT}}

\chapt{4}
\v{1}``The coming day\fnote{I.e. the Day of the \fbib{\divine{Lord}}} is certainly going to burn like a furnace. All the arrogant and all who practice evil will be stubble---the coming day will set them on fire,'' says the \divine{Lord} of the Heavenly Armies, ``so that it will leave them neither root nor branch. \v{2}But the Sun of Righteousness will arise with healing in its light\fnote{Lit. \fbib{wings}; i.e. its sun beams} for those who fear my name. You will go out and leap like calves released\fnote{The Heb. lacks \fbib{released}} from their stalls \v{3}and trample down the wicked. Indeed, they will become ashes under the soles of your feet on the day I do this,'' says the \divine{Lord} of the Heavenly Armies.
\passage{The Coming of Elijah the Prophet}

\v{4}``Remember the Law of Moses my servant that I gave him at Horeb for all Israel---both the decrees and laws.

\v{5}``Pay attention! I'm sending Elijah the prophet to you before the great and dreadful Day of the \divine{Lord} comes, \v{6}and he will turn the hearts of fathers to children, and the hearts of children to their fathers. Otherwise, I'll come, strike the land, and utterly destroy it.''
