\bookheader{1 Thessalonians}
\labelbook{1Thes}

\bookpretitle{The Letter from Paul Called}
\booktitle{First Thessalonians}

\labelchapt{1}
\passage{Greetings}

\chapt{1}
\v{1}From:\fnote{The Gk. lacks \fbib{From}} Paul, Silvanus,\fnote{I.e. Silas} and Timothy.

To: The church of the Thessalonians in union with God the Father and the Lord Jesus, the Messiah.\fnote{Or \fbib{Christ}}

May grace and peace from God our Father and the Lord Jesus, the Messiah,\fnote{Or \fbib{Christ}; other mss. lack \fbib{from God our Father and the Lord Jesus, the Messiah}} be yours!
\passage{Paul's Prayer for the Thessalonians}

\v{2}We always thank God for all of you when we mention you in our prayers. \v{3}In the presence of our God and Father, we constantly remember how your faith is active,\fnote{Lit. \fbib{your work of faith}} your love is hard at work,\fnote{Lit. \fbib{your labor of love}} and your hope in our Lord Jesus the Messiah\fnote{Or \fbib{Christ}} is enduring.\fnote{Lit. \fbib{the endurance of hope of our Lord Jesus, the Messiah}} \v{4}Brothers whom God loves, we know that he has chosen you, \v{5}for the gospel we brought\fnote{Lit. \fbib{our gospel}} did not come to you in words only, but also with power, with the Holy Spirit, and with deep conviction. Indeed,\fnote{Lit. \fbib{Just as}} you know what kind of people we proved to be while we were with you, acting on your behalf.

\v{6}You became imitators of us and of the Lord. In spite of a great deal of suffering, you welcomed the word with the joy that the Holy Spirit produces.\fnote{Or \fbib{the joy of the Holy Spirit}} \v{7}As a result, you became a model for all the believers in Macedonia and Achaia. \v{8}From you the word of the Lord has spread out not only in Macedonia and Achaia, but also in every place where your faith in God has become known. As a result, we do not need to say anything about it.

\v{9}For people\fnote{Lit. \fbib{they}} keep telling us what kind of welcome you gave us and how you turned away from idols to serve a living and true God \v{10}and to wait for his Son whom he raised from the dead to come back\fnote{The Gk. lacks \fbib{to come back}} from heaven. This Jesus is the one who rescues us from the coming wrath.
\labelchapt{2}
\passage{Paul Recalls His Visit to the Thessalonians}

\chapt{2}
\v{1}For you yourselves know, brothers, that our visit to you was not a waste of time. \v{2}As you know, we suffered persecution and were mistreated in Philippi. Yet we were encouraged by our God to tell you his\fnote{Lit. \fbib{God's}} gospel in spite of strong opposition. \v{3}For our appeal to you\fnote{The Gk. lacks \fbib{to you}} does not spring from deceit, impure motives, or trickery. \v{4}Rather, because we have been approved by God to be entrusted with the gospel, we speak as we do, not trying to please people but God, who tests our motives.

\v{5}As you know, we did not come with flattering words or with a scheme to make money. God is our witness! \v{6}We did not seek praise from people---from you or from anyone else--- \v{7}even though as apostles of the Messiah\fnote{Or \fbib{Christ}} we might have made such demands. Instead, we were gentle\fnote{Other mss. read \fbib{infants}} among you, like a nursing mother tenderly caring for her own children. \v{8}We cared so deeply for you that we were determined to share with you not only the gospel of God but our very lives. That is how dear you were to us. \v{9}Brothers, you remember our labor and toil. We worked night and day so that we would not become a burden to any of you while we proclaimed the gospel of God to you. \v{10}You and God are witnesses of how pure, honest, and blameless our conduct was among you who believe. \v{11}You know very well that we treated each of you the way a father treats\fnote{The Gk. lacks \fbib{treats}} his children. \v{12}We comforted and encouraged you, urging you to live in a manner worthy of God, who calls\fnote{Other mss. read \fbib{called}} you into his kingdom and glory.
\passage{How the Thessalonians Welcomed the Gospel}

\v{13}Here is another reason why we constantly give thanks to God: When you received God's word, which you heard from us, you did not accept it as the word of humans but for what it really is---the word of God, which is at work in you who believe. \v{14}For you, brothers, became imitators of the churches of God in Judea that are in union with the Messiah\fnote{Or \fbib{Christ}} Jesus. You suffered the same persecutions from the people of your own country as they did from those Jews \v{15}who killed the Lord Jesus and the\fnote{Other mss. read \fbib{their own}} prophets, who have persecuted us, and who please neither God nor any group of people, \v{16}as they try to keep us from telling the gentiles how they can be saved. As a result, they are constantly adding to the number of sins they have committed.\fnote{The Gk. lacks \fbib{they have committed}} However, wrath has overtaken them at last!
\passage{Timothy's Report to Paul}

\v{17}Brothers, although we have been separated from you for a little while---in person but not in heart---we eagerly desire to see you again face to face. \v{18}That is why we wanted to come to you. Certainly I, Paul, wanted to come\fnote{The Gk. lacks \fbib{wanted to come}} time and again, but Satan blocked our way. \v{19}After all, who is our hope, joy, or reason for\fnote{Lit. \fbib{or crown of}} rejoicing in the presence of our Lord Jesus at his coming? It is you, isn't it? \v{20}Yes, you are our glory and joy!
\labelchapt{3}

\chapt{3}
\v{1}Therefore, when we could stand it no longer, we decided to remain alone in Athens \v{2}and send Timothy, our brother who works with us for God in the gospel of the Messiah,\fnote{Or \fbib{Christ}; other mss. lack \fbib{of the Messiah}} to strengthen and encourage you in your faith, \v{3}so that no one would be shaken by these persecutions, for which you are aware that we were destined. \v{4}In fact, when we were with you, we told you ahead of time that we were going to suffer persecution. And as you know, that is what happened. \v{5}But when I could stand it no longer, I sent Timothy\fnote{The Gk lacks \fbib{Timothy}} to find out about your faith. I was afraid that the tempter had tempted you in some way, and that our work had been a waste of time.
\passage{Paul Rejoices about the Thessalonians}

\v{6}But Timothy has just now returned from visiting\fnote{The Gk. lacks \fbib{visiting}} you and has told us the good news about your faith and love. He also told us that you always have fond memories of us and want to see us, just as we want to see you. \v{7}That's why, brothers, in all our distress and persecution we have been encouraged about you by your faith. \v{8}For now we can go on living, as long as you continue to stand firm in the Lord. \v{9}How can we thank God enough for you in return for all the joy that we have in God's presence because of you? \v{10}We pray very hard night and day that we may see you again face to face, so that we may equip you with\fnote{The Gk. lacks \fbib{you with}} whatever is lacking in your faith.

\v{11}Now may our God and Father and our Lord Jesus provide a way for us to visit you. \v{12}May the Lord greatly increase your love\fnote{Lit. \fbib{cause you to increase and abound in love}} for each other and for all people, just as we love you.\fnote{Lit. \fbib{as we for you}} \v{13}Then your hearts will be strong, blameless, and holy in the presence of God, who is our Father, when our Lord Jesus appears with all his saints.\fnote{Or \fbib{holy ones}; other mss. read \fbib{saints. Amen}}
\labelchapt{4}
\passage{Instructions on the Way Christians Should Live}

\chapt{4}
\v{1}Now then, brothers, you learned from us how you ought to live and to please God, as in fact you are doing. We ask and encourage you in the Lord to do so even more. \v{2}You know what instructions we gave you through the Lord Jesus. \v{3}For it is God's will that you be sanctified: You must abstain from sexual immorality. \v{4}Each of you must know how to control his own body\fnote{Lit. \fbib{vessel}} in a holy and honorable manner, \v{5}not with passion and lust like the gentiles who do not know God. \v{6}Furthermore, you must never take advantage of or exploit a brother in this regard, because the Lord avenges all these things, just as we already told you and warned you. \v{7}For God did not call us to be impure, but to be holy. \v{8}Therefore, whoever rejects this instruction\fnote{The Gk. lacks \fbib{instruction}} is not rejecting human authority but God, who gives you his Holy Spirit.

\v{9}Now you do not need anyone to write to you about brotherly love, since you have been taught by God to love each other. \v{10}In fact, you are showing love to all the brothers throughout Macedonia, but we urge you, brothers, to keep on doing this even more. \v{11}Also, make it your goal to live quietly, to mind your own business, and to work with your own hands, as we instructed you, \v{12}so that you may win the respect of outsiders, and have need of nothing.
\passage{Comfort about Christians who Have Died}

\v{13}But we do not want you to be ignorant, brothers, about those who have died, so that you may not grieve like other people who have no hope. \v{14}For if we believe that Jesus died and rose again, even so through Jesus God will bring those who have died with him. \v{15}For we declare to you what the Lord has told us to say:\fnote{Lit. \fbib{you by the word of the Lord:}} We who are alive and remain until the coming of the Lord will by no means precede those who have died. \v{16}With a shout of command, with the archangel's call, and with the sound of God's trumpet, the Lord himself will come down from heaven, and the dead who belong to the Messiah\fnote{Or \fbib{Christ}} will rise first. \v{17}Then we who are alive and remain will be caught up in the clouds together with them to meet the Lord in the air. And so we will be with the Lord forever. \v{18}So then, encourage one another with these words.
\labelchapt{5}
\passage{Be Ready for the Day of the Lord}

\chapt{5}
\v{1}Now you do not need to have anything written to you about times and dates, brothers, \v{2}for you yourselves know very well that the Day of the Lord will come like a thief in the night. \v{3}When people\fnote{Lit. \fbib{they}} say, ``There is peace and security,'' destruction will strike them as suddenly as labor pains come\fnote{The Gk. lacks \fbib{come}} to a pregnant woman, and they will not be able to escape.

\v{4}However, brothers, you are not in the darkness, in order that the Day of the Lord\fnote{The Gk. lacks \fbib{of the Lord}} might surprise you like a thief. \v{5}For all of you are children of the light and children of the day. We do not belong to the night or to darkness. \v{6}Therefore, let's not fall asleep like others do, but let's stay awake and be sober. \v{7}For people who go to sleep, go to sleep at night; and people who get drunk, get drunk at night. \v{8}But since we belong to the day, let's be sober. We must put on the breastplate of faith and love, and the hope of salvation as a helmet. \v{9}For God has not destined us to receive\fnote{The Gk. lacks \fbib{receive}} wrath but to obtain salvation through our Lord Jesus, the Messiah,\fnote{Or \fbib{Christ}} \v{10}who died for us in order that, whether we are awake or asleep, we may live together with him. \v{11}So then, encourage one another and build each other up, as you are doing.
\passage{Paul Gives Final Instructions to the Church}

\v{12}Brothers, we ask you to show your appreciation for those who work among you, set an example for you in the Lord, and instruct\fnote{Or \fbib{admonish}} you. \v{13}Hold them in the highest regard, loving them because of their work. Live in peace with each other. \v{14}We urge you, brothers, to admonish\fnote{Or \fbib{instruct}} those who are idle,\fnote{Or \fbib{disorderly}} cheer up those who are discouraged, and help those who are weak. Be patient with everyone. \v{15}Make sure that no one pays back evil for evil. Instead, always pursue what is good for each other and for everyone else. \v{16}Always be joyful. \v{17}Continually be prayerful. \v{18}In everything be thankful, because this is God's will for you in the Messiah\fnote{Or \fbib{Christ}} Jesus. \v{19}Do not put out the Spirit's fire.\fnote{I.e. stifle the Spirit's work} \v{20}Do not despise prophecies. \v{21}Instead, test everything. Hold on to what is good. \v{22}Keep away from every kind of evil.
\passage{Final Greeting}

\v{23}May the God of peace himself make you holy in every way. And may your whole being---spirit, soul, and body---remain blameless when our Lord Jesus, the Messiah,\fnote{Or \fbib{Christ}} appears. \v{24}The one who calls you is faithful, and he will continue to be faithful.\fnote{The Gk. lacks \fbib{faithful}} \v{25}Brothers, pray\fnote{Other mss. read \fbib{also pray}} for us. \v{26}Greet all the brothers with a holy kiss. \v{27}I order you by the Lord to have this letter read to all the brothers. \v{28}May the grace of our Lord Jesus, the Messiah,\fnote{Or \fbib{Christ}} be with you! Amen.\fnote{Other mss. lack \fbib{Amen}}
