\bookheader{2 Thessalonians}
\labelbook{2Thes}

\bookpretitle{The Letter of Paul Called}
\booktitle{Second Thessalonians}

\labelchapt{1}
\passage{Greetings}

\chapt{1}
\v{1}From:\fnote{\fbackref{1:1} The Gk. lacks \fbib{From}} Paul, Silvanus,\fnote{\fbackref{1:1} I.e. Silas} and Timothy.

To: The church of the Thessalonians in union with God our Father and the Lord Jesus, the Messiah.\fnote{\fbackref{1:1} Or \fbib{Christ}}

\v{2}May grace and peace from God our Father and the Lord Jesus, the Messiah,\fnote{\fbackref{1:2} Or \fbib{Christ}} be yours!
\passage{Endurance in Affliction}

\v{3}Brothers, at all times we are obligated to thank God for you. It is right to do this\fnote{\fbackref{1:3} Lit. \fbib{As is right}} because your faith is growing all the time and the love every one of you has for each other is increasing. \v{4}As a result, we rejoice about you among God's churches---about your endurance and faith through all the persecutions and afflictions you are experiencing. \v{5}This is evidence of God's righteous judgment and is intended to make you worthy of God's kingdom, for which you are suffering. \v{6}Certainly it is right for God to pay back those who afflict you with affliction, \v{7}and to give us who are afflicted relief when the Lord Jesus is revealed from heaven with his mighty angels \v{8}in blazing fire. He will take revenge on those who do not know God and on those who refuse to obey the gospel of our Lord Jesus. \v{9}Such people will suffer the punishment of eternal destruction by being separated from the Lord's presence and from his glorious power \v{10}when he comes to be glorified by his saints and to be regarded with wonder on that day by all who have believed---including you, because you believed our testimony.

\v{11}With this in mind, we always pray for you, asking\fnote{\fbackref{1:11} The Gk. lacks \fbib{asking}} that our God might make you worthy of his calling and that through his power he might help you accomplish every good desire and faithful action. \v{12}That way the name of our Lord Jesus will be glorified by you, and you by him, according to the grace of our God and Lord, Jesus, the Messiah.\fnote{\fbackref{1:12} Or \fbib{Christ}}
\labelchapt{2}
\passage{The Lawless One}

\chapt{2}
\v{1}Now we ask you, brothers, regarding the coming of our Lord Jesus, the Messiah,\fnote{\fbackref{2:1} Or \fbib{Christ}} and our gathering together to him, \v{2}not to be so quickly upset\fnote{\fbackref{2:2} Lit. \fbib{shaken in mind}} or alarmed when someone claims that we said,\fnote{\fbackref{2:2} Lit. \fbib{as though by us}} either by some spirit, conversation, or letter that the Day of the Lord has already come. \v{3}Do not let anyone deceive you in any way, for it will not come unless the rebellion\fnote{\fbackref{2:3} Or \fbib{apostasy}} takes place first and the man of sin,\fnote{\fbackref{2:3} Other mss. read \fbib{man of lawlessness}} who is destined for destruction,\fnote{\fbackref{2:3} Lit. \fbib{the son of destruction}} is revealed. \v{4}He opposes and exalts himself above every so-called god and object of worship. As a result, he seats himself in the sanctuary\fnote{\fbackref{2:4} Or \fbib{temple}} of God and himself declares that he is God.

\v{5}Don't you remember that I repeatedly told you about these things when I was still with you? \v{6}You know what it is that is now holding him back, so that he will be revealed when his time comes. \v{7}For the secret of this lawlessness is already at work, but only until the person now holding it back gets out of the way. \v{8}Then the lawless one will be revealed, whom the Lord\fnote{\fbackref{2:8} Some MSS read \fbib{Lord Jesus}} will destroy with the breath of his mouth, rendering him powerless by the manifestation of his coming.

\v{9}The coming of the lawless one\fnote{\fbackref{2:9} Lit. \fbib{His coming}} will be accompanied by the power of Satan. He will use every kind\fnote{\fbackref{2:9} Lit. \fbib{In every kind}} of power, including miraculous signs, lying wonders, \v{10}and every type of evil to deceive\fnote{\fbackref{2:10} Lit. \fbib{every evil deception}} those who are dying, those who refused to love the truth that would save them.\fnote{\fbackref{2:10} Lit. \fbib{so that they might be saved}} \v{11}For this reason, God will send them a powerful delusion so that they will believe the lie. \v{12}Then all who have not believed the truth but have taken pleasure in unrighteousness will be condemned.

\v{13}At all times we are obligated to thank God for you, brothers who are loved by the Lord, because God chose you to be the first fruits\fnote{\fbackref{2:13} Other mss. read \fbib{from the beginning}} for salvation through sanctification by the Spirit and through faith in the truth. \v{14}With this purpose in mind, he called you through our proclamation of the\fnote{\fbackref{2:14} The Gk. lacks \fbib{proclamation of the}} gospel so that you would obtain the glory of our Lord Jesus, the Messiah.\fnote{\fbackref{2:14} Or \fbib{Christ}} \v{15}So then, brothers, stand firm, and cling to the traditions that you were taught by us, either by word of mouth\fnote{\fbackref{2:15} Lit. \fbib{by word}} or by our letter. \v{16}May our Lord Jesus, the Messiah\fnote{\fbackref{2:16} Or \fbib{Christ}} himself, and may God our Father, who loved us and by his grace gave us eternal comfort\fnote{\fbackref{2:16} Or \fbib{encouragement}} and good hope, \v{17}encourage your hearts and strengthen you in every good action and word.
\labelchapt{3}
\passage{An Example to Follow}

\chapt{3}
\v{1}Finally, brothers, pray for us---that the word of the Lord may spread rapidly, and that it may be honored the way it is among you. \v{2}Also pray\fnote{\fbackref{3:2} The Gk. lacks \fbib{pray}} that we may be rescued from worthless and evil people, since not everyone holds to the faith.\fnote{\fbackref{3:2} Or \fbib{has faith}} \v{3}But the Lord is faithful and will strengthen you and protect you from the evil one. \v{4}We have confidence in the Lord\fnote{\fbackref{3:4} Lit. \fbib{in the Lord concerning you}} that you are doing and will continue to do what we command. \v{5}May the Lord direct your hearts to the love of God and to the endurance of the Messiah.\fnote{\fbackref{3:5} Or \fbib{Christ}}

\v{6}In the name of our Lord Jesus, the Messiah,\fnote{\fbackref{3:6} Or \fbib{Christ}} we command you, brothers, to keep away from every brother who is living in idleness\fnote{\fbackref{3:6} Or \fbib{is leading a disorderly life}} and not living\fnote{\fbackref{3:6} The Gk. lacks \fbib{living}} according to the tradition that they received\fnote{\fbackref{3:6} Other mss. read \fbib{you received}} from us. \v{7}For you yourselves know what you must do to imitate us. We never lived in idleness\fnote{\fbackref{3:7} Or \fbib{We did not lead a disorderly life}} among you. \v{8}We did not eat anyone's food without paying for it. Instead, with toil and labor we worked night and day in order not to be a burden to any of you. \v{9}It is not as though we did not have that right, but we wanted\fnote{\fbackref{3:9} The Gk. lacks \fbib{we wanted}} to give you an example to follow. \v{10}While we were with you, we gave this order: ``If anyone doesn't want to work, he shouldn't eat.''

\v{11}We hear that some of you are living in idleness.\fnote{\fbackref{3:11} Or \fbib{living disorderly lives}} You are not busy working\fnote{\fbackref{3:11} Lit. \fbib{ergazomenous} (working)}---you are busy interfering\fnote{\fbackref{3:11} Lit. \fbib{periergazomenous} (uselessly working)} in other people's lives! \v{12}We order and encourage such people by the Lord Jesus, the Messiah,\fnote{\fbackref{3:12} Or \fbib{Christ}} to do their work quietly and to earn their own living. \v{13}Brothers, do not get tired of doing what is right.

\v{14}If anyone does not obey what we say\fnote{\fbackref{3:14} Lit. \fbib{our word}} in this letter, take note of him. Have nothing to do with him so that he will feel ashamed. \v{15}Yet, don't treat him like an enemy, but warn\fnote{\fbackref{3:15} Or \fbib{instruct}} him like a brother. \v{16}Now may the Lord of peace give you his peace at all times and in every way. May the Lord be with all of you.
\passage{Final Greeting}

\v{17}I, Paul, am writing this greeting with my own hand. This is the mark in every letter of mine. It is the way I write. \v{18}May the grace of our Lord Jesus, the Messiah,\fnote{\fbackref{3:18} Or \fbib{Christ}} be with all of you. Amen.\fnote{\fbackref{3:18} Other mss. lack \fbib{Amen}}
