\bookheader{Titus}
\labelbook{Titus}

\bookpretitle{The Letter from Paul to}
\booktitle{Titus}

\labelchapt{1}
\passage{Greetings}

\chapt{1}
\v{1}From:\fnote{\fbackref{1:1} The Gk. lacks \fbib{From}} Paul, a servant of God, and also an apostle of Jesus the Messiah,\fnote{\fbackref{1:1} Or \fbib{Christ}} to bring the faith to those chosen by God, along with full knowledge of the truth that leads to\fnote{\fbackref{1:1} Lit. \fbib{that is according to}} godliness, \v{2}which is based on the hope of eternal life that God, who cannot lie, promised before the world\fnote{\fbackref{1:2} Or \fbib{the ages}} began. \v{3}At the right time he revealed his message through the proclamation that was entrusted to me by the command of God our Savior.

\v{4}To: Titus, a genuine child in the faith that we share.

May grace and peace\fnote{\fbackref{1:4} Other mss. read \fbib{grace, mercy, and peace}} from God the Father and the Messiah,\fnote{\fbackref{1:4} Or \fbib{Christ}} Jesus our Savior, be yours!
\passage{Qualifications for Leaders in the Church}

\v{5}The reason I left you in Crete was to complete what still needed to be done and to appoint elders in every city, as I myself commanded you. \v{6}An elder must be\fnote{\fbackref{1:6} Lit. \fbib{If anyone is}} blameless. He must be the husband of one wife\fnote{\fbackref{1:6} Or \fbib{devoted to his wife}; lit. \fbib{a man of one woman}} and have children who are believers and who are not accused of having wild lifestyles or of being rebellious. \v{7}Because an overseer is God's servant manager, he must be blameless. He must not be arrogant or irritable. He must not drink too much, be a violent person, or make money in shameful ways. \v{8}Instead, he must be hospitable to strangers, must appreciate what is good, and be sensible, honest, moral, and self-controlled. \v{9}He must be devoted to the trustworthy message that agrees with what we teach, so that he may be able to encourage others with healthy doctrine and refute those who oppose it.
\passage{Guard What is True}

\v{10}For there are many people who are rebellious, especially those who are converts from Judaism.\fnote{\fbackref{1:10} Lit. \fbib{those of the circumcision}} They speak utter nonsense and deceive people. \v{11}They must be silenced, because they are the kind of people who ruin whole families by teaching what they should not teach in order to make money in a shameful way. \v{12}One of their very own prophets said,

\begin{poetry}
\poeml ``Liars ever, men of Crete, \\
\poemll    savage brutes that live to eat.''\fnote{\fbackref{1:12} Epimenides (6\textsuperscript{th} to 5\textsuperscript{th} century BC)}
\end{poetry}

\v{13}That statement is true. For this reason, refute them sharply so that they may become healthy in the faith \v{14}and not pay attention to Jewish myths or commands given by people who reject the truth. \v{15}Everything is clean to those who are clean, but nothing is clean to those who are corrupt and unbelieving. Indeed, their very way of thinking and their consciences have been corrupted. \v{16}They claim to know God, but they deny him by their actions. They are detestable, disobedient, and disqualified to do anything good.
\labelchapt{2}
\passage{Guidelines for Christian Living}

\chapt{2}
\v{1}But as for you, teach what is consistent with healthy doctrine. \v{2}Older men are to be sober, serious, sensible, and sound in faith, love, and endurance. \v{3}Likewise, older women are to show their reverence for God by their behavior. They are not to be gossips or addicted to alcohol, but to be examples\fnote{\fbackref{2:3} Or \fbib{teachers}} of goodness. \v{4}They should encourage the younger women to love their husbands, to love their children, \v{5}to be sensible and pure, to manage their households, to be kind, and to submit themselves to their husbands. Otherwise, the word of God may be discredited.\fnote{\fbackref{2:5} Or \fbib{blasphemed}}

\v{6}Likewise, encourage the younger men to be sensible. \v{7}Always set an example for others by doing good actions. Teach with integrity and dignity. \v{8}Use wholesome speech that cannot be condemned. Then any opponent will be ashamed because he cannot say anything bad about us.

\v{9}Slaves are to submit to their masters in everything, aiming to please them and not argue with them \v{10}or steal from them. Instead, they are to show complete and perfect loyalty, so that in every way they may make the teaching about God our Savior more attractive.

\v{11}For the grace of God has appeared, bringing salvation to all people. \v{12}It trains us to renounce ungodly living and worldly passions so that we might live sensible, honest, and godly lives in the present age \v{13}as we wait for the blessed hope and glorious appearance of our great God and Savior, Jesus the Messiah.\fnote{\fbackref{2:13} Or \fbib{Christ}} \v{14}He gave himself for us to set us free from every wrong and to cleanse us so that we could be his special people who are enthusiastic about doing good deeds.

\v{15}These are the things you should teach. Encourage and refute with full authority. Do not let anyone look down on you.
\labelchapt{3}
\passage{Concentrate on Doing What is Good}

\chapt{3}
\v{1}Remind believers\fnote{\fbackref{3:1} Lit. \fbib{them}} to submit to rulers and authorities, to be obedient, and to be ready to do any honorable kind of work. \v{2}They are not to insult\fnote{\fbackref{3:2} Or \fbib{slander}} anyone or be argumentative. Instead, they are to be gentle and perfectly courteous to everyone. \v{3}After all, we ourselves were once foolish, disobedient, and misled. We were slaves to many kinds of lusts and pleasures, spending our days in malice and jealousy. We were despised, and we hated one another.

\begin{poetry}
\poeml \v{4}In grace our Savior God appeared, \\
\poemll    to make his love for mankind clear. \\
\poeml \v{5}`Twas not for deeds that we had done, \\
\poemll    but by his steadfast love\fnote{\fbackref{3:5} Or \fbib{his mercy}} alone, \\
\poeml he saved us through a second birth, \\
\poemll    renewed us by the Spirit's\fnote{\fbackref{3:5} Lit. \fbib{the Holy Spirit's}} work, \\
\poeml \v{6}and poured him out upon us, too, \\
\poemll    through Jesus the Messiah\fnote{\fbackref{3:6} Or \fbib{Christ}} our Savior true. \\
\poeml \v{7}And so, made heirs by his own grace, \\
\poemll    eternal life we now embrace.\fnote{\fbackref{3:7} Lit. \fbib{we have become heirs according to the hope of eternal life}}
\end{poetry}

\v{8}This saying is trustworthy.\fnote{\fbackref{3:8} This formula accompanied early Christian sayings on which full reliance could be placed.} I want you to insist on these things, so that those who have put their faith in God may devote themselves to good actions. These things are good and helpful to other people.

\v{9}But avoid foolish controversies, arguments about genealogies, quarrels, and fights about the Law. These things are useless and worthless. \v{10}Have nothing to do with a divisive person after you have warned him once or twice. \v{11}For you know that a person like this is corrupt and keeps on sinning, being self-condemned.
\passage{Final Instructions to Titus}

\v{12}As soon as I send Artemas to you, or perhaps Tychicus, do your best to come to me at Nicopolis, for I have decided to spend the winter there. \v{13}Do all you can to send Zenas the expert in the Law and Apollos on their way, and see that they have everything they need. \v{14}Our own people should also learn to make good deeds a priority when urgent needs arise, so they won't be unproductive.
\passage{Final Greeting}

\v{15}All who are with me send you greetings. Greet our fellow believers who love us. May grace be with all of you! Amen.\fnote{\fbackref{3:15} Other mss. lack \fbib{Amen}}
