\bookheader{1 Peter}
\labelbook{1Pet}

\bookpretitle{The Letter Called}
\booktitle{First Peter}

\labelchapt{1}
\passage{Greetings}

\chapt{1}
\v{1}From:\fnote{\fbackref{1:1} The Gk. lacks \fbib{From}} Peter, an apostle of Jesus, the Messiah.\fnote{\fbackref{1:1} Or \fbib{Christ}}

To: The exiles of the Dispersion\fnote{\fbackref{1:1} I.e. the Jewish communities outside of Israel} in Pontus, Galatia, Cappadocia, Asia, and Bithynia, \v{2}the people chosen according to the foreknowledge of God the Father through the sanctifying action of the Spirit to be obedient to Jesus, the Messiah,\fnote{\fbackref{1:2} Or \fbib{Christ}} and to be sprinkled with his blood.

May grace and peace be yours in abundance!
\passage{Our Hope and Joy are in the Messiah}

\v{3}Blessed be the God and Father of our Lord Jesus, the Messiah!\fnote{\fbackref{1:3} Or \fbib{Christ}} Because of his great mercy he has granted us a new birth, resulting in an immortal hope through the resurrection of Jesus, the Messiah,\fnote{\fbackref{1:3} Or \fbib{Christ}} from the dead \v{4}and to an inheritance kept in heaven for you that can't be destroyed, corrupted, or changed. \v{5}Through faith you are being protected by God's power for a salvation that is ready to be revealed at the end of this era. \v{6}You greatly rejoice in this, even though you have to suffer various kinds of trials for a little while, \v{7}so that your genuine faith, which is more valuable than gold that perishes when tested by fire, may result in praise, glory, and honor when Jesus, the Messiah,\fnote{\fbackref{1:7} Or \fbib{Christ}} is revealed.

\v{8}Though you have not seen\fnote{\fbackref{1:8} Other mss. read \fbib{known}} him, you love him. And even though you do not see him now, you believe in him and rejoice with an indescribable and glorious joy, \v{9}because you are receiving the goal of your faith, the salvation of your souls.

\v{10}Even the prophets, who prophesied about the grace that was to be yours, carefully researched and investigated this salvation. \v{11}They tried to find out what era or specific time the Spirit of the Messiah\fnote{\fbackref{1:11} Or \fbib{Christ}} in them kept referring to when he predicted the sufferings of the Messiah\fnote{\fbackref{1:11} Or \fbib{Christ}} and the glories that would follow. \v{12}It was revealed to them that they were not serving themselves but you in regard to the things that have now been announced to you by those who brought you the good news through the Holy Spirit sent from heaven. These are things that even the angels desire to look into.
\passage{Be Holy}

\v{13}Therefore, prepare your minds for action, keep a clear head, and set your hope completely on the grace to be given you when Jesus, the Messiah,\fnote{\fbackref{1:13} Or \fbib{Christ}} is revealed. \v{14}As obedient children, do not be shaped by the desires that used to influence you when you were ignorant. \v{15}Instead, be holy in every aspect of your life, just as the one who called you is holy. \v{16}For it is written, ``You must be holy, because I am holy.''\fnote{\fbackref{1:16} Cf. Lev 11:44-45; Lev 19:2}

\v{17}If you call ``Father'' the one who judges everyone impartially according to what they have done, you must live in reverent fear as long as you are strangers in a strange land. \v{18}For you know that it was not with perishable things like silver or gold that you have been ransomed from the worthless way of life handed down to you by your ancestors, \v{19}but with the precious blood of the Messiah,\fnote{\fbackref{1:19} Or \fbib{Christ}} like that of a lamb without blemish or defect. \v{20}On the one hand, he was foreknown before the creation\fnote{\fbackref{1:20} Or \fbib{foundation}} of the world, but on the other hand, he was revealed at the end of time for your sake. \v{21}Through him you believe in God, who raised him from the dead and gave him glory, so that your faith and hope might be in God.
\passage{Love One Another}

\v{22}Now that you have obeyed the truth\fnote{\fbackref{1:22} Other mss. read \fbib{the truth through the Spirit}} and have purified your souls to love your brothers sincerely, you must love one another intensely and with a pure heart. \v{23}For you have been born again, not by a seed that perishes but by one that cannot perish---by the living and everlasting word of God.\fnote{\fbackref{1:23} Or \fbib{by the word of the living and everlasting God}} \v{24}For

\begin{poetry}
\poeml ``All human life\fnote{\fbackref{1:24} Lit. \fbib{all flesh}} is like grass, \\
\poemll    and all its glory is like a flower in the grass. \\
\poeml The grass dries up and the flower drops off, \\
\poeml \v{25}but the word of the Lord\fnote{\fbackref{1:25} MT source citation reads \fbib{}\divine{Lord}} lasts forever.''\fnote{\fbackref{1:25} Cf. Isa 40:6-8}
\end{poetry}

Now this word is the good news that was announced to you.
\labelchapt{2}
\passage{Live as God's Chosen People}

\chapt{2}
\v{1}Therefore, rid yourselves of every kind of evil and deception, hypocrisy, jealousy, and every kind of slander. \v{2}Like newborn babies, thirst for the pure milk of the word so that by it you may grow in your salvation. \v{3}Surely you have tasted that the Lord is good!

\v{4}As you come to him, the living stone who was rejected by people but was chosen and precious in God's sight, \v{5}you, too, as living stones, are building yourselves up into a spiritual house and a holy priesthood, so that you may offer spiritual sacrifices that are acceptable to God through Jesus, the Messiah.\fnote{\fbackref{2:5} Or \fbib{Christ}} \v{6}This is why it says in Scripture:

\begin{poetry}
\poeml ``Look! I am laying a chosen, precious cornerstone\fnote{\fbackref{2:6} Or \fbib{capstone}} in Zion. \\
\poemll    The one who believes in him will never be ashamed.''\fnote{\fbackref{2:6} Cf. Isa 28:16}
\end{poetry}

\v{7}Therefore he is precious to you who believe, but to those who do not believe,

\begin{poetry}
\poeml ``The stone that the builders rejected \\
\poemll    has become the cornerstone,\fnote{\fbackref{2:7} Or \fbib{capstone}} \\
\poeml \v{8}a stone they stumble over \\
\poemll    and a rock they trip on.''\fnote{\fbackref{2:8} Cf. Ps 118:22; Isa 8:14}
\end{poetry}

They keep on stumbling because they disobey the word, as they were destined to do.

\v{9}But you are a chosen people, a royal priesthood, a holy nation, a people to be his very own and to proclaim the wonderful deeds\fnote{\fbackref{2:9} Or \fbib{the excellence}; cf. Isa 43:21} of the one who called you out of darkness into his marvelous light.

\begin{poetry}
\poeml \v{10}Once you were not a people, \\
\poemll    but now you are the people of God. \\
\poeml Once you had not received mercy, \\
\poemll    but now you have received mercy.
\end{poetry}
\passage{Live as God's Servants}

\v{11}Dear friends, I urge you as aliens and exiles to keep on abstaining from the desires of the flesh that wage war against the soul. \v{12}Continue to live such upright lives among the gentiles that, when they slander you as practicers of evil, they may see your good actions and glorify God when he visits them.\fnote{\fbackref{2:12} Lit. \fbib{God on the day of visitation}}

\v{13}For the Lord's sake submit yourselves to every human authority: whether to the king as supreme, \v{14}or to governors who are sent by him to punish those who do wrong and to praise those who do right. \v{15}For it is God's will that by doing right you should silence the ignorant talk\fnote{\fbackref{2:15} Lit. \fbib{the ignorance}} of foolish people. \v{16}Live like free people, and do not use your freedom as an excuse for doing evil. Instead, be God's servants. \v{17}Honor everyone. Keep on loving the community of believers,\fnote{\fbackref{2:17} Lit. \fbib{the brotherhood}} fearing God, and honoring the king.
\passage{Suffer Patiently}

\v{18}You household servants must submit yourselves to your masters out of respect, not only to those who are kind and fair, but also to those who are unjust. \v{19}For it is a fine thing if, when moved by your conscience to please God, you suffer patiently when wronged. \v{20}What good does it do if, when you sin, you patiently receive punishment for it? But if you suffer for doing good and receive it patiently, you have God's approval. \v{21}This is, in fact, what you were called to do, because:

\begin{poetry}
\poeml The Messiah\fnote{\fbackref{2:21} Or \fbib{Christ}} also suffered for you \\
\poemll    and left an example for you \\
\poemlll       to follow in his steps. \\
\poeml \v{22}``He never sinned, \\
\poemll    and he never told a lie.''\fnote{\fbackref{2:22} Lit. \fbib{and no deceit was found in his mouth}; cf. Isa 53:9} \\
\poeml \v{23}When he was insulted, \\
\poemll    he did not retaliate. \\
\poeml When he suffered, \\
\poemll    he did not threaten. \\
\poeml It was his habit \\
\poemll    to commit the matter to the one who judges fairly. \\
\poeml \v{24}``He himself bore our sins''\fnote{\fbackref{2:24} Cf. Isa 53:4,12} in his body on the tree, \\
\poemll    so that we might die to those sins \\
\poemlll       and live righteously. \\
\poeml ``By his wounds \\
\poemll    you have been healed.''\fnote{\fbackref{2:24} Cf. Isa 53:5} \\
\poeml \v{25}You were ``like sheep that kept going astray,''\fnote{\fbackref{2:25} Cf. Isa 53:6} \\
\poemll    but now you have returned to the shepherd \\
\poemlll       and overseer of your souls.
\end{poetry}
\labelchapt{3}
\passage{Wives and Husbands}

\chapt{3}
\v{1}In a similar way, you wives must submit yourselves to your husbands so that, even if some of them refuse to obey the word, they may be won over without a word through your conduct as wives \v{2}when they see your pure and reverent lives.

\v{3}Your beauty should not be an external one, consisting of braided hair or the wearing of gold ornaments and dresses. \v{4}Instead, it should be the inner disposition of the heart, consisting in the imperishable quality of a gentle and quiet spirit, which God values greatly.\fnote{\fbackref{3:4} Lit. \fbib{which is of great value before God}} \v{5}After all, this is how holy women who set their hope on God used to make themselves beautiful in the past. They submitted themselves to their husbands, \v{6}just as Sarah obeyed Abraham and called him lord. You have become her daughters by doing good and by not letting anything terrify you.

\v{7}In a similar way, you husbands must live with your wives in an understanding manner, as with a most delicate partner.\fnote{\fbackref{3:7} Lit. \fbib{with the weaker vessel}} Honor them as heirs with you of the gracious gift of life, so that nothing may interfere with your prayers.
\passage{When You are Wronged}

\v{8}Finally, all of you must live in harmony, be sympathetic, love as brothers, and be compassionate and humble. \v{9}Do not pay others back evil for evil or insult for insult. Instead, keep blessing them, because you were called to inherit a blessing.

\begin{poetry}
\poeml \v{10}``For the person who wants to love life \\
\poemll    and see good days \\
\poeml must keep his tongue from evil \\
\poemll    and his lips from speaking deceit. \\
\poeml \v{11}He must turn away from evil and do good. \\
\poemll    He must seek peace and pursue it. \\
\poeml \v{12}For the Lord\fnote{\fbackref{3:12} MT source citation reads \fbib{}\divine{Lord}} watches the righteous,\fnote{\fbackref{3:12} Lit. \fbib{the eyes of the Lord are on the righteous}} \\
\poemll    and he pays attention to their prayers.\fnote{\fbackref{3:12} Lit. \fbib{and his ears are attentive to their prayer}} \\
\poeml But the Lord\fnote{\fbackref{3:12} MT source citation reads \fbib{}\divine{Lord}} opposes those\fnote{\fbackref{3:12} Lit. \fbib{the face of the Lord is against}} who do wrong.''\fnote{\fbackref{3:12} Cf. Ps 34:12-16}
\end{poetry}

\v{13}Who will harm you if you are devoted to doing what is good? \v{14}But even if you should suffer for doing what is right, you are blessed. ``Never be afraid of their threats, and never get upset. \v{15}Instead, exalt\fnote{\fbackref{3:15} Or \fbib{set apart}} the Messiah''\fnote{\fbackref{3:15} Or \fbib{Christ}; cf. Isa 8:12} as Lord in your lives.\fnote{\fbackref{3:15} Lit. \fbib{hearts}} Always be prepared to give a defense to everyone who asks you to explain the hope you have. \v{16}But do this\fnote{\fbackref{3:16} The Gk. lacks \fbib{do this}} gently and respectfully, keeping a clear conscience, so that those who speak evil of your good conduct in the Messiah\fnote{\fbackref{3:16} Or \fbib{Christ}} will be ashamed of slandering you. \v{17}After all, if it is the will of God, it is better to suffer for doing right than for doing wrong.

\begin{poetry}
\poeml \v{18}For the Messiah\fnote{\fbackref{3:18} Or \fbib{Christ}} also suffered\fnote{\fbackref{3:18} Other mss. read \fbib{died}} for sins once for all, \\
\poemll    an innocent person for the guilty, \\
\poemlll       so that he could bring you\fnote{\fbackref{3:18} Other mss. read \fbib{us}} to God. \\
\poeml He was put to death in a mortal body \\
\poemll    but was brought to life by the Spirit,
\end{poetry}

\v{19}in which he went and made a proclamation to those imprisoned spirits \v{20}who disobeyed long ago in the days of Noah, when God waited patiently while the ark was being built. In it a few, that is, eight persons, were saved by water. \v{21}Baptism, which is symbolized by that water, now saves you also, not by removing dirt from the body, but by asking God for a clear\fnote{\fbackref{3:21} Lit. \fbib{clean}} conscience based on the resurrection of Jesus, the Messiah,\fnote{\fbackref{3:21} Or \fbib{Christ}} \v{22}who has gone to heaven and is at the right hand of God, where angels, authorities, and powers have been made subject to him.
\labelchapt{4}
\passage{Good Managers of God's Grace}

\chapt{4}
\v{1}Therefore, since the Messiah\fnote{\fbackref{4:1} Or \fbib{Christ}} suffered in a mortal body,\fnote{\fbackref{4:1} Other mss. read \fbib{suffered for us}; still other mss. read \fbib{suffered for you}} you, too, must arm yourselves with the same determination, because the person who has suffered in a mortal body has stopped sinning, \v{2}so that he can live the rest of his mortal life\fnote{\fbackref{4:2} Lit. \fbib{time}} guided, not by human desires, but by the will of God. \v{3}For you spent enough time in the past doing what the gentiles like to do, living in sensuality, sinful desires, drunkenness, wild celebrations, drinking parties, and detestable idolatry. \v{4}They insult you now because they are surprised that you are no longer joining them in the same excesses of wild living. \v{5}They will give an account to the one who is ready to judge the living and the dead. \v{6}Indeed, this is why the gospel was proclaimed even to those who have died, so that they could be judged in their mortal flesh like all humans and live in the spiritual realm like God.

\v{7}Because everything will soon come to an end, be sensible and clear-headed, so you can pray. \v{8}Above all, continue to love each other deeply, because love covers a multitude of sins. \v{9}Show hospitality to one another without complaining. \v{10}As good servant managers of God's grace in its various forms, serve one another with the gift each of you has received. \v{11}Whoever speaks must speak God's words.\fnote{\fbackref{4:11} Lit. \fbib{If anyone speaks as the words of God}} Whoever serves must serve with the strength\fnote{\fbackref{4:11} Lit. \fbib{Whoever serves as with the strength}} that God supplies, so that in every way God may be glorified through Jesus, the Messiah.\fnote{\fbackref{4:11} Or \fbib{Christ}} Glory and power belong to him forever and ever! Amen.
\passage{Suffering as a Christian}

\v{12}Dear friends, do not be surprised by the fiery ordeal that is taking place among you to test you, as though something strange were happening to you. \v{13}Instead, because you are participating in the sufferings of the Messiah,\fnote{\fbackref{4:13} Or \fbib{Christ}} keep on rejoicing, so that you may be glad and shout for joy when his glory is revealed. \v{14}If you are insulted because of the name of the Messiah,\fnote{\fbackref{4:14} Or \fbib{Christ}} you are blessed, for the glorious Spirit of God is resting on you.\fnote{\fbackref{4:14} Other mss. read \fbib{on you. For their sake he is being blasphemed, but for your sake he is being glorified.}}

\v{15}Of course, none of you should suffer for being a murderer, thief, criminal, or troublemaker. \v{16}But if you suffer for being a Christian, do not feel ashamed, but glorify God with that name. \v{17}For the time has come for judgment to begin with the household of God. And if it begins with us, what will be the outcome for those who refuse to obey the gospel of God?

\begin{poetry}
\poeml \v{18}``If it is hard for the righteous person to be saved, \\
\poemll    what will happen to the ungodly and sinful person?''\fnote{\fbackref{4:18} Cf. Prov 11:31 (LXX)}
\end{poetry}

\v{19}So then, those who suffer according to God's will should entrust their souls to their faithful Creator and continue to do what is good.
\labelchapt{5}
\passage{Be Shepherds of God's Flock}

\chapt{5}
\v{1}Therefore, as a fellow elder, a witness of the Messiah's\fnote{\fbackref{5:1} Or \fbib{Christ's}} sufferings, and one who shares in the glory to be revealed, I appeal to the elders among you: \v{2}Be shepherds of God's flock that is among you, watching over it, not because you must but because you want to, and not greedily but eagerly, as God desires. \v{3}Do not lord it over the people entrusted to you, but be examples to the flock. \v{4}Then, when the Chief Shepherd appears, you will receive the victor's crown of glory that will never fade away.
\passage{Be Humble and Alert}

\v{5}In a similar way, you young people must submit to the elders.\fnote{\fbackref{5:5} Or \fbib{to those who are older}} All of you must clothe yourselves with humility for the sake of each other, because:

\begin{poetry}
\poeml ``God opposes the arrogant, \\
\poemll    but gives grace to the humble.''\fnote{\fbackref{5:5} Cf. Prov 3:34 (LXX)}
\end{poetry}

\v{6}Therefore, humble yourselves under the mighty hand of God, so that at the proper time he may exalt you. \v{7}Throw all your worry on him, because he cares for you. \v{8}Be clear-minded and alert. Your opponent, the devil, is prowling around like a roaring lion, looking for someone to devour. \v{9}Resist him and be firm in the faith, because you know that your brothers throughout the world are undergoing the same kinds of suffering. \v{10}After you have suffered for a little while, the God of all grace, who called you by the Messiah\fnote{\fbackref{5:10} Or \fbib{Christ}} Jesus\fnote{\fbackref{5:10} Other mss. lack \fbib{Jesus}} to his eternal glory, will restore you, establish you, strengthen you, and support you. \v{11}Power belongs\fnote{\fbackref{5:11} Other mss. read \fbib{Glory and power belong}} to him forever and ever! Amen.
\passage{Final Greeting}

\v{12}Through Silvanus, whom I regard as a faithful brother, I have written this short letter to encourage you and to testify that this is to be the true grace of God in which you are to stand firm! \v{13}Your sister church\fnote{\fbackref{5:13} Lit. \fbib{She who is}} in Babylon, chosen along with you, sends you greetings, as does Mark, whom I regard as a son. \v{14}Greet one another with a loving kiss. Peace be to all of you who are in the Messiah!\fnote{\fbackref{5:14} Or \fbib{Christ}; other mss. read \fbib{the Messiah Jesus! Amen}}
