\bookheader{2 Peter}
\labelbook{2Pet}

\bookpretitle{The Letter Called}
\booktitle{Second Peter}

\labelchapt{1}
\passage{Greetings}

\chapt{1}
\v{1}From:\fnote{\fbackref{1:1} The Gk. lacks \fbib{From}} Simeon\fnote{\fbackref{1:1} Other mss. read \fbib{Simon}} Peter, a servant\fnote{\fbackref{1:1} Or \fbib{slave}} and apostle of Jesus, the Messiah.\fnote{\fbackref{1:1} Or \fbib{Christ}}

To: Those who have received faith that is as valuable as ours through the righteousness of our God and Savior, Jesus the Messiah.\fnote{\fbackref{1:1} Or \fbib{Christ}}

\v{2}May grace and peace be yours in abundance through full knowledge of God and of Jesus our Lord!
\passage{We are Called to Holy Living}

\v{3}His divine power has given us everything we need for life and godliness through the full knowledge of the one who called us by his own glory and excellence. \v{4}Through these he has given us his precious and wonderful promises, so that through them you may participate in the divine nature, seeing that you have escaped the corruption that is in the world caused by evil desires. \v{5}For this very reason, you must make every effort to supplement your faith with moral character, your moral character with knowledge, \v{6}your knowledge with self-control, your self-control with endurance, your endurance with godliness, \v{7}your godliness with brotherly kindness, and your brotherly kindness with love. \v{8}For if you possess these qualities, and if they continue to increase among you, they will keep you from being ineffective and unproductive in attaining a full knowledge of our Lord Jesus, the Messiah.\fnote{\fbackref{1:8} Or \fbib{Christ}} \v{9}For the person who lacks these qualities is blind and shortsighted, and has forgotten the cleansing that he has received from his past sins.

\v{10}So then, my brothers, be all the more eager to make your calling and election certain, for if you keep on doing this you will never fail. \v{11}For in this way you will be generously granted entry into the eternal kingdom of our Lord and Savior Jesus, the Messiah.\fnote{\fbackref{1:11} Or \fbib{Christ}}

\v{12}Therefore, I intend to keep on reminding you about these things, even though you already know them and are firmly established in the truth that you now have. \v{13}Yet I think it is right to refresh your memory as long as I am living in this bodily tent, \v{14}because I know that the removal of my bodily tent will come soon, as indeed our Lord Jesus, the Messiah,\fnote{\fbackref{1:14} Or \fbib{Christ}} has shown me. \v{15}And I will make every effort to see that you will always remember these things after I am gone.
\passage{Pay Attention to God's Word}

\v{16}When we told you about the power and coming of our Lord Jesus, the Messiah,\fnote{\fbackref{1:16} Or \fbib{Christ}} we did not follow any clever myths. Rather, we were eyewitnesses of his majesty. \v{17}For he received honor and glory from God the Father when these words from the Majestic Glory were spoken about him: ``This is my Son, whom I love. I am pleased with him.'' \v{18}We ourselves heard this voice that came from heaven when we were with him on the holy mountain. \v{19}Therefore we regard the message of the prophets as confirmed beyond doubt, and you will do well to pay attention to it, as to a lamp that is shining in a gloomy place, until the day dawns and the morning star rises in your hearts. \v{20}First of all, you must understand this: No prophecy in Scripture is a matter of one's own interpretation, \v{21}because no prophecy ever originated through a human decision. Instead, men spoke from God as they were carried along by the Holy Spirit.
\labelchapt{2}
\passage{Warning against False Teachers}

\chapt{2}
\v{1}Now there were false prophets among the people, just as there also will be false teachers among you, who will secretly introduce destructive heresies and even deny the Master who bought them, bringing swift destruction on themselves. \v{2}Many people will follow their immoral ways, and because of them the way of truth will be maligned.\fnote{\fbackref{2:2} Or \fbib{blasphemed}} \v{3}In their greed they will exploit you with deceptive words. The ancient verdict against them is still in force, and their destruction is not delayed.\fnote{\fbackref{2:3} Lit. \fbib{asleep}}

\v{4}For if God did not spare angels when they sinned, but threw them into the lowest hell\fnote{\fbackref{2:4} Gk. \fbib{Tartarus}; a reference to the realm of the dead} and imprisoned them in chains\fnote{\fbackref{2:4} Other mss. read \fbib{pits}} of deepest darkness, holding them for judgment; \v{5}and if he did not spare the ancient world but protected Noah, a righteous preacher, along with seven others when he brought the flood on the world of ungodly people; \v{6}and if he condemned the cities of Sodom and Gomorrah and destroyed them by burning them to ashes, making them an example to ungodly people of what is going to happen to them; \v{7}and if he rescued Lot, a righteous man who was greatly distressed by the immoral conduct of lawless people--- \v{8}for as long as that righteous man lived among them, day after day he was being tortured in his righteous soul by what he saw and heard in their lawless actions--- \v{9}then the Lord knows how to rescue godly people from their trials and to hold unrighteous people for punishment on the day of judgment, \v{10}especially those who satisfy their flesh by indulging in its passions and who despise authority.

Being bold and arrogant, they are not afraid to slander glorious beings. \v{11}Yet even angels, although they are greater in strength and power, do not bring a slanderous accusation against them from the Lord. \v{12}These people, like irrational animals, are mere creatures of instinct that are born to be caught and killed. They insult what they don't understand, and like animals they, too, will be destroyed, \v{13}suffering harm as punishment for their wrongdoing. They take pleasure in wild parties in broad daylight. They are stains and blemishes, reveling in their deceitful pleasures\fnote{\fbackref{2:13} Other mss. read \fbib{in their love feasts}} while they eat with you. \v{14}With eyes full of adultery, they cannot get enough of sin. They seduce unsteady souls and have had their hearts expertly trained in greed. They are doomed to a curse.\fnote{\fbackref{2:14} Lit. \fbib{children of a curse}} \v{15}They have left the straight path and wandered off to follow the path of Bosor's\fnote{\fbackref{2:15} Other mss. read \fbib{Beor's}} son Balaam, who loved the reward he got for doing wrong. \v{16}But he was rebuked for his offense. A donkey that normally cannot talk spoke with a human voice and restrained the prophet's insanity.

\v{17}These men are dried-up springs, mere clouds driven by a storm. Gloomy darkness is reserved for them. \v{18}By talking high-sounding nonsense and using sinful cravings of the flesh, they entice people who have just escaped from those who live in error. \v{19}Promising them freedom, they themselves are slaves to depravity, for a person is a slave to whatever conquers him.

\v{20}For if, after escaping the world's corruptions through a full knowledge of our Lord and Savior Jesus, the Messiah,\fnote{\fbackref{2:20} Or \fbib{Christ}} they are again entangled and conquered by those corruptions,\fnote{\fbackref{2:20} Lit. \fbib{by them}} then their last condition is worse than their former one. \v{21}It would have been better for them not to have known the way of righteousness than to know it and turn their backs on the holy commandment that was committed to them. \v{22}The proverb is true that describes what has happened to them: ``A dog returns to its vomit,''\fnote{\fbackref{2:22} Cf. Prov 26:11} and ``A pig that is washed goes back to wallow in the mud.''\fnote{\fbackref{2:22} The source of this quotation appears to be the Syrian \fbib{Story of Ahikar}.}
\labelchapt{3}
\passage{Be Ready for the Day of the Lord}

\chapt{3}
\v{1}Dear friends, this is now the second of two letters\fnote{\fbackref{3:1} Lit. \fbib{second letter}} I am writing to you, in which I have been trying to stimulate your pure minds by reminding you \v{2}to recall the words spoken in the past by the holy prophets and the commandment of our Lord and Savior spoken\fnote{\fbackref{3:2} The Gk. lacks \fbib{spoken}} through your apostles.

\v{3}First of all you must understand this: In the last days mockers will come and, following their own desires, will ridicule us\fnote{\fbackref{3:3} Cf. Jude 18} \v{4}by saying, ``What happened to the Messiah's\fnote{\fbackref{3:4} Lit. \fbib{to his}} promise to return? Ever since our ancestors died,\fnote{\fbackref{3:4} Lit. \fbib{fell asleep}} everything continues as it did from the beginning of creation.'' \v{5}But they deliberately ignore the fact that long ago the heavens existed and the earth was formed by God's word out of water and with water, \v{6}by which the world at that time was deluged with water and destroyed. \v{7}Now by that same word, the present heavens and earth have been reserved for fire and are being kept for the day when ungodly people will be judged and destroyed.

\v{8}Don't forget this fact, dear friends: With the Lord a single day is like a thousand years, and a thousand years are like a single day. \v{9}The Lord is not slow about his promise, as some people understand slowness, but is being patient with you. He does not want anyone to perish, but wants\fnote{\fbackref{3:9} The Gk. lacks \fbib{wants}} everyone to repent. \v{10}But the Day of the Lord will come like a thief. On that day\fnote{\fbackref{3:10} Lit. \fbib{On it}} the heavens will disappear with a roaring sound, the elements will be destroyed by fire, and the earth and everything done on it will be exposed.

\v{11}Since everything will be destroyed in this way, think of the kind of holy and godly people you ought to be \v{12}as you look forward to and hasten the coming of the day of God, when the heavens will be set ablaze and dissolved and the elements will melt with fire. \v{13}But in keeping with his promise, we are looking forward to new heavens and a new earth, where righteousness is at home.

\v{14}So then, dear friends, since you are looking forward to this, make every effort to have the Lord\fnote{\fbackref{3:14} Lit. \fbib{have him}} find you at peace and without spot or fault. \v{15}Think of our Lord's patience as facilitating salvation, just as our dear brother Paul also wrote to you according to the wisdom given him. \v{16}He speaks about this subject in all his letters. Some things in them are hard to understand, which ignorant and unstable people distort, leading to their own destruction, as they do the rest of the Scriptures.

\v{17}And so, dear friends, since you already know these things, continuously be on your guard not to be carried away by the deception of lawless people. Otherwise, you may\fnote{\fbackref{3:17} Lit. \fbib{people, lest you}} fall from your secure position. \v{18}Instead, continue to grow in the grace and knowledge of our Lord and Savior Jesus, the Messiah.\fnote{\fbackref{3:18} Or \fbib{Christ}} Glory belongs to him both now and on that eternal day! Amen.\fnote{\fbackref{3:18} Other mss. lack \fbib{Amen}}
