\bookheader{2 John}
\labelbook{2John}

\bookpretitle{The Letter Called}
\booktitle{Second John}

\passage{Greetings from John}

\v{1}From:\fnote{The Gk. lacks \fbib{From}} The Elder

To: The chosen lady and her children, whom I genuinely love, and not only I but also all who know the truth, \v{2}that is present in us and will be with us forever.

\v{3}Grace, mercy, and peace will be with us from God the Father and from Jesus\fnote{Other mss. read \fbib{the Lord Jesus}} the Messiah,\fnote{Or \fbib{Christ}} the Father's Son, in truth and love.
\passage{Living in the Truth}

\v{4}I was overjoyed to find some of your\fnote{Lit. \fbib{your} (sing.)} children living truthfully, just as the Father has commanded us. \v{5}Dear lady, I am now requesting of you\fnote{Lit. \fbib{you} (sing.)} that we all continue to love one another. It is not as though I am writing to give you\fnote{Lit. \fbib{you} (sing.)} a new commandment, but one that we have had from the beginning. \v{6}And this is what demonstrates\fnote{The Gk. lacks \fbib{what demonstrates}} love: that we live according to God's\fnote{Lit. \fbib{his}} commandments. Just as you\fnote{Lit. \fbib{you} (pl.)} have heard from the beginning what he commanded, you\fnote{Lit. \fbib{you} (pl.)} must live by it.
\passage{Reject False Teachers}

\v{7}For many deceivers have gone out into the world. They refuse to acknowledge Jesus the Messiah\fnote{Or \fbib{Christ}} as having become human. Any such person is a deceiver and an antichrist. \v{8}See\fnote{The Gk. verb is pl.} to it that you\fnote{Lit. \fbib{you} (pl.)} don't destroy what we have\fnote{Other mss. read \fbib{you have}} worked for, but that you\fnote{Lit. \fbib{you} (pl.)} receive your\fnote{Lit. \fbib{your} (pl.)} full reward. \v{9}Everyone who does not remain true to the teaching of the Messiah,\fnote{Or \fbib{Christ}} but goes beyond it, does not have God. The person who remains true to the teaching of the Messiah\fnote{Or \fbib{Christ}} has both the Father and the Son. \v{10}If anyone comes to you\fnote{Lit. \fbib{you} (pl.)} but does not present his teachings,\fnote{Lit. \fbib{this teaching}} do not receive\fnote{The Gk. verb is pl.} him into your house or even welcome\fnote{The Gk. verb is pl.} him, \v{11}because the one who welcomes him shares in his evil deeds.
\passage{Final Greeting}

\v{12}Although I have a great deal to write to you,\fnote{Lit. \fbib{you} (pl.)} I would prefer not to use paper and ink. Instead, I hope to come to you and talk face to face, so that our joy may be complete. \v{13}The children of your\fnote{Lit. \fbib{you} (sing.)} chosen sister greet you.\fnote{Other mss. read \fbib{you. Amen}}
