\bookheader{Jude}
\labelbook{Jude}

\bookpretitle{The Letter from}
\booktitle{Jude}

\passage{Greetings}

\v{1}From: Jude, a servant of Jesus the Messiah,\fnote{Or \fbib{Christ}} and yet a brother of James.

To: Those who have been called, who are loved\fnote{Other mss. read \fbib{sanctified}} by God the Father and kept safe by Jesus, the Messiah.\fnote{Or \fbib{Christ}}

\v{2}May mercy, peace, and love be yours in abundance!
\passage{Warning about False Teachers}

\v{3}Dear friends, although I was eager to write to you about the salvation we share, I found it necessary to write to you and urge you to continue your vigorous defense of the faith that was passed down to the saints once and for all. \v{4}For some people have slipped in among you unnoticed. They were written about long ago as being deserving of this condemnation because they are ungodly. They turn the grace of our God into uncontrollable lust and deny our only Master and Lord, Jesus the Messiah.\fnote{Or \fbib{Christ}}

\v{5}Now I want to remind you, even though you are fully aware of these things, that the Lord who once saved his people from the land of Egypt later destroyed those who did not believe. \v{6}He has also held in eternal chains those angels who did not keep their own position but abandoned their assigned place. They are held in deepest darkness for judgment on the great day.\fnote{I.e. the day of judgment} \v{7}Likewise, Sodom and Gomorrah and the cities near them, which like them committed sexual sins and pursued homosexual activities,\fnote{Lit. \fbib{pursued other flesh}} serve as an example of the punishment of eternal fire.

\v{8}In a similar way, these dreamers also defile their flesh, reject the Lord's authority,\fnote{Lit. \fbib{reject dominions}} and slander his glorious beings. \v{9}Even the archangel Michael, when he argued with the devil and fought over the body of Moses, did not dare to bring a slanderous accusation against him. Instead, he said, ``May the Lord rebuke you!''\fnote{This incident is possibly based on \fbib{The Assumption of Moses,} an apocryphal Jewish writing.} \v{10}Whatever these people do not understand, they slander. Like irrational animals, they are destroyed by the very things they know by instinct. \v{11}How terrible it will be for them! For they lived like Cain did\fnote{Lit. \fbib{they followed Cain's path}}, rushed headlong into Balaam's error to make a profit, and destroyed themselves, as happened\fnote{The Gk. lacks \fbib{as happened}} in Korah's rebellion. \v{12}These people are stains on your love feasts.\fnote{Some early Christians had a meal along with the Lord's Supper.} They feast with you without any sense of awe.\fnote{Or \fbib{without fear}} They are shepherds who care only for themselves. They are waterless clouds blown about by the winds. They are autumn trees that are fruitless, totally\fnote{Lit. \fbib{twice}} dead, and uprooted. \v{13}They are wild waves of the sea, churning up the foam of their own shame. They are wandering stars for whom the deepest darkness has been reserved forever.

\v{14}Enoch, in the seventh generation from Adam, prophesied about these people when he said,

\begin{poetry}
\poeml ``Look! The Lord has come with countless thousands of his holy ones. \v{15}He will judge all people and convict everyone of all the ungodly things that they have done in such an ungodly way, including all the harsh things that these ungodly sinners have said about him.''\fnote{Cf. 1 Enoch 1:9 (\fbib{The Apocrypha})}
\end{poetry}

\v{16}These people are complainers and faultfinders, following their own desires. They say arrogant things and flatter people in order to take advantage of them.
\passage{Advice to the Readers}

\v{17}But you, dear friends, must remember the statements and predictions of the apostles of our Lord Jesus, the Messiah.\fnote{Or \fbib{Christ}} \v{18}They kept telling you, ``In the last times there will be mockers, following their own ungodly desires.''\fnote{Cf. 2 Peter 3:3} \v{19}These are the people who cause divisions. They are worldly, devoid of the Spirit.

\v{20}But you, dear friends, must continue to build your most holy faith for your own benefit. Furthermore, continue to pray in the Holy Spirit. \v{21}Remain in God's love as you look for the mercy of our Lord Jesus the Messiah,\fnote{Or \fbib{Christ}} which brings eternal life. \v{22}Show mercy to those who have doubts. \v{23}Save others by snatching them from the fire. To others, show mercy with fear, hating even the clothes stained by their sinful lives.\fnote{Lit. \fbib{by their flesh}}
\passage{Final Prayer}

\v{24}Now to the one who is able to keep you from falling and to make you stand joyful and faultless in his glorious presence, \v{25}to the only God, our Savior, through Jesus the Messiah,\fnote{Or \fbib{Christ}} our Lord, be glory, majesty, power, and authority before all time and for all eternity! Amen.
