\bookheader{Revelation}
\labelbook{Rev}

\bookpretitle{The Book of}
\booktitle{The Revelation to John}

\labelchapt{1}
\passage{The Revelation of Jesus the Messiah to the Seven Churches}

\chapt{1}
\v{1}This is\fnote{The Gk. lacks \fbib{This is}} the revelation of Jesus the Messiah,\fnote{Or \fbib{Christ}} which God gave him to show his servants the things that must happen soon. He made it known by sending his messenger\fnote{Or \fbib{angel}} to his servant John, \v{2}who testified about this message from God and the testimony about Jesus the Messiah.\fnote{Or \fbib{Christ}} \v{3}How blessed is the one who reads aloud and those who hear the words of this prophecy and obey what is written in it, for the time is near!

\v{4}From\fnote{The Gk. lacks \fbib{From}} John to the seven churches in Asia. May grace and peace be yours from the one who is, who was, and who is coming, from the seven spirits who are in front of his throne, \v{5}and from Jesus the Messiah,\fnote{Or \fbib{Christ}} the witness, the faithful one,\fnote{Or \fbib{Jesus the Messiah, the faithful witness}} the firstborn from the dead, and the ruler over the kings of the earth. To the one who loves us and has freed\fnote{Other mss. read \fbib{has washed}} us from our sins by his blood \v{6}and has made us a kingdom, priests for his God and Father, be glory and power forever and ever! Amen.

\begin{poetry}
\poeml \v{7}Look! He is coming in the clouds. \\
\poemll    Every eye will see him, \\
\poemlll       even those who pierced him, \\
\poeml and all the tribes of the earth will mourn \\
\poemll    because of him.
\end{poetry}

So be it! Amen.

\v{8}\red{``I am the Alpha and the Omega,''} declares the Lord God,\red{ ``the one who is, who was, and who is coming, the Almighty.''}
\passage{John's Vision of the Messiah}

\v{9}I am John, your brother and partner in the oppression, kingdom, and patience that comes because of Jesus. I was on the island called Patmos because of the word of God and the testimony about Jesus. \v{10}I came to be in the Spirit on the Day of the Lord, when I heard a loud voice behind me like a trumpet, \v{11}saying, \red{``Write on a scroll what you see, and send it to the seven churches: Ephesus, Smyrna, Pergamum, Thyatira, Sardis, Philadelphia, and Laodicea.''}

\v{12}Then I turned to see who\fnote{Lit. \fbib{see the voice that}} was talking to me, and when I turned I saw seven gold lamp stands. \v{13}Among the lamp stands there was someone like the Son of Man. He was wearing a long robe with a gold sash around his chest.\fnote{Or \fbib{waist}} \v{14}His head and his hair were white like wool, in fact, as white as snow. His eyes were like flames of fire, \v{15}his feet were like glowing bronze refined in a furnace, and his voice was like the sound of raging waters. \v{16}In his right hand he held seven stars, and out of his mouth came a sharp, two-edged sword. His face was like the sun when it shines with full force.

\v{17}When I saw him, I fell down at his feet like a dead man. But he placed his right hand on me and said, \red{``Stop being afraid! I am the first and the last,} \v{18}\red{the living one. I was dead---but look!---I am alive forever and ever! I have the keys of Death and Hades.}\fnote{I.e. the realm of the dead} \v{19}\red{Therefore, write down what you have seen, what is, and what is going to happen after this.} \v{20}\red{The secret meaning of the seven stars that you saw in my right hand and the seven gold lamp stands is this: the seven stars are the messengers\fnote{Or \fbib{angels}} of the seven churches, and the seven lamp stands are the seven churches.''}
\labelchapt{2}
\passage{The Letter to the Church in Ephesus}

\chapt{2}
\v{1}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Ephesus, write:}

\red{`The one who holds the seven stars in his right hand, the one who walks among the seven gold lamp stands, says this:}

\v{2}\red{`I know what you've been doing, your toil, and your endurance. I also know that you cannot tolerate evil people. You have tested those who call themselves apostles, but are not, and have found them to be false.} \v{3}\red{You have endured and suffered because of my name, yet you have not grown weary.} \v{4}\red{However, I have this against you: You have abandoned the love you had at first.} \v{5}\red{Therefore, remember how far you have fallen. Repent and go back to what you were doing at first. If you don't, I will come to you and remove your lamp stand from its place---unless you repent.} \v{6}\red{But this is to your credit: You hate the actions of the Nicolaitans, which I also hate.}

\v{7}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches. To everyone who conquers I will give the privilege of eating from the tree of life that is in God's paradise.'\,''}
\passage{The Letter to the Church in Smyrna}

\v{8}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Smyrna, write:}

\red{`The first and the last, who was dead and became alive, says this:}

\v{9}\red{`I know your suffering and your poverty---though you are rich---and the slander committed by those who claim to be Jews but are not. They are the synagogue of Satan.} \v{10}\red{Don't be afraid of what you are going to suffer. Look! The devil is going to throw some of you into prison so that you may be tested. For ten days you will undergo suffering. Be faithful until death, and I will give you the victor's crown of life.}

\v{11}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches. The one who conquers will never be hurt by the second death.'\,''}
\passage{The Letter to the Church in Pergamum}

\v{12}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Pergamum, write:}

\red{`The one who holds the sharp, two-edged sword, says this:}

\v{13}\red{`I know where you live. Satan's throne is there. Yet you hold on to my name and have not denied your faith in me,\fnote{Or \fbib{my faith}} even in the days of Antipas, my faithful witness, who was killed in your presence, where Satan lives.} \v{14}\red{But I have a few things against you: You have there some who hold to the teaching of Balaam, the one who taught Balak to put a stumbling block before the people of Israel so that they would eat food sacrificed to idols and practice immorality.} \v{15}\red{You also have some who hold to the teaching of the Nicolaitans.} \v{16}\red{So repent. If you don't, I will come to you quickly and wage war against them with the sword of my mouth.}

\v{17}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches. To the one who conquers I will give some of the hidden manna. I will also give him a white stone. On the white stone is written a new name that no one knows except the person who receives it.'\,''}
\passage{The Letter to the Church in Thyatira}

\v{18}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Thyatira, write:}

\red{`The Son of God, whose eyes are like flaming fire and whose feet are like glowing bronze, says this:}

\v{19}\red{`I know what you've been doing---your love, faithfulness,\fnote{Or \fbib{faith}} service, and endurance---and that your last actions are greater than the first.} \v{20}\red{But I have this against you: You tolerate that woman Jezebel, who calls herself a prophet and who teaches and leads my servants to practice immorality and to eat food sacrificed to idols.} \v{21}\red{I gave her time to repent, but she refused to repent of her immorality.} \v{22}\red{Look! I am going to strike her with illness.\fnote{Lit. \fbib{to throw her into a bed}} Those who commit adultery with her will also suffer greatly, unless they repent from acting like her.} \v{23}\red{I will strike her children dead. Then all the churches will know that I am the one who searches minds and hearts. I will reward each of you as your actions deserve.}

\v{24}\red{`But as for the rest of you in Thyatira---you who do not hold to this teaching and who have not learned what some people call the deep things of Satan---I won't burden you with anything else.} \v{25}\red{Just hold on to what you have until I come.} \v{26}\red{To the person who conquers and continues to do what I've commanded\fnote{Lit. \fbib{do my works}} to the end, I will give authority over the nations.}

\begin{poetry}
\poeml \v{27}\red{`He will rule them with an iron scepter;} \\
\poemll    \red{shattering them like clay pots.'}\fnote{Cf. Isa 30:14; Jer 19:11}
\end{poetry}

\v{28}\red{`Just as I have received authority from my Father, I will also give him the morning star.}

\v{29}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches.'\,''}
\labelchapt{3}
\passage{The Letter to the Church in Sardis}

\chapt{3}
\v{1}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Sardis, write:}

\red{`The one who has the seven spirits of God and the seven stars says this:}

\red{`I know what you've been doing. You are known for being alive, but you are dead.} \v{2}\red{Be alert, and strengthen the things that are left, which are about to die. I note that your actions are incomplete before my God.} \v{3}\red{So remember what you received and heard. Obey it, and repent. If you are not alert, I will come like a thief, and you won't know the time when I will come to you.} \v{4}\red{But you have a few people in Sardis who have not soiled their clothes. They will walk with me in white clothes because they are worthy.} \v{5}\red{The person who conquers in this way will wear white clothes, and I will never erase his name from the Book of Life. I will acknowledge his name in the presence of my Father and his angels.}

\v{6}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches.'\,''}
\passage{The Letter to the Church in Philadelphia}

\v{7}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Philadelphia, write:}

\begin{poetry}
\poeml \red{`The one who is holy, who is true,} \\
\poemll    \red{who has the key of David,} \\
\poeml \red{who opens a door that\fnote{Lit. \fbib{who opens and}} no one can shut,} \\
\poemll    \red{and who shuts a door that\fnote{Lit. \fbib{who shuts and}} no one can open,}
\end{poetry}

\red{`says this:}

\v{8}\red{`I know what you've been doing. Look! I have put in front of you an open door that no one can shut. You have only a little strength, but you have obeyed my word and have not denied my name.} \v{9}\red{I will make those who belong to the synagogue of Satan---those who claim to be Jews and aren't, but are lying---come and bow down at your feet. Then they will realize that I have loved you.} \v{10}\red{Because you have obeyed my command to endure,\fnote{Lit. \fbib{my word of endurance}} I will keep you from the hour of testing that is coming to the whole world to test those living on the earth.} \v{11}\red{I am coming soon! Hold on to what you have so that no one takes your victor's crown.} \v{12}\red{I will make the one who conquers to become a pillar in the sanctuary of my God, and he will never go out of it again. I will write on him the name of my God, the name of the city of my God (the new Jerusalem coming down out of heaven from God), and my own new name.}

\v{13}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches.'\,''}
\passage{The Letter to the Church in Laodicea}

\v{14}\red{``To the messenger\fnote{Or \fbib{angel}} of the church in Laodicea, write:}

\red{`The Amen, the witness who is faithful and true, the originator of God's creation, says this:}

\v{15}\red{`I know your actions, that you are neither cold nor hot. I wish you were cold or hot.} \v{16}\red{Since you are lukewarm and neither hot nor cold, I am going to spit you out of my mouth.} \v{17}\red{You say, ``I am rich. I have become wealthy. I don't need anything.'' Yet you don't realize that you are miserable, pitiful, poor, blind, and naked.} \v{18}\red{Therefore, I advise you to buy from me gold purified in fire so you may be rich, white clothes to wear so your shameful nakedness won't show, and ointment to put on your eyes so you may see.} \v{19}\red{I correct and discipline those whom I love, so be serious and repent!} \v{20}\red{Look! I am standing at the door and knocking. If anyone listens to my voice and opens the door, I will come in to him and eat with him, and he will eat\fnote{The Gk. lacks \fbib{will eat}} with me.} \v{21}\red{To the one who conquers I will give a place to sit with me on my throne, just as I have conquered and have sat down with my Father on his throne.}

\v{22}\red{`Let everyone\fnote{Lit. \fbib{Let the one who has an ear}} listen to what the Spirit says to the churches.'\,''}
\labelchapt{4}
\passage{The Vision of God's Throne}

\chapt{4}
\v{1}After these things, I saw a door standing open in heaven. The first voice that I had heard speaking to me like a trumpet said, \red{``Come up here, and I will show you what must happen after this.''} \v{2}Instantly I was in the Spirit, and I saw a throne in heaven with a person seated on the throne. \v{3}The person sitting there looked like jasper and carnelian, and there was a rainbow around the throne that looked like an emerald. \v{4}Around the throne were 24 other thrones, and on these thrones sat 24 elders wearing white robes and gold victor's crowns on their heads. \v{5}Flashes of lightning, noises, and peals of thunder came from the throne. Burning in front of the throne were seven flaming torches, which are the seven spirits of God.

\v{6}In front of the throne was something like a sea of glass as clear as crystal. In the center of the throne and on each side of the throne were four living creatures full of eyes in front and in back. \v{7}The first living creature was like a lion, the second living creature was like an ox, the third living creature had a face like a human, and the fourth living creature was like a flying eagle. \v{8}Each of the four living creatures had six wings and were full of eyes inside and out. Without stopping day or night they were saying,

\begin{poetry}
\poeml ``Holy, holy, holy \\
\poemll    is the Lord God Almighty, \\
\poemlll       who was, who is, and who is coming.''
\end{poetry}

\v{9}Whenever the living creatures give glory, honor, and thanks to the one who sits on the throne, who lives forever and ever, \v{10}the 24 elders bow down and worship in front of the one who sits on the throne, the one who lives forever and ever. They throw their victor's crowns in front of the throne and say,

\begin{poetry}
\poeml \v{11}``You are worthy, our Lord and God, \\
\poemll    to receive glory, honor, and power, \\
\poemlll       because you created all things; \\
\poeml they came into existence \\
\poeml and were created because of your will.''
\end{poetry}
\labelchapt{5}
\passage{The Vision of the Scroll with Seven Seals}

\chapt{5}
\v{1}Then I saw in the right hand of the one who sits on the throne a scroll written on the inside and on the outside, sealed with seven seals. \v{2}I also saw a powerful angel proclaiming with a loud voice, ``Who is worthy to open the scroll and break its seals?'' \v{3}No one in heaven, on earth, or under the earth could open the scroll or look inside it. \v{4}I began to cry bitterly because no one was found worthy to open the scroll or look inside it.

\v{5}``Stop crying,'' one of the elders told me. ``Look! The Lion from the tribe of Judah, the Root of David, has conquered. He can open the scroll and its seven seals.''
\passage{The Vision of the Lamb Taking the Scroll}

\v{6}Then I saw a lamb standing in the middle of the throne, the four living creatures, and the elders. He looked\fnote{The Gk. lacks \fbib{He looked}} like he had been slaughtered. He had seven horns and seven eyes, which are the seven spirits of God sent into all the earth. \v{7}He went and took the scroll from the right hand of the one who sits on the throne.

\v{8}When the lamb had taken the scroll, the four living creatures and the twenty-four elders bowed down in front of him. Each held a harp and a gold bowl full of incense, the prayers of the saints. \v{9}They sang a new song:

\begin{poetry}
\poeml ``You are worthy to take the scroll and open its seals, \\
\poemll    because you were slaughtered. \\
\poeml With your blood you purchased people\fnote{Some MSS read \fbib{us}; the Gk. lacks \fbib{people}} for God \\
\poemll    from every tribe, language, people, and nation. \\
\poeml \v{10}You made them a kingdom and priests for our God, \\
\poemll    and they will reign on the earth.''
\end{poetry}
\passage{The Vision of the Song of the Lamb}

\v{11}Then I looked, and I heard the voices of many angels, the living creatures, and the elders surrounding the throne. They numbered 10,000's times 10,000 and thousands times thousands. \v{12}They sang with a loud voice,

\begin{poetry}
\poeml ``Worthy is the lamb who was slaughtered \\
\poemll    to receive power, wealth, wisdom, strength, honor, glory, and praise!''
\end{poetry}

\v{13}I heard every creature in heaven, on earth, under the earth, and on the sea, and everything that is in them, saying,

\begin{poetry}
\poeml ``To the one who sits on the throne and to the lamb \\
\poemll    be praise, honor, glory, and power forever and ever!''
\end{poetry}

\v{14}Then the four living creatures said, ``Amen!'', and the elders bowed down and worshipped.
\labelchapt{6}
\passage{The Vision of the First Seal Opened}

\chapt{6}
\v{1}Then I saw the lamb open the first of the seven seals. I heard one of the four living creatures say with a voice like thunder, ``Go!'' \v{2}Then I looked, and there was a white horse! Its rider had a bow, and a victor's crown had been given to him. He went out as a conqueror to conquer.
\passage{The Vision of the Second Seal Opened}

\v{3}When the lamb\fnote{Lit. \fbib{he}} opened the second seal, I heard the second living creature say, ``Go!'' \v{4}A second horse went out. It was fiery red, and its rider was given permission to take peace away from the earth and to make people slaughter one another. So he was given a large sword.
\passage{The Vision of the Third Seal Opened}

\v{5}When the lamb\fnote{Lit. \fbib{he}} opened the third seal, I heard the third living creature say, ``Go!'' I looked, and there was a black horse! Its rider held a scale in his hand. \v{6}I heard what sounded like a voice from among the four living creatures, saying, ``One day's ration of wheat for a day's wage, or three day's ration of barley for a day's wage!\fnote{A denarius was equivalent to a day's wage for a laborer.} But don't damage the olive oil or the wine!''
\passage{The Vision of the Fourth Seal Opened}

\v{7}When the lamb\fnote{Lit. \fbib{he}} opened the fourth seal, I heard the voice of the fourth living creature say, ``Go!'' \v{8}I looked, and there was a pale green horse! Its rider's name was Death, and Hades\fnote{I.e. the realm of the dead} followed him. They were given authority over one-fourth of the earth to kill people using wars, famines, plagues, and the wild animals of the earth.
\passage{The Vision of the Fifth Seal Opened}

\v{9}When the lamb\fnote{Lit. \fbib{he}} opened the fifth seal, I saw under the altar the souls of those who had been slaughtered because of the word of God and the testimony they had given. \v{10}They cried out in a loud voice,

\begin{poetry}
\poeml ``Holy and true Sovereign, \\
\poemll    how long will it be before you judge \\
\poeml and take revenge on those living on the earth \\
\poemll    who shed our blood?''
\end{poetry}

\v{11}Each of them was given a white robe. They were told to rest a little longer until the number of\fnote{The Gk. lacks \fbib{the number of}} their fellow servants and their brothers was completed, who would be killed as they themselves had been.
\passage{The Vision of the Sixth Seal Opened}

\v{12}Then I saw the lamb\fnote{Lit. \fbib{him}} open the sixth seal. There was a powerful earthquake. The sun turned as black as sackcloth made of hair, and the full moon turned as red as blood.\fnote{Lit. \fbib{became like blood}} \v{13}The stars in the sky fell to the earth like a fig tree drops its fruit when it is shaken by a strong wind. \v{14}The sky vanished like a scroll being rolled up, and every mountain and island was moved from its place. \v{15}Then the kings of the earth, the important people, the generals, the rich, the powerful, and all the slaves and free people concealed themselves in caves and among the rocks in the mountains. \v{16}They told the mountains and rocks, ``Fall on us and hide us from the face of the one who sits on the throne and from the wrath of the lamb. \v{17}For the great day of their wrath has come, and who is able to endure it?''
\labelchapt{7}
\passage{The Vision of the Sealing of 144,000 People}

\chapt{7}
\v{1}After this, I saw four angels standing at the four corners of the earth. They were holding back the four winds of the earth so that no wind could blow on the land, on the sea, or on any tree. \v{2}I saw another angel coming from the east having the seal of the living God. He cried out in a loud voice to the four angels who had been permitted to harm the land and sea, \v{3}``Don't harm the land, the sea, or the trees until we have marked the servants of our God with a seal on their foreheads.''

\v{4}I heard the number of those who were sealed: 144,000. Those who were sealed were from every tribe of Israel: \v{5}12,000 from the tribe of Judah were sealed, 12,000 from the tribe of Reuben, 12,000 from the tribe of Gad, \v{6}12,000 from the tribe of Asher, 12,000 from the tribe of Naphtali, 12,000 from the tribe of Manasseh, \v{7}12,000 from the tribe of Simeon, 12,000 from the tribe of Levi, 12,000 from the tribe of Issachar, \v{8}12,000 from the tribe of Zebulun, 12,000 from the tribe of Joseph, and 12,000 from the tribe of Benjamin were sealed.
\passage{The Vision of Tribulation Saints}

\v{9}After these things, I looked, and there was a crowd so large that no one was able to count it! They were from every nation, tribe, people, and language. They were standing in front of the throne and the lamb and were wearing white robes, with palm branches in their hands. \v{10}They cried out in a loud voice,

\begin{poetry}
\poeml ``Salvation belongs to our God, \\
\poemll    who sits on the throne, \\
\poemlll       and to the lamb!''
\end{poetry}

\v{11}All the angels stood around the throne and around the elders and the four living creatures. They fell on their faces in front of the throne and worshipped God, \v{12}saying,

\begin{poetry}
\poeml ``Amen! Praise, glory, wisdom, thanks, honor, power, and strength \\
\poemll    be to our God forever and ever! Amen!''
\end{poetry}

\v{13}``Who are these people wearing white robes,'' one of the elders asked me, ``and where did they come from?''

\v{14}I told him, ``Sir, you know.''

Then he told me, ``These are the people who are coming out of the terrible suffering.\fnote{Or \fbib{great tribulation}} They have washed their robes and made them white in the blood of the lamb. \v{15}That is why:

\begin{poetry}
\poeml ``They are in front of the throne of God \\
\poemll    and worship\fnote{Or \fbib{serve}} him night and day in his Temple. \\
\poeml The one who sits on the throne will shelter them. \\
\poeml \v{16}They will never be hungry or thirsty again. \\
\poemll    Neither the sun nor its heat will ever beat down on them, \\
\poeml \v{17}because the lamb in the center of the throne will be their shepherd. \\
\poemll    He will lead them to springs filled with the water of life, \\
\poemlll       and God will wipe every tear from their eyes.''\fnote{Isa 25:8}
\end{poetry}
\labelchapt{8}
\passage{The Vision of the Seventh Seal Opened}

\chapt{8}
\v{1}When the lamb\fnote{Lit. \fbib{he}} opened the seventh seal, there was silence in heaven for about half an hour.
\passage{The Vision of Seven Angels Given Seven Trumpets}

\v{2}Then I saw the seven angels who stand in God's presence, and seven trumpets were given to them. \v{3}Another angel came with a gold censer and stood at the altar. He was given a large quantity of incense to offer on the gold altar before the throne, along with the prayers of all the saints. \v{4}The smoke from the incense and the prayers of the saints went up from the angel's hand to God. \v{5}The angel took the censer, filled it with fire from the altar, and threw it on the earth. Then there were peals of thunder, noises, flashes of lightning, and an earthquake. \v{6}The seven angels who had the seven trumpets got ready to blow them.
\passage{The Vision of the First Four Trumpets}

\v{7}When the first angel blew his trumpet, hail and fire were mixed with blood and thrown on the earth. One-third of the earth was burned up, one-third of the trees was burned up, and all the green grass was burned up.

\v{8}When the second angel blew his trumpet, something like a huge mountain burning with fire was thrown into the sea. One-third of the sea turned into blood, \v{9}one-third of the creatures that were living in the sea died, and one-third of the ships was destroyed.

\v{10}When the third angel blew his trumpet, a huge star blazing like a torch fell from heaven. It fell on one-third of the rivers and on the springs of water. \v{11}The name of the star is Wormwood. One-third of the water turned into wormwood, and many people died from the water because it had turned bitter.

\v{12}When the fourth angel blew his trumpet, one-third of the sun, one-third of the moon, and one-third of the stars were struck so that one-third of them turned dark. One-third of the day was kept from having light, as was the night.
\passage{The Vision of the Eagle Flying}

\v{13}Then I looked, and I heard an eagle flying overhead say in a loud voice,

\begin{poetry}
\poeml ``How terrible, how terrible, how terrible \\
\poemll    for those living on the earth, \\
\poeml because of the blasts of the remaining trumpets \\
\poemll    that the three angels are about to blow!''
\end{poetry}
\labelchapt{9}
\passage{The Vision of the Fifth Trumpet}

\chapt{9}
\v{1}When the fifth angel blew his trumpet, I saw a star that had fallen to earth\fnote{I.e. Lucifer} from the sky.\fnote{Or \fbib{from heaven}} The star\fnote{Lit. \fbib{It}} was given the key to the shaft of the bottomless pit.\fnote{I.e. the realm of punishment in the afterlife} \v{2}It opened the shaft of the bottomless pit,\fnote{I.e. the realm of punishment in the afterlife} and smoke came out of the shaft like the smoke from a large furnace. The sun and the air were darkened with the smoke from the shaft. \v{3}Locusts came out of the smoke onto the earth, and they were given power like that of earthly scorpions. \v{4}They were told not to harm the grass on the earth, any green plant, or any tree. They could harm\fnote{The Gk. lacks \fbib{They could harm}} only the people who do not have the seal of God on their foreheads. \v{5}They were not allowed to kill them, but were only allowed\fnote{The Gk. lacks \fbib{were only allowed}} to torture them for five months. Their torture was like the pain of a scorpion when it stings someone. \v{6}In those days people will seek death, but never find it. They will long to die, but death will escape them.

\v{7}The locusts looked like horses prepared for battle. On their heads were victor's crowns that looked like gold, and their faces were like human faces. \v{8}They had hair like women's hair and teeth like lions' teeth. \v{9}They had breastplates like iron, and the noise of their wings was like the roar of chariots with many horses rushing into battle. \v{10}They had tails and stingers like scorpions, and they had the power to hurt people with their tails for five months. \v{11}They had the angel of the bottomless pit\fnote{I.e. the realm of punishment in the afterlife} ruling over them as king. In Hebrew he is called Abaddon,\fnote{I.e. the realm of destruction in the afterlife} and in Greek he is called Apollyon.\fnote{I.e. Destroyer}

\v{12}The first catastrophe is over. After these things, there are still two more catastrophes to come.
\passage{The Vision of the Sixth Trumpet}

\v{13}When the sixth angel blew his trumpet, I heard a voice from the four\fnote{Other mss. lack \fbib{four}} horns of the gold altar in front of God. \v{14}It told the sixth angel who had the trumpet, ``Release the four angels who are held at the great Euphrates River.'' \v{15}So the four angels who were ready for that hour, day, month, and year were released to kill one-third of humanity. \v{16}The number of cavalry troops was 200,000,000. I heard how many there were.\fnote{Lit. \fbib{heard their number}}

\v{17}This was how I saw the horses in my vision: The riders wore breastplates that had the color of fire, sapphire, and sulfur. The heads of the horses were like lions' heads, and fire, smoke, and sulfur came out of their mouths. \v{18}By these three plagues---the fire, the smoke, and the sulfur that came out of their mouths---one-third of humanity was killed. \v{19}For the power of these horses is in their mouths and their tails. Their tails have heads like snakes, which they use to inflict pain.

\v{20}The rest of the people who survived these plagues did not repent from their evil actions\fnote{Lit. \fbib{from the works of their hands}} or stop worshiping demons and idols made of gold, silver, bronze, stone, and wood, which cannot see, hear, or walk. \v{21}They did not repent from their murders, their witchcraft, their sexual immorality, or their thefts.
\labelchapt{10}
\passage{The Vision of the Powerful Angel}

\chapt{10}
\v{1}Then I saw another powerful angel come down from heaven. He was dressed in a cloud, and there was a rainbow over his head. His face was like the sun, and his legs were like columns of fire. \v{2}He held a small, opened scroll in his hand. Setting his right foot on the sea and his left foot on the land, \v{3}he shouted in a loud voice as a lion roars. When he shouted, the seven thunders spoke with voices of their own. \v{4}When the seven thunders spoke, I was going to write, but I heard a voice from heaven say, ``Seal up what the seven thunders have said, and don't write it down.''

\v{5}Then the angel whom I saw standing on the sea and on the land raised his right hand to heaven. \v{6}He swore an oath by the one who lives forever and ever, who created heaven and everything in it, the earth and everything in it, and the sea and everything in it: ``There will be no more delay. \v{7}When the time approaches\fnote{Lit. \fbib{days approach}} for the seventh angel to blow his trumpet, God's secret plan\fnote{The Gk. lacks \fbib{plan}} will be fulfilled, as he had announced to his servants, the prophets.''

\v{8}Then the voice that I had heard from heaven spoke to me again, saying, ``Go and take the opened scroll from the hand of the angel who is standing on the sea and on the land.''

\v{9}So I went to the angel and asked him to give me the small scroll. ``Take it and eat it,'' he told me. ``It will turn bitter in your stomach, but it will be as sweet as honey in your mouth.''

\v{10}So I took the small scroll from the angel's hand and ate it. It was as sweet as honey in my mouth, but when I had eaten it, it turned bitter in my stomach. \v{11}Then the seven thunders\fnote{Lit. \fbib{they}} told me, ``You must prophesy again about many people, nations, languages, and kings.''
\labelchapt{11}
\passage{The Vision of the Two Witnesses}

\chapt{11}
\v{1}Then I was given a stick like a measuring rod. I was told, ``Stand up and measure the Temple of God and the altar, and count\fnote{The Gk. lacks \fbib{count}} those who worship there. \v{2}But don't measure the courtyard outside the Temple. Leave that out, because it is given to the nations, and they will trample the Holy City\fnote{I.e. Jerusalem} for 42 months. \v{3}I will give my two witnesses who wear sackcloth the authority to prophesy for 1,260 days.''

\v{4}These witnesses\fnote{Lit. \fbib{These ones}} are the two olive trees and the two lamp stands standing in the presence of the Lord of the earth. \v{5}And if anyone should want to hurt them, fire comes out of their mouths and burns up their enemies. If anyone wants to hurt them, he must be killed in this manner. \v{6}These witnesses\fnote{Lit. \fbib{They}} have authority to close the heavens in order to keep rain from falling while they are\fnote{Lit. \fbib{falling during the days of their}} prophesying. They also have authority to turn bodies of water into blood and to strike the earth with any plague, as often as they desire.

\v{7}When they have finished their testimony, the beast that comes up from the bottomless pit\fnote{I.e. the realm of punishment in the afterlife} will wage war against them, conquer them, and kill them. \v{8}Their dead bodies will lie in the street of the great city that is spiritually called Sodom and Egypt, where their Lord was crucified. \v{9}For three and a half days some members of the people, tribes, languages, and nations will look at their dead bodies and will not allow them to be placed in a tomb. \v{10}Those living on earth will gloat over them, celebrate, and send gifts to each other, because these two prophets had tormented those living on earth.
\passage{The Resurrection of the Witnesses}

\v{11}But after the three and a half days, the breath of life from God entered them, and they stood on their feet. Those who watched them were terrified. \v{12}Then the witnesses\fnote{Lit. \fbib{they}; other mss. read \fbib{I}} heard a loud voice from heaven calling to them, ``Come up here!'' So they went up to heaven in a cloud, and their enemies watched them. \v{13}At that moment a powerful earthquake struck. One-tenth of the city collapsed, 7,000 people were killed by the earthquake, and the rest were terrified and gave glory to the God of heaven.

\v{14}The second catastrophe is over. The third catastrophe is coming very soon.
\passage{The Vision of the Seventh Angel Blowing His Trumpet}

\v{15}When the seventh angel blew his trumpet, there were loud voices in heaven, saying,

\begin{poetry}
\poeml ``The world's kingdom has become \\
\poemll    the kingdom of our Lord and of his Messiah,\fnote{Or \fbib{Christ}} \\
\poemlll       and he will rule forever and ever.''
\end{poetry}

\v{16}Then the twenty-four elders who were sitting on their thrones in God's presence fell on their faces and worshipped God. \v{17}They said,

\begin{poetry}
\poeml ``We give thanks to you, Lord God Almighty, \\
\poemll    who is and who was, \\
\poeml because you have taken your great power \\
\poemll    and have begun to rule. \\
\poeml \v{18}The nations were angry, \\
\poemll    but the time for\fnote{The Gk. lacks \fbib{the time for}} your wrath has come. \\
\poeml It is time for the dead to be judged--- \\
\poemll    to reward your servants, the prophets, the saints, and all who fear your name, \\
\poemlll       both unimportant and important, \\
\poemll    and to destroy those who destroy the earth.''
\end{poetry}

\v{19}Then the Temple of God in heaven was opened, and the ark of his covenant was seen inside his Temple. There were flashes of lightning, noises, peals of thunder, an earthquake, and heavy hail.
\labelchapt{12}
\passage{The Vision of a Woman Dressed with the Sun}

\chapt{12}
\v{1}A spectacular sign appeared in the sky: a woman dressed with the sun, who had the moon under her feet and a victor's crown of twelve stars on her head. \v{2}She was pregnant and was crying out from her labor pains, the agony of giving birth.
\passage{The Vision of the Red Dragon}

\v{3}Then another sign appeared in the sky: a huge red dragon with seven heads, ten horns, and seven royal crowns on its heads. \v{4}Its tail swept away one-third of the stars in the sky and knocked them down to the earth. Then the dragon stood in front of the woman who was about to give birth so that it could devour her child when it was born. \v{5}She gave birth to a son, a boy, who is to rule\fnote{Or \fbib{shepherd}} all the nations with an iron scepter. But her child was snatched away and taken to God and to his throne. \v{6}Then the woman fled into the wilderness, where a place had been prepared for her by God so that she might be taken care of for 1,260 days.
\passage{The Vision of War in Heaven}

\v{7}Then a war broke out in heaven. Michael and his angels fought with the dragon, and the dragon and its angels fought back. \v{8}But it was not strong enough, and there was no longer any place for them in heaven. \v{9}The huge dragon was hurled down. That ancient serpent, called the Devil and Satan, the deceiver of the whole world, was hurled down to the earth, along with its angels.
\passage{The Vision of the Cry of Victory}

\v{10}Then I heard a loud voice in heaven say,

\begin{poetry}
\poeml ``Now the salvation, the power, \\
\poemll    the kingdom of our God, \\
\poemlll       and the authority of his Messiah\fnote{Or \fbib{Christ}} have come. \\
\poeml For the one who accuses our brothers, \\
\poemll    who accuses them day and night \\
\poeml in the presence of our God, \\
\poemll    has been thrown out. \\
\poeml \v{11}Our brothers\fnote{Lit. \fbib{They}} conquered him by the blood of the lamb \\
\poemll    and by the word of their testimony, \\
\poeml for they did not cling to their lives \\
\poemll    even in the face of death. \\
\poeml \v{12}So be glad, heavens, and those who live in them! \\
\poeml How terrible it is for the earth and the sea, \\
\poemll    because the Devil has come down to you, filled with rage, \\
\poemlll       knowing that his time is short!''
\end{poetry}
\passage{The Vision of Persecution of the Woman and Her Children}

\v{13}When the dragon saw that it had been thrown down to the earth, it pursued\fnote{Or \fbib{persecuted}} the woman who had given birth to the boy. \v{14}However, the woman was given the two wings of a large eagle so that she could fly away from the serpent to her place in the wilderness, where she could be taken care of for a time, times, and half a time. \v{15}From its mouth the serpent spewed water like a river behind the woman in order to sweep her away with the flood. \v{16}But the earth helped the woman by opening its mouth and swallowing the river that the dragon had spewed from its mouth. \v{17}The dragon became angry with the woman and went away to do battle against the rest of her children, the ones who keep God's commandments and hold on to the testimony about Jesus. \v{18}Then the dragon\fnote{Lit. \fbib{Then it}; other mss. read \fbib{Then I}} stood on the sand of the seashore.\fnote{Some translations include this sentence within 12:17; other translations include this sentence as part of 13:1.}
\labelchapt{13}
\passage{The Vision of the Beast from the Sea}

\chapt{13}
\v{1}I saw a beast coming out of the sea. It had ten horns, seven heads, and ten royal crowns on its horns. On its heads were blasphemous names. \v{2}The beast that I saw was like a leopard. Its feet were like bear's feet, and its mouth was like a lion's mouth. The dragon gave it his power, his throne, and complete authority.

\v{3}One of the beast's\fnote{Lit. \fbib{its}} heads looked like it had sustained a mortal wound, but its fatal wound was healed. Rapt with amazement, the whole world followed the beast. \v{4}They worshipped the dragon because it had given authority to the beast. They also worshipped the beast, saying, ``Who is like the beast, and who can fight a war with it?'' \v{5}The beast was allowed\fnote{Lit. \fbib{was given a mouth}} to speak arrogant and blasphemous things, and it was given authority for 42 months. \v{6}It uttered\fnote{Lit. \fbib{It opened its mouth to utter}} blasphemies against God, against his name, and against his residence,\fnote{Lit. \fbib{tent}} that is, against those who are living in heaven. \v{7}It was allowed to wage war against the saints and to conquer them.\fnote{Other mss. lack \fbib{It was allowed to wage war against the saints and to conquer them.}} It was also given authority over every tribe, people, language, and nation. \v{8}All those who had become settled down and at home, living on the earth, will worship it, everyone whose name had not been written in the Book of Life belonging to the lamb that had been slaughtered since the foundation of the world.

\begin{poetry}
\poeml \v{9}Let everyone listen:\fnote{Lit. \fbib{Let the one who has an ear}} \\
\poeml \v{10}If anyone is to be taken captive, \\
\poemlll       into captivity he will go. \\
\poeml If anyone is to be killed with a sword, \\
\poemll    with a sword he will be killed.
\end{poetry}
\passage{The Vision of the Beast from the Earth}

Here is a call for endurance and faith of the saints:

\v{11}I saw another beast coming up out of the earth. It had two horns like a lamb and it talked like a dragon. \v{12}It uses all the authority of the first beast on its behalf,\fnote{Or \fbib{in its presence}} and it makes the earth and those living on it worship the first beast, whose mortal wound was healed. \v{13}It performs spectacular signs, even making fire come down from heaven to earth in front of people. \v{14}It deceives those living on earth with the signs that it is allowed to do on behalf of\fnote{Or \fbib{in the presence of}} the first\fnote{The Gk. lacks \fbib{first}} beast, telling them to make an image for the beast who was wounded by a sword and yet lived. \v{15}The second beast\fnote{Lit. \fbib{It}} was allowed to impart life to the image of the first\fnote{The Gk. lacks \fbib{first}} beast so that the image of the beast could talk and order the execution of those who would not worship the image of the beast. \v{16}The second beast\fnote{Lit. \fbib{It}} forces all people---important and unimportant, rich and poor, free and slaves---to be marked on their right hands or on their foreheads, \v{17}so that no one may buy or sell unless he has the mark, which is the beast's name or the number of its name.

\v{18}In this case wisdom is needed: Let the person who has understanding\fnote{Lit. \fbib{has a mind}} calculate\fnote{Or \fbib{decide}} the total\fnote{Or \fbib{the multitude}} of the beast, since it is a human multitude,\fnote{Or \fbib{is a multitude of a man}; or \fbib{is the number of a human being}} and the sum of the multitude\fnote{Or \fbib{total number}} is 600, 60, and six.\fnote{Other mss. read \fbib{616}; some scholars suggest that the Gk. letters {$\chi$}, {$\xi$}, and {$\varsigma$} (totaling 666) may resemble the proto-Arabic term for ``in the name of Allah''}
\labelchapt{14}
\passage{The Vision of the New Song on Mount Zion}

\chapt{14}
\v{1}Then I looked, and there was the lamb, standing on Mount Zion! With him were 144,000 people who had his name and his Father's name written on their foreheads.

\v{2}Then I heard a sound from heaven like that of many waters and like the sound of loud thunder. The sound I heard was like harpists playing on their harps. \v{3}They were singing a new song in front of the throne, the four living creatures, and the elders. No one could learn the song except the 144,000 who had been redeemed from the earth. \v{4}They have not defiled themselves with women, for they are virgins, and they follow the lamb wherever he goes. They have been redeemed from among humanity as the first fruits for God and the lamb. \v{5}In their mouth no lie was found. They are blameless.
\passage{The Vision of Angels Sounding a Warning}

\v{6}Then I saw another angel flying overhead with the eternal gospel to proclaim to those who live\fnote{Lit. \fbib{sit}} on earth---to every nation, tribe, language, and people. \v{7}He said in a loud voice,

\begin{poetry}
\poeml ``Fear God and give him glory, \\
\poemll    because the time for him to judge has arrived. \\
\poeml Worship the one who made heaven and earth, \\
\poemll    the sea and springs of water.''
\end{poetry}

\v{8}Then another angel, a second one, followed him, saying,

\begin{poetry}
\poeml ``Fallen! Babylon the Great has fallen! \\
\poemll    She has made all nations drink the wine, \\
\poemlll       the wrath earned for her sexual sins.''
\end{poetry}

\v{9}Then another angel, a third one, followed them, saying in a loud voice, ``Whoever worships the beast and its image and receives a mark on his forehead or his hand \v{10}will drink the wine of God's wrath, which has been poured undiluted into the cup of his anger. He will be tortured with fire and sulfur in the presence of the holy angels and the lamb. \v{11}The smoke from their torture goes up forever and ever. There is no rest day or night for those who worship the beast and its image or for anyone who receives the mark of its name.''
\passage{A Call for Endurance}

\v{12}Here is a call for\fnote{The Gk. lacks \fbib{a call for}} the endurance of the saints, who keep the commandments of God and hold fast to their faithfulness in Jesus:

\v{13}I heard a voice from heaven say, ``Write this:

\begin{poetry}
\poeml How blessed are the dead, \\
\poemll    that is, those who die in the Lord from now on!'' \\
\poeml ``Yes,'' says the Spirit. \\
\poemll    ``Let them rest from their labors, \\
\poemlll       for their actions follow them.''
\end{poetry}
\passage{The Vision of Earth Harvested}

\v{14}Then I looked, and there was a white cloud! On the cloud sat someone who was like the Son of Man, with a gold victor's crown on his head and a sharp sickle in his hand. \v{15}Another angel came out of the Temple, crying out in a loud voice to the one who sat on the cloud,

\begin{poetry}
\poeml ``Swing your sickle, \\
\poemll    and gather the harvest, \\
\poeml for the hour has come to gather it, \\
\poemll    because the harvest of the earth is fully ripe.''
\end{poetry}

\v{16}The one who sat on the cloud swung his sickle across the earth, and the earth was harvested.

\v{17}Then another angel came out of the Temple in heaven. He, too, had a sharp sickle. \v{18}From the altar came another angel who had authority over fire. He called out in a loud voice to the angel\fnote{Lit. \fbib{to the one}} who had the sharp sickle,

\begin{poetry}
\poeml ``Swing your sharp sickle, \\
\poemll    and gather the bunches of grapes \\
\poeml from the vine of the earth, \\
\poemll    because those grapes are ripe.''
\end{poetry}

\v{19}So the angel swung his sickle in the earth, gathered the grapes from the earth, and threw them into the great winepress of God's wrath. \v{20}The wine press was trampled outside the city, and blood flowed from the wine press as high as a horse's bridle for about 1,600 stadia.\fnote{I.e. about 194 miles; one Roman stadion was about 640 feet}
\labelchapt{15}
\passage{The Vision of Seven Angels with Seven Plagues}

\chapt{15}
\v{1}I saw another sign in heaven. It was both spectacular and amazing. There were seven angels with the seven last plagues, with which God's wrath is completed.
\passage{The Vision of the Sea of Glass}

\v{2}Then I saw what looked like a sea of glass mixed with fire. Those who had conquered the beast, its image, and the number of its name were standing on the sea of glass holding God's harps in their hands. \v{3}They sang the song of God's servant Moses and the song of the lamb:

\begin{poetry}
\poeml ``Your deeds are both spectacular and amazing, \\
\poemll    Lord God Almighty. \\
\poeml Your ways are just and true, \\
\poemll    King of the nations.\fnote{Other mss. read \fbib{ages}} \\
\poeml \v{4}Lord, who won't fear and praise your name? \\
\poemll    For you alone are holy, \\
\poemlll       and all the nations will come and worship you \\
\poemll    because your judgments have been revealed.''
\end{poetry}
\passage{The Vision of the Temple Opened}

\v{5}After these things, I looked, and the Temple, which is the Tent of Testimony in heaven, was open! \v{6}The seven angels with the seven plagues came out of the Temple wearing clean, shining linen with gold sashes around their chests.\fnote{Or \fbib{waists}} \v{7}One of the four living creatures gave to the seven angels seven gold bowls full of the wrath of God, who lives forever and ever. \v{8}The Temple was filled with smoke from the glory of God and his power, and no one could enter the Temple until the seven plagues of the seven angels came to an end.
\labelchapt{16}
\passage{The Vision of the Seven Angels Pouring Out Their Bowls}

\chapt{16}
\v{1}Then I heard a loud voice from the Temple saying to the seven angels, ``Go and pour the seven bowls of God's wrath on the earth.'' \v{2}So the first angel went and poured his bowl on the earth. A horrible, painful sore appeared on the people who had the mark of the beast and worshipped the image.

\v{3}The second angel poured his bowl into the sea. It became like the blood of a dead body, and every living thing in the sea died.

\v{4}The third angel poured his bowl into the rivers and the springs of water, and they turned into blood. \v{5}Then I heard the angel in charge\fnote{The Gk. lacks \fbib{in charge}} of the water say,

\begin{poetry}
\poeml ``You are just. You are the one who is \\
\poemll    and who was, the Holy One, \\
\poemlll       because you have judged these things. \\
\poeml \v{6}You have given them blood to drink \\
\poemll    because they spilled the blood of saints and prophets. \\
\poemlll       This is what they deserve.''
\end{poetry}

\v{7}Then I heard the altar reply,

\begin{poetry}
\poeml ``Yes, Lord God Almighty, \\
\poemll    your judgments are true and just.''
\end{poetry}

\v{8}The fourth angel poured his bowl on the sun, which then was allowed to burn people with fire, \v{9}and they were burned by the fierce heat. They cursed the name of God, who has the authority over these plagues. They did not repent and give him glory.

\v{10}The fifth angel poured his bowl on the throne of the beast. Its kingdom was plunged into darkness. People\fnote{Lit. \fbib{They}} gnawed on their tongues in anguish \v{11}and cursed the God of heaven because of their pain-filled sores. But they did not repent of their behavior.

\v{12}The sixth angel poured his bowl on the great Euphrates River. Its water was dried up to prepare the way for the kings from the east. \v{13}Then I saw three disgusting spirits like frogs come out of the mouth of the dragon, out of the mouth of the beast, and out of the mouth of the false prophet. \v{14}They are demonic spirits that perform signs. They go to the kings of the whole earth and gather them for the war of the great Day of God Almighty.

\begin{poetry}
\poeml \v{15}\red{``See, I am coming like a thief.} \\
\poeml \red{How blessed is the person who remains alert} \\
\poemll    \red{and keeps his clothes on!} \\
\poeml \red{He won't have to go naked} \\
\poemll    \red{and let others see his shame.''}
\end{poetry}

\v{16}The spirits\fnote{Lit. \fbib{They}} gathered the kings\fnote{Lit. \fbib{them}} to the place that is called Armageddon in Hebrew.

\v{17}The seventh angel threw the contents of\fnote{The Gk. lacks \fbib{the contents}} his bowl across the sky. A loud voice came from the throne in the Temple and said, ``It has happened!'' \v{18}There were flashes of lightning, noises, peals of thunder, and a powerful earthquake. There has never been such a powerful earthquake since people have been on the earth. \v{19}The great city was split into three parts, and the cities of the nations fell. God remembered to give Babylon the Great the cup of wine filled with the fury of his wrath. \v{20}Every island vanished,\fnote{Or \fbib{fled}} and the mountains could no longer be found. \v{21}Huge hailstones, each weighing about a talent,\fnote{I.e. about 75 pounds; a talent weighed about 75 pounds} fell from the sky on people, who cursed God because the plague of hail was so terrible.
\labelchapt{17}
\passage{The Vision of Babylon the Great}

\chapt{17}
\v{1}Then one of the seven angels who held the seven bowls came and told me, ``Come, I will show you how the notorious prostitute who sits on many waters will be judged. \v{2}The kings of the earth committed sexual immorality with her, and those living on earth became drunk with the wine of her immorality.''

\v{3}Then the angel\fnote{Lit. \fbib{he}} carried me away in the Spirit into a wilderness. I saw a woman sitting on a scarlet beast that was controlled by blasphemy.\fnote{Lit. \fbib{was filled with blasphemous names}} It had seven heads and ten horns. \v{4}The woman wore purple and scarlet clothes and was adorned with gold, gems, and pearls. In her hand she was holding a gold cup filled with detestable things and the impurities of her immorality. \v{5}On her forehead was written a secret name:

\divine{Babylon the Great},

\divine{the Mother of Prostitutes}

\divine{and Detestable Things of the Earth}

\v{6}I saw that the woman was drunk with the blood of the saints and the blood of the witnesses to Jesus. I was very surprised when I saw her.

\v{7}``Why are you surprised?'' The angel asked me. ``I will tell you the secret of the woman and the beast with the seven heads and the ten horns that carries her. \v{8}The beast that you saw existed once, but is no longer, and is going to crawl out of the bottomless pit\fnote{I.e. the realm of punishment in the afterlife} and then proceed to its destruction. Those living on earth, whose names were not written in the Book of Life from the foundation of the world, will be surprised when they see the beast because it was, is no longer, and will come again. \v{9}This calls for a mind that has wisdom. The seven heads are seven mountains on which the woman is sitting. They are also seven kings. \v{10}Five of them have fallen, one is living, and the other has not yet come. When he comes, he must remain in power\fnote{The Gk. lacks \fbib{in power}} for a little while. \v{11}The beast that was and is no longer is the eighth king,\fnote{The Gk. lacks \fbib{king}} but it belongs with the seven kings\fnote{The Gk. lacks \fbib{kings}} and goes to its destruction. \v{12}The ten horns that you saw are ten kings who have not yet received a kingdom. They will receive authority to rule\fnote{The Gk. lacks \fbib{to rule}} as kings with the beast for one hour. \v{13}They have one purpose: to give their power and authority to the beast. \v{14}They will wage war against the lamb, but the lamb will conquer them because he is Lord of lords and King of kings. Those who are called, chosen, and faithful are with him.''

\v{15}The angel\fnote{Lit. \fbib{He}} also told me, ``The bodies of water you saw, on which the prostitute is sitting, are people, multitudes, nations, and languages. \v{16}The ten horns and the beast you saw will hate the prostitute. They will leave her abandoned and naked. They will eat her flesh and burn her up with fire, \v{17}for God has placed within them a desire to carry out his purpose by uniting to give their kingdom to the beast until God's words are fulfilled. \v{18}The woman you saw is the great city that rules over the kings of the earth.''
\labelchapt{18}
\passage{The Vision of the Fall of Babylon}

\chapt{18}
\v{1}After these things, I saw another angel coming down from heaven. He had great authority,\fnote{Or \fbib{tremendous power}} and the earth was made bright by his splendor. \v{2}He cried out in a powerful voice,

\begin{poetry}
\poeml ``Fallen! Babylon the Great has fallen! \\
\poemll    She has become a home for demons. \\
\poeml She is a prison for every unclean spirit, \\
\poemll    a prison for every unclean bird, \\
\poemll    and a prison for every unclean \\
\poemlll       and hated beast.
\end{poetry}

\begin{poetry}
\poeml \v{3}For all the nations have drunk \\
\poemll    from the wine of her sexual immorality, \\
\poeml and the kings of the earth have committed sexual immorality with her. \\
\poeml The world's businesses have become rich \\
\poemll    from her luxurious excesses.''
\end{poetry}
\passage{The Warning to Leave Babylon}

\v{4}Then I heard another voice from heaven saying,

\begin{poetry}
\poeml ``Come out of her, my people, \\
\poemll    so that you don't participate in her sins \\
\poemlll       and also suffer from her diseases. \\
\poeml \v{5}For her sins are piled as high as heaven, \\
\poemll    and God has remembered her crimes. \\
\poeml \v{6}Do to her as she herself has done, \\
\poemll    and give her double for her deeds. \\
\poeml Mix a double drink for her in the cup she mixed. \\
\poeml \v{7}Just as she glorified herself and lived in luxury, \\
\poemll    inflict on her just as much torture and misery. \\
\poeml In her heart she says, \\
\poemll    `I am a queen on a throne, not a widow. \\
\poemlll       I will never see misery.' \\
\poeml \v{8}For this reason, her diseases that result in death, misery, and famine \\
\poemll    will come in a single day. \\
\poeml She will be burned up in a fire, \\
\poemll    because powerful is the Lord God who judges her.''
\end{poetry}

\v{9}The kings of the earth, who committed sexual immorality with her and lived in luxury with her, will cry and mourn over her when they see the smoke rising from the fire that consumes her.\fnote{Lit. \fbib{from her fire}} \v{10}Frightened by the torture that she experiences,\fnote{Lit. \fbib{by her torture}} they will stand far away and cry out,

\begin{poetry}
\poeml ``How terrible, how terrible it is for that great city, \\
\poemll    the powerful city Babylon, \\
\poemlll       because your judgment arrived in a single hour!''
\end{poetry}

\v{11}The world's businesses cry and mourn over her, because no one buys their cargo anymore--- \v{12}cargo of gold, silver, gems, pearls, fine linen, purple cloth, silk, scarlet cloth, all kinds of scented wood, all articles made of ivory, all articles made of very costly wood, bronze, iron, marble, \v{13}cinnamon, spice, incense, myrrh, frankincense, wine, olive oil, flour, wheat, cattle, sheep, horses, chariots, and slaves (that is, human souls)---

\begin{poetry}
\poeml \v{14}``The fruit that you crafted has abandoned you. \\
\poemll    All your dainties and your splendor are lost,\fnote{Lit. \fbib{lost to you}} \\
\poemlll       and no one will ever find them again.''
\end{poetry}

\v{15}Frightened by the severity of her punishment, businesses that had become rich because of her will stand at a distance, crying and mourning:

\begin{poetry}
\poeml \v{16}``How terrible, how terrible it is for the great city \\
\poemll    that was clothed in fine linen, purple, and scarlet \\
\poemlll       and was adorned with gold, gems, and pearls, \\
\poeml \v{17}because all this wealth has been destroyed in a single hour!''
\end{poetry}

Every ship's captain, everyone who traveled by ship, sailors, and everyone who made a living from the sea stood far away. \v{18}When they saw the smoke rising from the fire that consumed her,\fnote{Lit. \fbib{from her fire}} they began to cry out, ``What city was like that great city?'' \v{19}Then they threw dust on their heads and shouted while crying and mourning:

\begin{poetry}
\poeml ``How terrible, how terrible it is for the great city, \\
\poemll    where all who had ships at sea became rich from her wealth, \\
\poemlll       because it has been destroyed in a single hour! \\
\poeml \v{20}Be happy about her, heaven, saints, apostles, and prophets, \\
\poemll    for God has condemned her for you!''
\end{poetry}
\passage{The Vision of the Powerful Angel with the Millstone}

\v{21}Then a powerful angel picked up a stone that was like a large millstone and threw it into the sea, saying,

\begin{poetry}
\poeml ``The great city Babylon will be thrown down violently--- \\
\poemll    and will never be found again. \\
\poeml \v{22}The sound of harpists, musicians, flutists, and trumpeters \\
\poemll    will never be heard within you again. \\
\poeml No artisan of any trade \\
\poemll    will ever be found within you again. \\
\poeml The sound of a millstone \\
\poemll    will never be heard within you again. \\
\poeml \v{23}The light from a lamp \\
\poemll    will never shine within you again. \\
\poeml The voice of a bridegroom and bride \\
\poemll    will never be heard within you again. \\
\poeml For your merchants were the important people of the world, \\
\poemll    and all the nations were deceived by your witchcraft. \\
\poeml \v{24}The blood of the world's prophets, saints, \\
\poemll    and all who had been murdered \\
\poemlll       was found within her.''
\end{poetry}
\labelchapt{19}
\passage{The Vision of Worship}

\chapt{19}
\v{1}After these things, I heard what sounded like the loud voice of a large crowd in heaven, saying,

\begin{poetry}
\poeml ``Hallelujah! \\
\poemll    Salvation, glory, and power belong to our God. \\
\poeml \v{2}His judgments are true and just. \\
\poemll    He has condemned the notorious prostitute \\
\poemlll       who corrupted the world with her immorality. \\
\poeml He has taken revenge on her \\
\poemll    for the blood of his servants.''
\end{poetry}

\v{3}A second time they said,

\begin{poetry}
\poeml ``Hallelujah! \\
\poemll    The smoke goes up from her forever and ever.''
\end{poetry}

\v{4}The twenty-four elders and the four living creatures bowed down and worshipped God, who was sitting on the throne. They said, ``Amen! Hallelujah!'' \v{5}A voice came from the throne, saying,

\begin{poetry}
\poeml ``Praise our God, \\
\poemll    all who serve and fear him, \\
\poeml from the least important \\
\poemll    to the most important.''
\end{poetry}
\passage{The Vision of the Marriage Supper of the Lamb}

\v{6}Then I heard what sounded like the voice of a large crowd, like the sound of raging waters, and like the sound of powerful thunderclaps, saying,

\begin{poetry}
\poeml ``Hallelujah! \\
\poemll    The Lord our God, the Almighty, is reigning. \\
\poeml \v{7}Let us rejoice, be glad, and give him glory, \\
\poemll    because the marriage of the lamb has come \\
\poemlll       and his bride has made herself ready. \\
\poeml \v{8}She has been given the privilege of wearing fine linen, \\
\poemll    dazzling and pure.''
\end{poetry}

(The fine linen represents the righteous deeds of the saints.)

\v{9}Then the angel\fnote{Lit. \fbib{he}} told me, ``Write this: `How blessed are those who are invited to the marriage supper of the lamb!'\,'' He also told me, ``These are the true words of God.'' \v{10}I bowed down at his feet to worship him, but he told me, ``Don't do that! I am a fellow servant with you and with your brothers who rely on what Jesus is saying. Worship God, because what Jesus is saying is the spirit of prophecy!''
\passage{The Vision of the Coming of the Messiah}

\v{11}Then I saw heaven standing open, and there was a white horse! Its rider is named Faithful and True. He administers justice and wages war righteously. \v{12}His eyes are like a flame of fire, and on his head are many royal crowns. He has a name written on him that nobody knows except himself. \v{13}He is dressed in a robe dipped in\fnote{Other mss. read \fbib{sprinkled with}} blood, and his name is called the Word of God. \v{14}The armies of heaven, wearing fine linen, white and pure, follow him on white horses. \v{15}A sharp sword comes out of his mouth to strike down the nations. He will rule\fnote{Or \fbib{shepherd}} them with an iron rod and tread the winepress of the fury of the wrath of God Almighty. \v{16}On his robe that covers\fnote{Lit. \fbib{robe and on}} his thigh he has a name written:

\divine{King of Kings and Lord of Lords}
\passage{The Vision of the Angel's Invitation}

\v{17}Then I saw an angel standing in the sun. He cried out in a loud voice to all the birds flying overhead, ``Come! Gather for the great supper of God. \v{18}Eat the flesh of kings, the flesh of commanders, the flesh of warriors,\fnote{Lit. \fbib{of the powerful}} the flesh of horses and their riders, and the flesh of all people, both free and slaves, both unimportant and important.''
\passage{The Vision of the Judgment of the Beast and False Prophet}

\v{19}Then I saw the beast, the kings of the earth, and their armies gathered to wage war against the rider on the horse and his army. \v{20}The beast was captured, along with the false prophet who had performed signs on its behalf.\fnote{Or \fbib{in its presence}} By these signs\fnote{Lit. \fbib{By which}} the false prophet\fnote{Lit. \fbib{he}} had deceived those who had received the mark of the beast and worshipped its image. Both of them were thrown alive into the lake of fire that burns with sulfur. \v{21}The rest were killed by the sword that belonged to the rider on the horse and that came from his mouth, and all the birds gorged themselves with their flesh.
\labelchapt{20}
\passage{The Vision of the Millennial Reign}

\chapt{20}
\v{1}Then I saw an angel coming down from heaven, holding the key to the bottomless pit,\fnote{I.e. the realm of punishment in the afterlife} with a large chain in his hand. \v{2}He captured the dragon, that ancient serpent, also known as the devil and Satan, and tied him up for a thousand years. \v{3}He threw him into the bottomless pit,\fnote{I.e. the realm of punishment in the afterlife} locked it, and sealed it over him to keep him from deceiving the nations anymore until the thousand years were over. After that, he must be set free for a little while.
\passage{The Vision of the First Resurrection}

\v{4}Then I saw thrones, and those who sat on them were given authority to judge. I also saw the souls of those who had been beheaded because of their testimony about Jesus and because of the word of God. They had not worshipped the beast or its image and had not received its mark on their foreheads or hands. They came back to life and ruled with the Messiah\fnote{Or \fbib{Christ}} for a thousand years. \v{5}The rest of the dead did not come back to life until the thousand years were over. This is the first resurrection. \v{6}How blessed and holy are those who participate in the first resurrection! The second death has no power over them. They will be priests of God and the Messiah,\fnote{Or \fbib{Christ}} and will rule with him for a thousand years.
\passage{The Vision of the Final Judgment}

\v{7}When the thousand years are over, Satan will be freed from his prison. \v{8}He will go out to deceive Gog and Magog, the nations at the four corners of the earth, and gather them for war. They are as numerous as the sands of the seashore. \v{9}They marched over the broad expanse of the earth and surrounded the camp of the saints and the beloved city. Fire came from God\fnote{Other mss. lack \fbib{from God}} out of heaven and burned them up, \v{10}and the devil who deceived them was thrown into the lake of fire and sulfur, where the beast and the false prophet were. They will be tortured day and night forever and ever.
\passage{The Vision of the White Throne Judgment}

\v{11}Then I saw a large, white throne and the one who was sitting on it. The earth and the heavens fled from his presence, and no place was found for them. \v{12}I saw the dead, both unimportant and important, standing in front of the throne, and books were open. Another book was opened---the Book of Life. The dead were judged according to their actions, as recorded in the books. \v{13}The sea gave up the dead that were in it, and Death and Hades\fnote{I.e. the realm of the dead} gave up the dead that were in them, and all were judged according to their actions. \v{14}Death and Hades\fnote{I.e. the realm of the dead} were thrown into the lake of fire. (This is the second death---the lake of fire.) \v{15}Anyone whose name was not found written in the Book of Life was thrown into the lake of fire.
\labelchapt{21}
\passage{The Vision of the New Heaven and the New Earth}

\chapt{21}
\v{1}Then I saw a new heaven and a new earth, because the first heaven and the first earth had disappeared, and the sea was gone. \v{2}I also saw the holy city, New Jerusalem, coming down from God out of heaven, prepared like a bride adorned for her husband.

\v{3}I heard a loud voice from the throne say,

\begin{poetry}
\poeml \red{``See, the tent of God is among humans!} \\
\poemll    \red{He will make his home with them,} \\
\poemlll       \red{and they will be his people.} \\
\poeml \red{God himself will be with them,} \\
\poemll    \red{and he will be their God.}\fnote{Other mss. lack \fbib{and he will be their God}} \\
\poeml \v{4}\red{He will wipe every tear from their eyes.} \\
\poemll    \red{There won't be death anymore.} \\
\poeml \red{There won't be any grief, crying, or pain,} \\
\poemll    \red{because the first things have disappeared.''}
\end{poetry}

\v{5}The one sitting on the throne said, \red{``See, I am making all things new!''}

He said, \red{``Write this: `These words are trustworthy and true.'\,''}

\v{6}Then he told me, \red{``It has happened! I am the Alpha and the Omega, the beginning and the end. I will freely give a drink from the spring of the water of life to the one who is thirsty.} \v{7}\red{The person who conquers will inherit these things. I will be his God, and he will be my son.} \v{8}\red{But people who are cowardly, unfaithful, detestable, murderers, sexually immoral, sorcerers, idolaters, and all liars will find themselves\fnote{Lit. \fbib{will have their part}} in the lake that burns with fire and sulfur. This is the second death.''}
\passage{The Vision of the New Jerusalem}

\v{9}Then one of the seven angels who had the seven bowls full of the seven last plagues came to me and said, ``Come! I will show you the bride, the wife of the lamb.'' \v{10}He carried me away in the Spirit to a large, high mountain and showed me the holy city, Jerusalem, coming down from God out of heaven. \v{11}The glory of God was its radiance, and its light was like a valuable gem, like jasper, as clear as crystal. \v{12}It had a large, high wall with twelve gates. Twelve angels were at the gates, and the names of the twelve tribes of Israel were written on the gates. \v{13}There were three gates on the east, three gates on the north, three gates on the south, and three gates on the west. \v{14}The wall of the city had twelve foundations, and the twelve names of the twelve apostles of the lamb were written on them.

\v{15}The angel who\fnote{Lit. \fbib{The one who}} was talking to me had a gold measuring rod to measure the city, its gates, and its walls. \v{16}The city was cubic in shape: its length was the same as its width. He measured the city with his rod, and it measured at 12,000 stadia:\fnote{I.e. about 1,454 miles long ; one Roman stadion was about 640 feet}Its length, width, and height were the same. \v{17}He also measured its wall. According to the human measurement that the angel was using, it was 144 cubits.\fnote{I.e. about 84 yards, if a royal cubit (about 21 inches long) is intended; because height, thickness, or width is not specified, since the city structure is three-dimensional, the wall may be a shell that is 144 cubits thick, enclosing the entire structure} \v{18}Its wall was made of jasper. The city was made of pure gold, as clear as glass.

\v{19}The foundations of the city wall were decorated with all kinds of gems: The first foundation\fnote{The Gk. lacks \fbib{foundation}} was jasper, the second sapphire, the third agate, the fourth emerald, \v{20}the fifth onyx, the sixth carnelian,\fnote{Or \fbib{olivine}; i.e. an olive-green mineral commonly found in igneous rocks} the seventh chrysolite,\fnote{I.e. a bright green variety of chalcedony, a transparent silica quartz} the eighth beryl,\fnote{I.e. an aluminum silicate varying in color from blue, rose, white, and golden, in both opaque and transparent varieties, including emerald and aquamarine} the ninth topaz,\fnote{I.e. a bright yellow (but occasionally blue, green, brown, pink, or colorless) aluminum silicate variety of sapphire or corundum, and serving as the second stone in the first row in the high priest's breast plate (cf. Exod 28:17)} the tenth chrysoprase,\fnote{I.e. a bright green variety of chalcedony, a transparent silica quartz} the eleventh jacinth\fnote{I.e. a red or cinnamon-colored variety of transparent zircon; sometimes also colored deep purple, blue, or violet and serving as the first stone in row three of the high priest's breast plate (cf. Exod 28:19)} and the twelfth amethyst. \v{21}The twelve gates were twelve pearls, and each gate was made of a single pearl. The street of the city was made of pure gold, as clear as glass.

\v{22}I saw no temple in it, because the Lord God Almighty and the lamb are its temple. \v{23}The city doesn't need any sun or moon to give it light, because the glory of God gave it light, and the lamb was its lamp. \v{24}The nations will walk in its light, and the kings of the earth will bring their glory into it. \v{25}Its gates will never be shut at the end of the\fnote{Lit. \fbib{shut by}} day---because there will be no night there. \v{26}People\fnote{Lit. \fbib{They}} will bring the glory and wealth\fnote{Or \fbib{honor}} of the nations into it. \v{27}Nothing unclean, or anyone who does anything detestable, and no one who tells lies will ever enter it. Only those whose names\fnote{Lit. \fbib{those who}} are written in the lamb's Book of Life will enter it.\fnote{The Gk. lacks \fbib{will enter it}}
\labelchapt{22}
\passage{The Vision of the River of the Water of Life}

\chapt{22}
\v{1}Then the angel\fnote{Lit. \fbib{he}} showed me the river of the water of life, as clear as crystal. It was flowing from the throne of God and the lamb. \v{2}Between the city street and the river, the tree of life was visible from each side. It produced twelve kinds of fruit, each month having its own fruit. The leaves of the tree are for the healing of the nations. \v{3}There will no longer be any curse. The throne of God and the lamb will be in the city.\fnote{Lit. \fbib{in it}} His servants will worship him \v{4}and see his face, and his name will be on their foreheads. \v{5}There will be no more night, and they will not need any light from lamps or the sun because the Lord God will shine on them. They will rule forever and ever.
\passage{He is Coming Soon}

\v{6}He told me, ``These words are trustworthy and true. The Lord God of the spirits and of the prophets has sent his messenger\fnote{Or \fbib{angel}} to show his servants the things that must happen soon.''

\v{7}\red{``See! I am coming soon! How blessed is the person who keeps the words of the prophecy in this book!''}
\passage{The Vision of an Exhortation to Worship God}

\v{8}I, John, heard and saw these things. When I had heard and seen them, I bowed down to worship at the feet of the angel who had been showing me these things. \v{9}But he told me, ``Don't do that! I am a fellow servant with you, your brothers the prophets, and those who keep the words in this book. Worship God!''

\v{10}Then he told me, ``Don't seal up the words of the prophecy in this book, because the time is near. \v{11}Let the one who does what is evil continue to do what is evil. Let the filthy person continue to be filthy. Let the righteous person continue to do what is right. And let the holy person continue to be holy.''
\passage{Concluding Benediction}

\v{12}\red{``See! I am coming soon! My reward is with me to repay everyone according to his behavior.} \v{13}\red{I am the Alpha and the Omega, the first and the last, the beginning and the end.}\fnote{The speaker possibly concludes his quotation here rather than at the end of vs. 16.}

\v{14}\red{``How blessed are those who wash their robes\fnote{Other mss. read \fbib{who do his commandments}} so that they may have the right to the tree of life and may go through the gates into the city!} \v{15}\red{Outside are dogs, sorcerers, immoral people, murderers, idolaters, and everyone who loves and practices falsehood.}

\v{16}\red{``I, Jesus, have sent my angel to give this testimony to you for the churches. I am the root and descendant of David, the bright morning star.''}
\passage{Concluding Invitation}

\v{17}The Spirit and the bride say, ``Come!''

Let everyone who hears this say, ``Come!''

Let everyone who is thirsty come!

Let anyone who wants the water of life take it as a gift!
\passage{Concluding Warning}

\v{18}I warn everyone who hears the words of the prophecy in this book: If anyone adds anything to them, God will strike him with the plagues that are written in this book. \v{19}If anyone takes away any words from the book of this prophecy, God will take away his portion of the tree of life and the holy city that are described in this book.
\passage{Epilogue}

\v{20}The one who is testifying to these things says, \red{``Yes, I am coming soon!''}

Amen! Come, Lord Jesus!

\v{21}May the grace of the Lord Jesus be with all the saints.

Amen.\fnote{Other mss. lack \fbib{Amen}}
