\bookheader{Genesis}
\labelbook{Gen}

\bookpretitle{The First Book of the Law called}
\booktitle{Genesis}

\labelchapt{1}
\passage{The Creation}

\chapt{1}
\v{1}In the beginning, God created the universe.\fnote{\fbackref{1:1} Lit. \fbib{the heavens and the earth}; i.e. space and matter} \v{2}When the earth\fnote{\fbackref{1:2} Or \fbib{\v{1}When God began to create the universe,} \fbib{\v{2}the earth}} was as yet unformed and desolate, with the surface of the ocean depths shrouded\fnote{\fbackref{1:2} The Heb. lacks \fbib{shrouded}} in darkness, and while the Spirit of God was hovering\fnote{\fbackref{1:2} Or \fbib{brooding}} over the surface of the waters, \v{3}God said, ``Let there be light!'' So there was light.

\v{4}God saw that the light was beautiful.\fnote{\fbackref{1:4} Or \fbib{good}} He\fnote{\fbackref{1:5} Lit. \fbib{God}} separated the light from the darkness, \v{5}calling the light ``day,'' and the darkness\fnote{\fbackref{1:5} Lit. \fbib{darkness he called}} ``night.'' The twilight and the dawn were day one.

\v{6}Then God said, ``Let there be a canopy\fnote{\fbackref{1:6} Or \fbib{an expanse}} between bodies of water,\fnote{\fbackref{1:6} Lit. \fbib{between waters}} separating bodies of water\fnote{\fbackref{1:6} Lit. \fbib{separating waters}} from bodies of water!''\fnote{\fbackref{1:6} Lit. \fbib{from waters}} \v{7}So God made a canopy\fnote{\fbackref{1:7} Or \fbib{made an expanse}} that separated the water beneath the canopy\fnote{\fbackref{1:7} Or \fbib{expanse}} from the water above it.\fnote{\fbackref{1:7} Lit. \fbib{above the canopy}} And that is what happened:\fnote{\fbackref{1:7} Lit. \fbib{And so it was}} \v{8}God called the canopy\fnote{\fbackref{1:8} Or \fbib{expanse}} ``sky.''\fnote{\fbackref{1:8} Or \fbib{Heaven}} The twilight and the dawn were the second day.

\v{9}Then God said, ``Let the water beneath the sky come together into one area, and let dry ground appear!'' And that is what happened:\fnote{\fbackref{1:9} Lit. \fbib{And so it was}} \v{10}God called the dry ground ``land,''\fnote{\fbackref{1:10} Or \fbib{Earth}} and he called the water that had come together ``oceans.'' And God saw how good it was.

